### 2024-11-23 paper/paper.tex
Initial commit

\documentclass[a4paper,12pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{color}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{subcaption}

\geometry{margin=1in}

\title{BART}
\author{
    Vincenzo Dorrello 
    \and
    Giulio Frey 
    \and
    Guido Rossetti 
    \and
    Giovanni Scarpato 
}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Your abstract here.
\end{abstract}

%\tableofcontents

\section{Introduction}
Your introduction here.

\end{document}

### 2024-11-23 code/ code.r
adding code file

@@ -1,0 +1,16 @@
+# Package names
+packages <- c(
+    "ggplot2", 
+    "dplyr",
+    "BayesTree", # For Bayesian Additive Regression Trees
+    "gbm", # For Generalized Boosted Regression Models
+    "randomForest" # For Random Forest
+)
+# Install packages not yet installed
+installed_packages <- packages %in% rownames(installed.packages())
+if (any(installed_packages == FALSE)) {
+  install.packages(packages[!installed_packages])
+}
+
+# Packages loading
+invisible(lapply(packages, library, character.only = TRUE))

### 2024-11-24 paper/paper.tex
adding bibliography
Work at Collegio

@@
-\documentclass[a4paper,12pt]{article}
+\documentclass[a4paper,11pt]{article}
 
@@
 \usepackage{geometry}
-\usepackage{natbib}
 \usepackage{color}
@@
 \usepackage{subcaption}
+\usepackage{bookmark}
+\usepackage[
+    backend=biber,
+    style=authoryear,
+  ]{biblatex}
 
-\geometry{margin=1in}
+\addbibresource{bibliography.bib}
 
-\title{BART}
+\title{BART: A Comparison with Non-bayesian Tree Methods}
 \author{
@@
 \section{Introduction}
-Your introduction here.
+
+Bayesian Regression Additive Trees (BART) were first introduced by \cite{chipmanBARTBayesianAdditive2010}
+
+\section{Non-Bayesian Tree Based Methods}
+
+Tree based method are more usefull compared to the linear regression model when the relation to be model is complex or non-linear. Indeed, the standard linear model is defined as:
+\begin{equation}
+  f(X) = \beta_0 + \sum_{j=1}^p X_j \beta_j
+\end{equation}
+Wheras the decision tree model as:
+\begin{equation}
+  f(X) = \sum_{m=1}^M c_m \cdot 1(X \in R_m)
+\end{equation}
+DESCRIBE BETTER THIS EQUATIONS
+
+We take most of our arguments for this section from the \cite[Chapter~8]{jamesIntroductionStatisticalLearning2021}.
+
+\subsection{Regression Trees}
+
+Tree based methods that we cover in this paper are all based on decision trees.
+Decision trees can be applied both to regression or classification problems, and are thus named accordingly to their application.
+Regression trees in particular are build in two steps.
+
+First, the prediction space (the set of all possible values of the vector of observables) is split into $J$ regions $R_j$ that are distinct and non-overlapping.
+Secondly, a prediction for the outcome variable is made for each observation in each $R_j$ regions, as the mean of the training observations in that same regions.
+
+For a predictor space of $P$ elements the regions are $p$ dimensional boxes, that are build by minimizing the Residual Sum Squared, as the equation \ref{rss_tree} shows.
+\begin{equation}
+  R_1,...,R_j=\operatorname*{argmin}_{R_j}\sum^J_{j=1}\sum_{i\in R_j}\left(y_i-\hat{y}_{r_j}\right)^2
+  \label{rss_tree}
+\end{equation}
+
+Instead of considering all the possible partitions of the predictor space, a top-down approach is used.
+First, all observations are considered such that they belong to the same region.
+Then, this process chooses the best split minimizing the RSS irrespective of future outcomes.
+This is then repeated for a number of times, until a stopping criterion is reached (usually a minimum number of observations in all the regions).
+
+As the regions visualizations are complex when more than 2 regressors are present, a tree based graph is used.
+
+The process described above will likely tend to overfit the data.
+There are numerous solutions to this, the most simple one being increasing the number of AND GENERALLY THE ENSAMBLING MODELSobservations that rise the stopping rule.
+A valid alternative is to grow a very large tree, and then reduce it to obtain a smaller subtree.
+This is known as tree pruning.
+
+ADD COST COMPLEXITY PRUNING
+
+One of the main disadvantages of this setting, is that trees are non-robust.
+In fact a small change in the data can lead to a large change in the estimation.
+This follows from the fact that Regression Trees suffer from high variance.
+This is indeed quite problematic and the main reason that simple regression trees are not used by practitioners.
+Ensemble methods aim at solving this issue.
+Those methods in the context of trees combine simpler models (called weak learners), whichy may lead to mediocre predictions, in a single more powerful and stable model.
+QUelli di cui sotto sopra abuiao detto sono tutti ensable models.
+
+\subsection{Bagging}
+Bagging is a generic method that aims at reducing the variance of a statistical learning method.
+Bagging is based on the fact that if we have independent observations $(Z_1,...,Z_n)$ with same variance $\sigma$, then the vairance of the sample mean of the observation will be lower, as it is $V(\bar{Z})=\frac{\sigma}{n}$.
+
+Bagging uses a bootstrap sampling procedure to obtain different samples from the data, and then fits a tree for each of the bootstrapped samples.
+The single estimates for each sample are then averaged, as equation \ref{eq_bagging} shows.
+\begin{equation}
+  \label{eq_bagging}
+  \hat{f}_{\text{bag}}(x) = \frac{1}{B} \sum_{b=1}^{B} \hat{f}^{*b}(x).
+\end{equation}
+
+\newpage
+
+\printbibliography
 

### 2024-12-02 paper/paper.tex
adding work before push on overleaf

@@
 \usepackage[utf8]{inputenc}
+\usepackage[margin=1in]{geometry}
 \usepackage[T1]{fontenc}
@@
 \usepackage{caption}
+\usepackage[ruled,vlined]{algorithm2e}
 \usepackage{subcaption}
@@
 \end{equation}
+Where B is the number of samples of the bootstrapping procedure.
+
+\subsection{Boosting}
+
+Similar to bagging, boosting is not specific to regression trees, it is rather a generic approach.
+Said so, it is quite effective when used on trees.
+In fact ,it approaches the overfitting problem but without using bootstrapping, as bagging does.
+Boosting relies on fitting trees sequentially, as algorithm \ref{alg_boosting} shows.
+
+\begin{algorithm}
+  \caption{Boosting for Regression Trees}
+  \label{alg_boosting}
+  \SetAlgoLined
+  \DontPrintSemicolon
+  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
+
+  Set $\hat{f}(x) = 0$ and $r_i = y_i$ for all $i$ in the training set.\;
+  \For{$b = 1, 2, \ldots, B$}{
+      (a) Fit a tree $f^b$ with $d$ splits ($d+1$ terminal nodes) to the training data $(X, r)$.\;
+      (b) Update $\hat{f}$ by adding in a shrunken version of the new tree:
+      \begin{equation}
+      \hat{f}(x) \gets \hat{f}(x) + \lambda f^b(x).
+      \end{equation}\;
+      (c) Update the residuals:
+      \begin{equation}
+      r_i \gets r_i - \lambda f^b(x_i).
+      \end{equation}\;
+  }
+  Output the boosted model:
+  \begin{equation}
+  \hat{f}(x) = \sum_{b=1}^B \lambda f^b(x).
+  \end{equation}\;
+
+\end{algorithm}
 

### 2024-12-02 paper/paper.tex
adding work on draft alone

@@
 \usepackage{color}
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{1em}
 \usepackage{booktabs}
@@
 \end{equation}
-Where B is the number of samples of the bootstrapping procedure.
+Where $B$ is the number of samples of the bootstrapping procedure.
 
@@
 Boosting relies on fitting trees sequentially, as algorithm \ref{alg_boosting} shows.
+
+The number of sequential trees to fit, as chosen by the parameter $B$ in algorithm  \ref{alg_boosting}, if chosen too large, can cause overfitting.
+The learning rate $\lambda$ is another tunable parameter.
+Often small values are chosen ($\lambda=10^{-2}$ or even smaller).
 

### 2024-12-02 code/ code.r
Work on code

@@ -2,7 +2,8 @@
 packages <- c(
     "ggplot2", 
     "dplyr",
-    "BayesTree", # For Bayesian Additive Regression Trees
+    "MASS",
+    "BART",
     "gbm", # For Generalized Boosted Regression Models
     "randomForest" # For Random Forest
 )
@@ -14,3 +15,34 @@ if (any(installed_packages == FALSE)) {
 
 # Packages loading
 invisible(lapply(packages, library, character.only = TRUE))
+
+# Settting the seed
+set.seed(123)
+
+x <- Boston[, c(6, 13)] 
+
+dim(x)
+
+y <- Boston$medv
+
+nd <- 200 # The number of posterior draws
+burn <- 50 # Number of MCMC iterations to be treated as burn in
+
+
+bart <- wbart(x, y, nskip = burn, ndpost = nd, ntree = 1000)
+bart.mse <- mean((bart$yhat.train.mean - y)^2)
+
+rf <- randomForest(x, y, ntree = 1000)
+
+
+# TODO understand the structure of the gbm object and why works only with data frame
+gbm <- gbm(medv ~ ., data = Boston, n.trees = 1000, distribution = "gaussian", interaction.depth = 1)
+
+#TODO implement bagging
+
+mse_df <- data.frame(
+    mse = c(bart.mse, mean(rf$mse), mean(gbm$train.error)),
+    method = c("BART", "Random Forest", "GBM")
+)
+
+print(mse_df)

### 2024-12-04 code/ code.r
modifying gitignore and adding work to clean the code

@@ -14,35 +14,37 @@ if (any(installed_packages == FALSE)) {
 }
 
 # Packages loading
-invisible(lapply(packages, library, character.only = TRUE))
+invisible(lapply(packages, library, character.only = TRUE, warn.conflicts = FALSE))
 
 # Settting the seed
 set.seed(123)
 
+# Load the Boston dataset and create a dataframe
 x <- Boston[, c(6, 13)] 
-
-dim(x)
-
+boston_df <- as.data.frame(Boston[, c(6, 13)])
+boston_df$medv <- Boston$medv
 y <- Boston$medv
 
 nd <- 200 # The number of posterior draws
 burn <- 50 # Number of MCMC iterations to be treated as burn in
 
-
+#Bayesian Additive Regression Trees
 bart <- wbart(x, y, nskip = burn, ndpost = nd, ntree = 1000)
 bart.mse <- mean((bart$yhat.train.mean - y)^2)
 
-rf <- randomForest(x, y, ntree = 1000)
+# Random Forest
+rf <- randomForest(medv ~ ., data = boston_df, ntree = 1000)
+
+# Generalized Boosted Regression Models
+gbm <- gbm(medv ~ ., data = boston_df, n.trees = 1000, distribution = "gaussian", interaction.depth = 1)
+
+# Bagging
+bag <- randomForest(medv ~ ., data = boston_df, mtry = 2, ntree = 1000)
 
 
-# TODO understand the structure of the gbm object and why works only with data frame
-gbm <- gbm(medv ~ ., data = Boston, n.trees = 1000, distribution = "gaussian", interaction.depth = 1)
-
-#TODO implement bagging
-
-mse_df <- data.frame(
-    mse = c(bart.mse, mean(rf$mse), mean(gbm$train.error)),
-    method = c("BART", "Random Forest", "GBM")
+valuation <- data.frame(
+    mse = c(bart.mse, mean(rf$mse), mean(gbm$train.error), mean(bag$mse)),
+    method = c("BART", "Random Forest", "GBM", "Bagging")
 )
 
 print(mse_df)

### 2024-12-04 code/ code.r
adding underspecified model

@@ -19,11 +19,15 @@ invisible(lapply(packages, library, character.only = TRUE, warn.conflicts = FALS
 # Settting the seed
 set.seed(123)
 
-# Load the Boston dataset and create a dataframe
-x <- Boston[, c(6, 13)] 
-boston_df <- as.data.frame(Boston[, c(6, 13)])
-boston_df$medv <- Boston$medv
+# Load the full Boston dataset and create a dataframe
+x <- Boston[, -14]
+boston_df <- as.data.frame(Boston)
 y <- Boston$medv
+
+# Load an underspecified dataset
+x_under <- Boston[, c(6, 13)]
+y_under <- Boston$medv
+boston_under_df <- as.data.frame(Boston[, c(6, 13, 14)])
 
 nd <- 200 # The number of posterior draws
 burn <- 50 # Number of MCMC iterations to be treated as burn in
@@ -32,19 +36,34 @@ burn <- 50 # Number of MCMC iterations to be treated as burn in
 bart <- wbart(x, y, nskip = burn, ndpost = nd, ntree = 1000)
 bart.mse <- mean((bart$yhat.train.mean - y)^2)
 
+#underspecified model
+bart_under <- wbart(x_under, y_under, nskip = burn, ndpost = nd, ntree = 1000)
+bart_under.mse <- mean((bart_under$yhat.train.mean - y_under)^2)
+
 # Random Forest
 rf <- randomForest(medv ~ ., data = boston_df, ntree = 1000)
+
+# underspecified model
+rf_under <- randomForest(medv ~ ., data = boston_under_df, ntree = 1000)
 
 # Generalized Boosted Regression Models
 gbm <- gbm(medv ~ ., data = boston_df, n.trees = 1000, distribution = "gaussian", interaction.depth = 1)
 
-# Bagging
-bag <- randomForest(medv ~ ., data = boston_df, mtry = 2, ntree = 1000)
+# underspecified model
+gbm_under <- gbm(medv ~ ., data = boston_under_df, n.trees = 1000, distribution = "gaussian", interaction.depth = 1)
+
+# Bagging (Random Forest with mtry = 12, full variables considered)
+bag <- randomForest(medv ~ ., data = boston_df, mtry = 12, ntree = 1000)
+
+#underspecified model
+bag_under <- randomForest(medv ~ ., data = boston_under_df, mtry = 2, ntree = 1000)
 
 
 valuation <- data.frame(
[…]

### 2024-12-04 code/ code.r
improving out o9f sample predicyion

@@ -20,43 +20,63 @@ invisible(lapply(packages, library, character.only = TRUE, warn.conflicts = FALS
 set.seed(123)
 
 # Load the full Boston dataset and create a dataframe
-x <- Boston[, -14]
+train <- sample(1:nrow(Boston), nrow(Boston) / 2)
 boston_df <- as.data.frame(Boston)
-y <- Boston$medv
+x <- Boston[, 1:12]
+y <- Boston[, "medv"]
+xtrain <- x[train, ]
+ytrain <- y[train]
+xtest <- x[-train, ]
+ytest <- y[-train]
 
 # Load an underspecified dataset
-x_under <- Boston[, c(6, 13)]
-y_under <- Boston$medv
 boston_under_df <- as.data.frame(Boston[, c(6, 13, 14)])
+train_under <- sample(1:nrow(boston_under_df), nrow(boston_under_df) / 2)
+x_under <- Boston[, c(6, 13)]
+y_under <- Boston[, "medv"]
+xtrain_under <- x_under[train_under, ]
+ytrain_under <- y_under[train_under]
+xtest_under <- x_under[-train_under, ]
+ytest_under <- y_under[-train_under]
 
 nd <- 200 # The number of posterior draws
 burn <- 50 # Number of MCMC iterations to be treated as burn in
 
 #Bayesian Additive Regression Trees
-bart <- wbart(x, y, nskip = burn, ndpost = nd, ntree = 1000)
-bart.mse <- mean((bart$yhat.train.mean - y)^2)
+bart <- gbart(xtrain, ytrain, x.test = xtest)
+bart.mse <- mean((bart$yhat.test.mean - y)^2)
 
 #underspecified model
-bart_under <- wbart(x_under, y_under, nskip = burn, ndpost = nd, ntree = 1000)
-bart_under.mse <- mean((bart_under$yhat.train.mean - y_under)^2)
+bart_under <- wbart(xtrain_under, ytrain_under, x.test = xtest_under)
+bart_under.mse <- mean((bart_under$yhat.test.mean - y)^2)
 
 # Random Forest
-rf <- randomForest(medv ~ ., data = boston_df, ntree = 1000)
+rf <- randomForest(medv ~ ., data = boston_df, ntree = 1000, subset = train)
+rf.mse >- mean((rf$predicted - y)^2)
 
 # underspecified model
-rf_under <- randomForest(medv ~ ., data = boston_under_df, ntree = 1000)
[…]

### 2024-12-07 code/ code.r
created code for multiple runs to test optimal number of trees

@@ -5,7 +5,8 @@ packages <- c(
     "MASS",
     "BART",
     "gbm", # For Generalized Boosted Regression Models
-    "randomForest" # For Random Forest
+    "randomForest", # For Random Forest,
+    "ucimlrepo" # For the Abalone age prediction dataset,
 )
 # Install packages not yet installed
 installed_packages <- packages %in% rownames(installed.packages())
@@ -19,71 +20,72 @@ invisible(lapply(packages, library, character.only = TRUE, warn.conflicts = FALS
 # Settting the seed
 set.seed(123)
 
-# Load the full Boston dataset and create a dataframe
-train <- sample(1:nrow(Boston), nrow(Boston) / 2)
-boston_df <- as.data.frame(Boston)
-x <- Boston[, 1:12]
-y <- Boston[, "medv"]
+#load the crab dataset
+abalone <- fetch_ucirepo(name = "Abalone")
+
+abalone_df <- abalone$data$original
+
+
+#convert the categorical variables to factors
+abalone_df$Sex <- as.factor(abalone_df$Sex)
+
+#convert rings to age
+abalone_df$Age <- abalone_df$Rings + 1.5
+#remove the rings column
+abalone_df <- abalone_df[, setdiff(names(abalone_df), "Rings")]
+
+
+train <- sample(1:nrow(abalone_df), nrow(abalone_df) / 2)
+x <- abalone_df[,  setdiff(names(abalone_df), "Age")]
+y <- abalone_df[, "Age"]
 xtrain <- x[train, ]
 ytrain <- y[train]
 xtest <- x[-train, ]
 ytest <- y[-train]
 
-# Load an underspecified dataset
-boston_under_df <- as.data.frame(Boston[, c(6, 13, 14)])
-train_under <- sample(1:nrow(boston_under_df), nrow(boston_under_df) / 2)
-x_under <- Boston[, c(6, 13)]
-y_under <- Boston[, "medv"]
-xtrain_under <- x_under[train_under, ]
-ytrain_under <- y_under[train_under]
-xtest_under <- x_under[-train_under, ]
-ytest_under <- y_under[-train_under]
 
-nd <- 200 # The number of posterior draws
-burn <- 50 # Number of MCMC iterations to be treated as burn in
-
-#Bayesian Additive Regression Trees
-bart <- gbart(xtrain, ytrain, x.test = xtest)
[…]

### 2024-12-07 code/plots.r
Adding plots

@@ -1,0 +1,33 @@
+library(ggplot2)
+
+# import the data
+data <- read.csv("outputs/abalone_results.csv")
+
+mse_plot <- ggplot(data, aes(x = ntree, y = mse, color = method)) +
+  geom_line() +
+  geom_point() +
+  labs(title = "MSE vs. Number of Trees",
+       x = "Number of Trees",
+       y = "Mean Squared Error",
+       color = "Method"
+       ) +
+  theme_minimal() +
+  theme(legend.position = "bottom") +
+  scale_color_manual(values = c("red", "blue", "green", "purple"))
+
+
+
+time_plot <-ggplot(data, aes(x = ntree, y = time, color = method)) +
+  geom_line() +
+  geom_point() +
+  labs(title = "Time vs. Number of Trees",
+       x = "Number of Trees",
+       y = "Time (s)",
+       color = "Method") +
+  theme_minimal() +
+  theme(legend.position = "bottom") +
+  scale_color_manual(values = c("red", "blue", "green", "purple"))
+
+#save the plots
+ggsave("outputs/mse_plot.pdf", mse_plot, width = 6, height = 4)
+ggsave("outputs/time_plot.pdf", time_plot, width = 6, height = 4)

### 2024-12-07 paper/paper.tex
Adding section of paper

@@
 \usepackage{geometry}
+\usepackage{listing}
 \usepackage{color}
@@
 
+\section{Data Application: Predicting the age of an Abalone}
+
+Abalone are marine snails found in various regions around the world.
+Determining their age is a labor-intensive process, as it requires cutting through the shell, staining it, and then counting the growth rings under a microscope.
+However, predicting the age of an abalone can also be done by examining other physical characteristics of the individual, which are much easier to assess.
+This is precisely the motivation behind our application.
+
+We use data Abalone characteristics data coming from the UC Irvine Machine Learning Repository \parencite{warwicknashAbalone1994a}, that is described in table \ref{table1}.
+No data manipulation is necessary, aside from creating a new variable, \( \text{Age} = \text{Rings} + 1.5 \), and removing the Rings column.
+
+ We run BART from \cite{mccullochBARTBayesianAdditive2024}; Boosting from \cite{ridgewayGbmGeneralizedBoosted2024}; Random Forest and Bagging from \cite{breimanRandomForestBreimanCutlers2024}.
+We use half of the sample to fit $\hat{y}$, that is the Age variable, and half the sample to give us out of bag Mean Squared Errors (MSE) for predicting the same variable.
+Models are run for various values of TT, and for each value, both the mean squared error (MSE) of out-of-sample predictions (shown in plot \ref{plot_mse}) and the computation time (depicted in plot \ref{plot_time}) are recorded.
+
+\begin{table}[]
+  \centering
+  \begin{tabular}{llllll}
+
+  \toprule
+  Variable Name & Role & Type & Description & Units & Missing Values\\  
+  \midrule
+   Sex                  & Feature       & Categorical   & M, F, and I (infant) & -             & No                      \\   Length               & Feature       & Continuous    & Longest shell measurement & mm         & No                      \\   Diameter             & Feature       & Continuous    & Perpendicular to length & mm           & No                      \\   Height               & Feature       & Continuous    & With meat in shell      & mm           & No                      \\   Whole\_weight         & Feature       & Continuous    & Whole abalone           & grams        & No                      \\   Shucked\_weight       & Feature       & Continuous    & Weight of meat          & grams        & No                      \\   Viscera\_weight       & Feature       & Continuous    & Gut weight (after bleeding) & grams    & No                      \\   Shell\_weight         & Feature       & Continuous    & After being dried       & grams        & No                      \\   Rings                & Target        & Integer       & +1.5 gives the age in years & -          & No                      \\
+   \bottomrule
+  \end{tabular}
+  \caption{Table of Variables}
+  \label{table1}
+  \end{table}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.7\textwidth]{../outputs/mse_plot.pdf}
+  \caption{Mean Squared Error for different values of T}
+  \label{plot_mse}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.7\textwidth]{../outputs/time_plot.pdf}
+  \caption{Time of computations for different values of T}
+  \label{plot_time}
+\end{figure}
+
 \newpage

### 2024-12-07 paper/paper.tex
commit with code on simulated data

@@
 
- We run BART from \cite{mccullochBARTBayesianAdditive2024}; Boosting from \cite{ridgewayGbmGeneralizedBoosted2024}; Random Forest and Bagging from \cite{breimanRandomForestBreimanCutlers2024}.
-We use half of the sample to fit $\hat{y}$, that is the Age variable, and half the sample to give us out of bag Mean Squared Errors (MSE) for predicting the same variable.
-Models are run for various values of TT, and for each value, both the mean squared error (MSE) of out-of-sample predictions (shown in plot \ref{plot_mse}) and the computation time (depicted in plot \ref{plot_time}) are recorded.
-
 \begin{table}[]
@@
   \toprule
-  Variable Name & Role & Type & Description & Units & Missing Values\\  
+  Variable Name & Role & Type & Description & Units \\
   \midrule
-   Sex                  & Feature       & Categorical   & M, F, and I (infant) & -             & No                      \\   Length               & Feature       & Continuous    & Longest shell measurement & mm         & No                      \\   Diameter             & Feature       & Continuous    & Perpendicular to length & mm           & No                      \\   Height               & Feature       & Continuous    & With meat in shell      & mm           & No                      \\   Whole\_weight         & Feature       & Continuous    & Whole abalone           & grams        & No                      \\   Shucked\_weight       & Feature       & Continuous    & Weight of meat          & grams        & No                      \\   Viscera\_weight       & Feature       & Continuous    & Gut weight (after bleeding) & grams    & No                      \\   Shell\_weight         & Feature       & Continuous    & After being dried       & grams        & No                      \\   Rings                & Target        & Integer       & +1.5 gives the age in years & -          & No                      \\
+  Sex                  & Feature       & Categorical   & M, F, and I (infant) & -    \\
+  Length               & Feature       & Continuous    & Longest shell measurement & mm\\
+  Diameter             & Feature       & Continuous    & Perpendicular to length & mm  \\
+  Height               & Feature       & Continuous    & With meat in shell      & mm  \\
+  Whole\_weight         & Feature       & Continuous    & Whole abalone           & gram\\
+  Shucked\_weight       & Feature       & Continuous    & Weight of meat          & gram\\
+  Viscera\_weight       & Feature       & Continuous    & Gut weight (after bleeding) & \\
+  Shell\_weight         & Feature       & Continuous    & After being dried       & gram\\
+  Rings                & Target        & Integer       & +1.5 gives the age in years & - \\
+
    \bottomrule
@@
   \end{table}
+
+We run BART from \cite{mccullochBARTBayesianAdditive2024}; Boosting from \cite{ridgewayGbmGeneralizedBoosted2024}; Random Forest and Bagging from \cite{breimanRandomForestBreimanCutlers2024}.
+We use half of the sample to fit $\hat{y}$, that is the Age variable, and half the sample to give us out of bag Mean Squared Errors (MSE) for predicting the same variable.
+Models are run for various values of T, and for each value, both the mean squared error (MSE) of out-of-sample predictions (shown in plot \ref{plot_mse}) and the computation time (depicted in plot \ref{plot_time}) are recorded.
 

### 2024-12-07 code/ code.r

@@ -2,7 +2,7 @@
 packages <- c(
     "ggplot2", 
     "dplyr",
-    "MASS",
+    "tree",
     "BART",
     "gbm", # For Generalized Boosted Regression Models
     "randomForest", # For Random Forest,
@@ -19,6 +19,70 @@ invisible(lapply(packages, library, character.only = TRUE, warn.conflicts = FALS
 
 # Settting the seed
 set.seed(123)
+
+
+# simulate data from sin function
+x <- seq(0, 10, length.out = 1000)
+y <- 2*sin(x) + x + rnorm(1000, 0, 1)
+
+train <- sample(1:1000, 500)
+xtrain <- x[train]
+ytrain <- y[train]
+xtest <- x[-train]
+ytest <- y[-train]
+
+# fit a bart model
+bart_sin <- gbart(xtrain, ytrain, x.test = xtest)
+
+
+# plot the results using ggplot2
+plot_sin <- ggplot(data = data.frame(xtest, ytest), aes(x = xtest, y = ytest)) +
+  geom_point() +  
+  geom_line(aes(x = xtest, y = bart_sin$yhat.test.mean), color = "blue") +
+  geom_line(aes(x = xtest, y = 2*sin(xtest) + xtest), color = "green") +
+  labs(title = "BART vs. True Function",
+       x = "x",
+       y = "2sin(x)+x") +
+  theme_minimal()
+
+ggsave("outputs/sin_plot.pdf", plot_sin, width = 6, height = 4)
+
+
+
+plot_sigma <- ggplot(data = data.frame(samples = 1:1100, sigma = bart_sin$sigma), aes(x = samples, y = sigma)) +
+  geom_line() +
+  labs(title = "Estimated Sigma vs. Samples",
+       x = "Samples",
+       y = "Sigma") +
+  theme_minimal()
+
+ggsave("outputs/sigma_plot.pdf", plot_sigma, width = 6, height = 4)
+
+small <- sample(1:1000, 200)
+xsmall <- x[small]
+ysmall <- y[small]
+train_small <- sample(1:200, 100)
+xtrain_small <- xsmall[train_small]
+ytrain_small <- ysmall[train_small]
+xtest_small <- xsmall[-train_small]
+ytest_small <- ysmall[-train_small]
+
+bart_sig_s1 <- gbart(xtrain_small, ytrain_small, x.test = xtest_small, sigest = 100)
+bart_sig_s2 <- gbart(xtrain_small, ytrain_small, x.test = xtest_small, sigest = 20)
[…]

### 2024-12-07 code/tempCodeRunnerFile.r

@@ -1,0 +1,1 @@
+ytrain

### 2024-12-07 code/ code.r
finalizing code

@@ -3,6 +3,7 @@ packages <- c(
     "ggplot2", 
     "dplyr",
     "tree",
+    "xtable",
     "BART",
     "gbm", # For Generalized Boosted Regression Models
     "randomForest", # For Random Forest,
@@ -143,6 +144,9 @@ evaluate_models <- function(ntree) {
     time = c(bart$time, rf$time, gbm$time, bag$time)
   ))
 }
+
+df_100 <- evaluate_models(1000)
+print(xtable(df_100), type = "latex", file = "outputs/results_100.tex")
 
 run_results <- FALSE
 

### 2024-12-07 paper/paper.tex
Add work on afternoon

@@
 
-\section{Data Application: Predicting the age of an Abalone}
+\section{Data Application}
+
+We now move to applications of the theoretical insights derived in the previous sections.
+Firstly, we will focus on simulated data, and then we will cover an application using real word data on abalones.
+
+\subsection{Simulated data}
+We simulate a dataframe with the following generating non-linear function:
+\begin{equation}
+  y = 2\sin(x)+x+Z \quad Z \sim \mathcal{N}(0,1)
+\end{equation}
+The dataframe is split in two smaller dataframes to obtain training and testing data.
+
+We first run a single decision tree on the simulated data and we obtain as a result 5 terminal nodes and a 
+residual mean deviance of 1.164.
+
+We now fit a BART model with default parameters:  MCMC have 1100 iterations in total of which 100 are burn-in.
+We see from figure \ref{plot_sin1} that the BART model is able to fit correctly this non-linear trend.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.7\textwidth]{../outputs/sin_plot.pdf}
+  \caption{Plot of the estimates (blue) against observations (black) and generating function(green)}
+  \label{plot_sin1}
+\end{figure}
+
+To check convergence of the posterior, we plot in figure \ref{plot_sin2} the posterior draws of $\sigma$, the uniquely identified parameter.
+We observe that the MCMC draw converges quickly to the true parameter, although they oscillate with some autocorrelation due to the posterior sampling procedure.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.7\textwidth]{../outputs/sigma_plot.pdf}
+  \caption{Value of $\sigma$ over MCMC iterations}
+  \label{plot_sin2}
+\end{figure}
+
+To check robustness of BART to different prior specifications, we consider different values for the prior mean of $\sigma$,  that are 100, 0.001 and 1 (the true one).
+With a large $\sigma$, we observe in plot \ref{plot_sin3} underfitting of the data.
+With a small $\sigma$, holding all the other parameter constant, the regularization prior is still strong enough to avoid the overfitting.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.7\textwidth]{../outputs/sin_plot_diff_sigma.pdf}
+  \caption{DIfferent prior mean of $\sigma$}
+  \label{plot_sin3}
+\end{figure}
+
+\subsection{Predicting the age of an Abalone}
 
@@
 
-We use data Abalone characteristics data coming from the UC Irvine Machine Learning Repository \parencite{warwicknashAbalone1994a}, that is described in table \ref{table1}.
-No data manipulation is necessary, aside from creating a new variable, \( \text{Age} = \text{Rings} + 1.5 \), and removing the Rings column.
+We use data abalone characteristics data coming from the UC Irvine Machine Learning Repository \parencite{warwicknashAbalone1994a}, that is described in table \ref{table1}.
+No data manipulation is necessary, aside from creating a new variable, \( \text{Age} = \text{Rings} + 1.5 \), and removing the column identifying the number of rings.
 
@@
 
-We run BART from \cite{mccullochBARTBayesianAdditive2024}; Boosting from \cite{ridgewayGbmGeneralizedBoosted2024}; Random Forest and Bagging from \cite{breimanRandomForestBreimanCutlers2024}.
-We use half of the sample to fit $\hat{y}$, that is the Age variable, and half the sample to give us out of bag Mean Squared Errors (MSE) for predicting the same variable.
+We run BART from \cite{mccullochBARTBayesianAdditive2024}; Boosting from \cite{ridgewayGbmGeneralizedBoosted2024}; Random Forest and Bagging from \cite{breimanRandomForestBreimanCutlers2024}.Our aim is to compare the predictive performance of those methods.
+In fact,  we use half of the sample to fit $\hat{y}$, that is the age variable, and half the sample to give us out of bag Mean Squared Errors (MSE) for predicting the same variable.
 Models are run for various values of T, and for each value, both the mean squared error (MSE) of out-of-sample predictions (shown in plot \ref{plot_mse}) and the computation time (depicted in plot \ref{plot_time}) are recorded.
+
+We observe with a number of trees estimated larger than 500, the BART model is the best performing one, and it reaches also the best performance overall when run with 500 iterations.
+Although it is the best performer, it is also the model that takes the most to be run, and time of computations scales linearly with the number of trees that are fitted.
+
+\begin{table}[ht]
+  \centering
+  \begin{tabular}{lcc}
+  \toprule
+  Model           & MSE & Time (s) \\
+  \midrule
+  BART            & 4.45    & 66.7  \\
+  Random Forest   & 4.53    & 4.74  \\
+  Boosting        & 4.87    & 0.220 \\
+  Bagging         & 4.54    & 12.4  \\
+  \bottomrule
+  \end{tabular}
+  \caption{Model performance with $T=500$}
+  \end{table}
 
@@
 
-\newpage
-
 \printbibliography

### 2024-12-07 code/ code.r

@@ -34,6 +34,9 @@ ytest <- y[-train]
 
 # fit a bart model
 bart_sin <- gbart(xtrain, ytrain, x.test = xtest)
+# fit a tree model
+tree_sin <- tree(ytrain ~ xtrain)
+print(summary(tree_sin))
 
 
 # plot the results using ggplot2
@@ -69,8 +72,8 @@ xtest_small <- xsmall[-train_small]
 ytest_small <- ysmall[-train_small]
 
 bart_sig_s1 <- gbart(xtrain_small, ytrain_small, x.test = xtest_small, sigest = 100)
-bart_sig_s2 <- gbart(xtrain_small, ytrain_small, x.test = xtest_small, sigest = 20)
-bart_sig_s3 <- gbart(xtrain_small, ytrain_small, x.test = xtest_small, sigest = 1)
+bart_sig_s2 <- gbart(xtrain_small, ytrain_small, x.test = xtest_small, sigest = 1)
+bart_sig_s3 <- gbart(xtrain_small, ytrain_small, x.test = xtest_small, sigest = 0.001)
 
 plot_diff_sigma <- ggplot(data = data.frame(xtest_small, ytest_small), aes(x = xtest_small, y = ytest_small)) +
   geom_point() +  

### 2024-12-07 paper/paper.tex
Work online by vincenzo and guido

@@
 \section{Introduction}
+summary of all
+\cite{chipmanBARTBayesianAdditive2010}
 
-Bayesian Regression Additive Trees (BART) were first introduced by \cite{chipmanBARTBayesianAdditive2010}
+\section{Decision trees}
+CART (Classification And Regression Trees) trees, where first introduced by (\cite{Breiman et al.
+1984}) and represent a statistical method for constructing decision trees for both regression and classification problems.
+A CART is an upside-down tree, in the sense that leaves are at the bottom of the tree.
+Terminal nodes, that is nodes without descendants, are called leaves, whereas the other nodes are called internal nodes.
+Each internal node has two child nodes, hence the tree is said to be binary.
+The segments of the trees that connect the nodes are called branches.
+The internal nodes are labeled by a condition while the leaves by a value of the dependent variable.
+Hence, when a tree is given, is easy to do forecasting.
+Indeed, to get the predicted value $\hat{y}$  for a given covariate vector $x$, is enough to go sequentially answer the decision rules starting form the root of the tree.
+For each decision rule, if the condition of is verified you go to one side of the node, if not, you go to the other.
+This process continues until the terminal node is reached.
+The set of bottom nodes (or leaves) of the trees give a partition of the space of covariates: it partitions the space of covariates $x$ into $J$ distinct non-overlapping regions $R_1, \ldots, R_J$, each corresponding to a terminal node of the tree.
 
-\section{Non-Bayesian Tree Based Methods}
-
-Tree based method are more usefull compared to the linear regression model when the relation to be model is complex or non-linear. Indeed, the standard linear model is defined as:
+Regression trees are very different models compared with the more common linear regression models.
+Suppose there are $n$ observations of $y$ and $\mathbf{x} = (x_1, \ldots, x_p)$.
+Overall, covariates and output are related by the equation
+\begin{equation}
+    Y = f(X) + \epsilon, \quad \epsilon \sim \mathcal{N}(0, \sigma^2), \label{regression}
+\end{equation}
+where we assumed normality of the error for simplicity.
+Both regression trees and linear models attempt at making inference about the unknown function \( f \), which is left completely unrestricted, using the \( p \)-dimensional vector of inputs \( x = (x_1, \dots, x_p) \) and the output $y$.
+How $f(X)$ is proxied depends on the model chosen. On the one hand, linear regression models assume a functional form of the type:
 \begin{equation}
@@
 \end{equation}
-Wheras the decision tree model as:
+On the other hand, regression trees approximate the function $f(X)$ with a step-wise form: 
 \begin{equation}
-  f(X) = \sum_{m=1}^M c_m \cdot 1(X \in R_m)
+  f(X) = \sum_{j=1}^J \mu_j \cdot 1(X \in R_j) \label{stepwise}
 \end{equation}
-DESCRIBE BETTER THIS EQUATIONS
 
-We take most of our arguments for this section from the \cite[Chapter~8]{jamesIntroductionStatisticalLearning2021}.
+where $R_1, ..., R_J$ represent the partition of the covariates space generated by the leaves of the tree.
+The preferable method depends on the particular problem we are presented with.
+Indeed, if the relation among explanatory variables and the dependent variable is well approximated by a linear model, it is likely that a linear regression will outperform a regression tree model.
+On the contrary, if there is a highly non-linear and complex relation between the explanatory variables and the dependent one, then a regression tree is more likely to outperform a simple linear regression model.
+Moreover, when fitting a regression tree instead of a linear model, we do not have to care about the scale of covariates.
+From now on our analyses will only focus on models of the form of \eqref{stepwise}.
+
+We take most of our arguments for this section from the \cite[Chapter~8]{jamesIntroductionStatisticalLearning2021}, and from \cite{Random Forest with R}.
 
@@
 
-Tree based methods that we cover in this paper are all based on decision trees.
-Decision trees can be applied both to regression or classification problems, and are thus named accordingly to their application.
-Regression trees in particular are build in two steps.
+Regression trees are built in two steps.
+First, the predictor space (the set of all possible values of the vector of observables $X_1, ...., X_p$) is split into $J$ regions $R_j$ that are distinct and non-overlapping.
+Secondly, given a new observation $x$ that falls in a region $R_j$, the model predicts the corresponding outcome variable $\hat{y}$ to be the mean of the outcome variables corresponding to the observations in the training set that fall in that same region.
 
-First, the prediction space (the set of all possible values of the vector of observables) is split into $J$ regions $R_j$ that are distinct and non-overlapping.
-Secondly, a prediction for the outcome variable is made for each observation in each $R_j$ regions, as the mean of the training observations in that same regions.
-
-For a predictor space of $P$ elements the regions are $p$ dimensional boxes, that are build by minimizing the Residual Sum Squared, as the equation \ref{rss_tree} shows.
+To begin, we have to construct the regions $R_1,..., R_j$.
+For simplicity, regions are chosen as $p$-dimensional boxes, where $p$ is the number of regressors.
+The partition is chosen is such a way to minimize the RSS:
 \begin{equation}
-  R_1,...,R_j=\operatorname*{argmin}_{R_j}\sum^J_{j=1}\sum_{i\in R_j}\left(y_i-\hat{y}_{r_j}\right)^2
+  \{R_1,...,R_J \}=\operatorname*{argmin}_{ \{R_j\}_{j=1}^J}\sum^J_{j=1}\sum_{i\in R_j}\left(y_i-\hat{y}_{R_j}\right)^2
   \label{rss_tree}
 \end{equation}
+where $\hat{y}_{R_j}$ is the mean of the training observation in the $j$-th box.
 
-Instead of considering all the possible partitions of the predictor space, a top-down approach is used.
-First, all observations are considered such that they belong to the same region.
-Then, this process chooses the best split minimizing the RSS irrespective of future outcomes.
-This is then repeated for a number of times, until a stopping criterion is reached (usually a minimum number of observations in all the regions).
+It is computatioanlly infeasible to do the optimization problem in \eqref{rss_tree}.
+Instead of considering all the possible partitions of the predictor space, a top-down, greedy approach, known as recursive binary splitting is used.
+The approach is \texttt{top-down} since it begins at the top of the tree, and then sequentially splits the predictor space; each split is then indicated by two new branches.
+Moreover, the process is \textit{greedy}, since for each step we make that split that generates the greatest reduction of the RSS at that particular step, and we do not choose the best sequence of steps looking ahead.
+\\To perform recursive binary splitting, the algorithm considers the set of all predictors $X_1, \ldots, X_p$, and the set of all possible values of the cutpoint $s$, and then choose the predictor $X_j$ and the cutpoint $s$ such that the branching of the tree corresponding tot the decision rule:  $$\{X \mid X_j < s\} \quad \text{and} \quad \{X \mid X_j \geq s\}$$ minimizes the RSS.
+Then, we repeat the process within the resulting subregion, looking again for the predictor and the cutpoint such that the subsequent decision rule minimize the RSS.
+It is important to note that this time we do not split the entire predictor space but one of the two previously identified regions.
+This process continues until a stopping condition is reached; the more common requires a minimum number of observation within each splitting nodes.
 
-As the regions visualizations are complex when more than 2 regressors are present, a tree based graph is used.
+\subsection{Pruning}
 
-The process described above will likely tend to overfit the data.
-There are numerous solutions to this, the most simple one being increasing the number of AND GENERALLY THE ENSAMBLING MODELSobservations that rise the stopping rule.
-A valid alternative is to grow a very large tree, and then reduce it to obtain a smaller subtree.
-This is known as tree pruning.
+Recursive binary splitting to build a regression tree  is likely to produce good prediction on the training-set but overfit the training data, leading to poor out-of-sample performance.The reason is that the above mentioned procedure is likely to give us a complex tree, which has probably a low bias but high variance in prediction.
+Constraining ex ante the size of trees would be a short-sighted solution, because one could possibly stop too soon, leaving potentially good splits further down in tree unachieved.
+\\A better strategy is pruning an already complex tree.
+This amounts to search for a good intermediate between two extreme cases: a fully developed tree which has high variance but low bias, and a tree consisting of only the root, with high bias but low variance.
+The strategy consists in growing a very large tree $T_0$, and then prune it back in order to obtain a subtree. To do so, we use cost complexity pruning, also know as weakest link pruning.
 
-ADD COST COMPLEXITY PRUNING
+Consider a sequence of subtrees of $T_0$, $\{T\}_\alpha$ indexed by a non-negative tuning parameter $\alpha$.
+For each value of $\alpha$, there is a subtree $T \subset T_0$ such that
+$$\sum_{m=1}^{|T|} \sum_{i : x_i \in R_m} \left( y_i - \hat{y}_{R_m} \right)^2 + \alpha |T|$$
+is as small as possible, where $|T|$ represents the number of terminal nodes of the tree T, $R_m$ is the subset of the predictor space corresponding to the $m$-th terminal node, $\hat{y}_{R_m}$ is the predicted response associated with $R_m$.
+The tuning parameter $\alpha$ controls the trade-off between a sub-tree complexity and and its fit to the training data, the higher the $\alpha$, the more terminal nodes a tree has, the smaller a subtree is.
+The value for $\alpha$ can be chosen using a validation set or trough cross-validation.
 
-One of the main disadvantages of this setting, is that trees are non-robust.
+\subsection{Pros and Cons of decision trees}
+Overall, CART offers much greater flexibility in handling the relation between covariates and output than linear models do.
+This comes from the fact that fitting a regression tree amounts to estimate a stepwise relation between $y$ and the covariates $x$ that can proxy well non-linear patterns.
+\\\\Greater flexibility of CART comes in handy also when dealing with missing values.
+CART offers an effective way to overcome this issue, by constructing surrogate variables.
+Exploiting correlation between predictors, CART may decrease the effect of missing data.
+
+So far we considered only binary splits, thus an interesting question would be if a model with more than two splits at each note is a valid idea.
+However, it turns out that in general is not a good strategy.
+The issue is that multiple splits fragment the dataset too quickly.
+Moreover, since multiple splits can be achieved by a series of binary splits, the latter approach is preferred.
+
+The interpretability of CART trees has been one of the driving forces of its success.
+Indeed, thanks to its structure, is straightforward to understand why, for a given $x$, we expect a particular value of $y$, following a series of decision rules.
+Small trees are visually easy to interpret.
+Moreover, trees build in such a way are a computationally fast method, that easily scales with large datasets.
+
+However, one of the main disadvantages of this setting, is that trees are non-robust.
 In fact a small change in the data can lead to a large change in the estimation.
-This follows from the fact that Regression Trees suffer from high variance.
-This is indeed quite problematic and the main reason that simple regression trees are not used by practitioners.
-Ensemble methods aim at solving this issue.
-Those methods in the context of trees combine simpler models (called weak learners), whichy may lead to mediocre predictions, in a single more powerful and stable model.
-QUelli di cui sotto sopra abuiao detto sono tutti ensable models.
+This follows from the fact that regression trees suffer from high variance.
+The main reason of this instability is the hierarchical structure of the process as the effect of an error in one of the first nodes is propagated to the other nodes below it.
+This is heightened by the fact that single trees tend to overfit, and spurious correlations in the training data can be mistaken as structural patterns and included in the decision rules.
+This may lead to bad out-of-sample predictive performance.
+High variance and tendence to overfit are the main reasons that simple regression trees are not used by practitioners.
+Instead, ensemble methods are generally preferred.
+
+\section{Ensemble methods}
+
+An ensemble method is an approach that aims at combining many ensemble methods that combine the fit from many (hundreds, thousands) simpler models to get an overall
+ predictor. These simpler models are often called weak learners, since often lead to mediocre prediction if considered singularly.
+All the models we are going to discuss, namely Bagging, Boosting, Random Forest and Bayesian Additive Regression Trees are ensemble methods, for which the simpler models consist of regression trees.
 
 \subsection{Bagging}
-Bagging is a generic method that aims at reducing the variance of a statistical learning method.
-Bagging is based on the fact that if we have independent observations $(Z_1,...,Z_n)$ with same variance $\sigma$, then the vairance of the sample mean of the observation will be lower, as it is $V(\bar{Z})=\frac{\sigma}{n}$.
+Bagging, or Bootstrap Aggregation, is a general method that aims at reducing the variance of a statistical learning method.
+The method is based on the fact that, given independent observations, averaging over them reduces the variance.
+For the case of regression trees, we would like to be able to take many different training sets from the population, to build a different trees for each training set and then to average the resulting predictions.
+However this is very unfeasible since we do not have access to many training sets.
+Hence, we resort to bootstrap.
+Bootstrapping consists of taking repeated samples, with replacement, from the single available training set. In such a way, we are able to construct $B$ different bootstrapped training sets.
 
-Bagging uses a bootstrap sampling procedure to obtain different samples from the data, and then fits a tree for each of the bootstrapped samples.
-The single estimates for each sample are then averaged, as equation \ref{eq_bagging} shows.
+Bagging consists in taking $B$ bootstrap samples from the training data, each of the same size as the training data, and to fit a large tree to each bootstrap sample.
+Then it combines the results from each of the $B$ trees to get an overall prediction.
+The single predictions for the target funciton $f(X)$ for each sample are averaged, as
 \begin{equation}
@@
 \end{equation}
-Where $B$ is the number of samples of the bootstrapping procedure.
+where $B$ is the number of samples of the bootstrapping procedure.
+
+In bagging the trees are grown big and are not pruned. Therefore, each single tree has high variance and low bias. Averaging these different trees reduce the variance.
+
+The drawback of bagging is that we lose interpretability to improve accuracy. When we bag a large number of trees, we can no longer represent the model with a single tree.
+
+\subsection{Random Forest}
+A possible problem with bagging is due to the lack of randomization between trees.
+Indeed, since trees are separately fitted on the bootstrapped training set, it may happen that they share to great amount the same decision rule.
+This is the case, for example, if there is very strong predictor in the dataset.
+Hence, in the collection of bagged trees a vast majority of the trees will use this strong predictor in the top split and tehy will all be hihgly correlated among themselves.
+\\Correlation among the trees in turn decrease the effectiveness of averaging, as it does not lead to a large reduction in variance, making the entire bagging algorithm less useful. 
+\\\\ Random Forests starts from bagging and adds another kind of randomization to reduce correlation among trees.
+Rather than searching over all the covariate space for each decision rule, as the greedy build of the big trees imposes, in random forests we randomly sample a subset of $m$ covariates to search over each time we make a split.
+Therefore, the split is allowed to consider only one of those m predictors.
+Moreover, at each split is considered a new sample of m predictors.
+This makes the big trees explore more the space of models
+\\\\Usually, we choose $m \approx \sqrt{p}$.
+On average, $(p - m)/ = p$ of the splits will not even consider the strong predictor, and so other predictors will have more of a chance.
+Therefore, we can think of this process as decorrelating the trees, making the average of the resulting trees less variable and hence more reliable.
 
@@
 
-Similar to bagging, boosting is not specific to regression trees, it is rather a generic approach.
-Said so, it is quite effective when used on trees.
-In fact ,it approaches the overfitting problem but without using bootstrapping, as bagging does.
-Boosting relies on fitting trees sequentially, as algorithm \ref{alg_boosting} shows.
-
-The number of sequential trees to fit, as chosen by the parameter $B$ in algorithm  \ref{alg_boosting}, if chosen too large, can cause overfitting.
-The learning rate $\lambda$ is another tunable parameter.
-Often small values are chosen ($\lambda=10^{-2}$ or even smaller).
-
-\begin{algorithm}
+Differently from bagging and random forests, boosting does not involve fitting independent trees from many bootstrap samples; instead, boosting relies on fitting trees sequentially, that is, using information from previously grown trees.
+This is shown algorithm \ref{alg_boosting} shows. 
+\begin{algorithm} [h]
   \caption{Boosting for Regression Trees}
@@
 \end{algorithm}
+
+By fitting sequentially new trees on the residual of the previous tree, voosting improves at each iteration the fit marginally.
+Indeed, since there is a constraining parameter $\lambda$, each tree is rather small, with just a few terminal nodes.
+By
+fitting small trees to the residuals, we slowly improve $\hat{f}$ in areas where it
+does not perform well. It is important to mention that in boosting, unlike in bagging, the construction of each tree strongly depends on the trees that have already grown.
+\\\\The model accounts for three different tuning parameters.
+The number of of trees $B$, chosen trough cross-validation.
+Interestingly, unlike bagging and random forests, boosting can suffer of overfitting if $B$ is too large.
+The shrinkage parameter $\lambda$, a small positive number.
+Common values for $\lambda$ are 0.01 or 0.001, depending on the problem at hand.
+However, very small values of $\lambda$ often require a very large B to achieve good performances.
+Lastly, $d$ represents the number of splits in each tree.
+Often $d=1$ works well.
+However in this case the tree is called stump since it contains only a single split.
+
+\section{Bayesian additive regression trees (BART)}
+The idea for Bayesian Additive Regression Trees (BART) was first developed by Chipman, George, McCulloch (2010 AOAS).
+Like other ensemble methods (bagging, boosting, random forest), to solve this regression problem \eqref{regression}, BART approximates \( f(x) = \mathbb{E}(y \mid x) \) by a sum of $m$ regression trees \( h(x) \equiv \sum_{j=1}^m g(x, T_j, M_j)\), where each \( g \) is a regression tree function that given a tree $\theta_j$ to each observation $x$  assigns a leaf $\mu_{ij}$ of that regression tree.
+\\\\The BART Ensemble Model is therefore
+\[
+y = \sum_{j=1}^m g(x; T_j, M_j) + \epsilon, \quad \epsilon \sim N(0, \sigma^2) \tag{3}
+\]
+where for each binary regression tree \( T_j \) and its associated terminal node parameters \( M_j \), \( g(x; T_j, M_j) \) is the function which assigns \( \mu_{ij} \in M_j \) to \( x \).
+Under \((3)\), \( \mathbb{E}(y \mid x) \) equals the sum of all the terminal node \( \mu_{ij} \)'s assigned to \( x \) by the \( g(x; T_j, M_j) \)'s.
+\\\\Like the other ensemble methods, to avoid overfitting each tree of the summation must be made a weak learner, explaining only part of the result.
+To do so, while boosting uses a shrinkage parameter $\lambda$, BART uses a regularization prior:
+\[
+p((T_1, M_1), (T_2, M_2), \ldots, (T_m, M_m), \sigma)
+\]
+The prior is set in a way to regularize the fit by keeping the individual tree effects small.
+It forces each tree to explain only a limited subset of the relationships between the covariates and the predictor variable (it is a marginal addition to the overall fit).
+By weakening the \( g_j \) effects, BART ends up with a sum of trees, each of which explains a small and different portion of \( f \).
+
+In effect, the \( g_j \)'s become a dimensionally adaptive random basis of weak learners. Bayesian Nonparametrics: Lots of parameters (to make model flexible).
+A strong prior to shrink towards simple structure (regularization).
+BART shrinks towards additive models with some interaction:Fit becomes the cumulative effort of many weak learners.Note that BART works with an ensemble of trees where each tree has an unknown size.
+As a result, the model is high and random dimensional.
+Note that The number of trees $m$ can be much larger than the sample size $n$.
+Indeed, $g(x; T_1, M_1), g(x; T_2, M_2), \ldots, g(x; T_m, M_m)$ is a highly redundant \textit{over-complete basis} with many parameters.
+Finally, dynamic Random Basis Elements:
+\[
+g(x; T_1, M_1), \ldots, g(x; T_m, M_m) \text{ are dimensionally adaptive.}
+\]
+It is a highly redundant “over-complete basis” with many many parameters.
+
+Finally, we use MCMC to get draws from the high-dimensional posterior so that  we can stochastically search the model space and gauge our uncertainty.
+To fit the sum-of-trees model, BART uses a tailored version of Bayesian backfitting MCMC \cite{hastie2000} that iteratively constructs and fits successive residuals.
+Inferences obtained from BART are based on successive iterations of the backfitting algorithm, which are effectively an MCMC sample from the induced posterior over the sum-of-trees model space.
+A single posterior mean estimate of \( f(x) = \mathbb{E}(Y \mid x) \) at any input value is obtained by a simple average of these successive sum-of-trees model draws evaluated at \( x \).
+Further, pointwise uncertainty intervals for \( f(x) \) are easily obtained from the corresponding quantiles of the sample of draws.
+
+\\\\Overall, Bart's characteristics are in how it weakens the individual trees by instead using a prior, and how it performs the iterative fitting by instead using Bayesian backfitting on a fixed number of trees.
+Conceptually, BART can be viewed as a Bayesian nonparametric approach that fits a parameter-rich model using a strongly influential prior distribution.
+Overall, BART is a highly-dimensional (each tree has unknown size) nonlinear model inspired by boosting and tree ensembles.
+the main difference is that, being Bayesian, it uses a regularization prior to shrink towards a simple model.
+It uses MCMC to get draws from the high dimensional posterior so that one can stochastically search the  model space and gauge the uncertainty.
+
+\subsection{The BART Model}
+The BART model consists of two parts: a sum-of-trees model and a regularization prior on the parameters of that model.
+
+\subsubsection{A Sum-of-Trees Model}
+Denote by \( T \) the tree structure, consisting of a set of interior node decision rules and a set of $b$ terminal nodes.
+The decision rules are binary splits of the predictor space of the form \( \{x \in A\} \) vs \( \{x \notin A\} \), where \( A \) is a subset of the range of \( x \).
+We only consider rules based on the single components of \( x = (x_1, \dots, x_p) \) and of the form \( \{x_i \leq c\} \) vs \( \{x_i > c\} \) for continuous \( x_i \).
+Let \( M = \{\mu_1, \mu_2, \dots, \mu_b\} \) denote a set of parameter values associated with each of the \( b \) terminal nodes of \( T \).
+Given the way it is constructed, the tree is a full binary tree, that is, each node has exactly zero or two children.
+\\\\ Let \( g(x; T, M) \) denote the regression tree function which assigns a \( \mu_i \in M \) to \( x \).
+Each \( x \) value is associated with a single terminal node of \( T \) by the sequence of decision rules from top to bottom, and is then assigned the \( \mu_i \) value associated with this terminal node.
+\\\\We denote by \( T \) the binary tree structure itself—the various binary split decision rules of the form \( \{x_q < c\} \) versus \( \{x_q \geq c\} \), \( c \in \mathbb{R} \).
+The vector of parameters associated with \( T \), i.e., the collection of terminal node parameters, is denoted by \( M = \{\mu_1, \ldots, \mu_b\} \), where \( b \) is the number of terminal nodes.
+Defining a regression tree as \( g(x; T, M) \), we can view \( g(x; T, M) \) as a function that assigns the conditional mean \( \mathbb{E}[y \mid x] \) to the parameter \( \mu_i \in M \), i.e., \( \mu_i = g(x; T, M) \rightarrow \mathbb{E}[y \mid x] \) via binary decision rules denoted as \( T \).
+\\\\Thus,
+\begin{equation}
+Y = g(x; T, M) + \epsilon, \quad \epsilon \sim N(0, \sigma^2)
+\end{equation}
+is a single tree model. Under (3), the conditional mean of \( Y \) given \( x \), \( \mathbb{E}(Y \mid x) \), equals the terminal node parameter \( \mu \) assigned by \( g(x \mid T, M) \).
+With this notation, the sum-of-trees model can be more explicitly expressed as
+\begin{equation}
+Y = \sum_{j=1}^m g(x; T_j, M_j) + \epsilon, \quad \epsilon \sim N(0, \sigma^2) \label{likelihood}
+\end{equation}
+where for each binary regression tree \( T_j \) and its associated terminal node parameters \( M_j \), \( g(x; T_j, M_j) \) is the function which assigns \( \mu_{j} \in M_j \) to \( x \).
+Under (4), \( \mathbb{E}(Y \mid x) \) equals the sum of all the terminal node \( \mu_{j} \)'s assigned by the \( g(x \mid T_j, M_j) \)'s.
+When the number of trees \( m > 1 \), each \( g(x; T_j, M_j) \) here is merely a part of \( \mathbb{E}(Y \mid x) \), unlike the single tree model.
+\\\\We can see that the quantity that is being ``summed'' and eventually allocated to \( \mathbb{E}(y \mid x) \) is not the regression tree or tree structure, but the value that each \( j \)th tree structure assigns to the subject.
+This is one way to think of a sum of regression trees.
+It allocates a sum of parameters \( \mu_{j} \) to \( \mathbb{E}[y \mid x] \) of the subject.
+
+\\\\A sum-of-trees model is an additive function, a sum of step functions of the individual components of \( x \).
+With a large number of trees, a sum-of-trees model gains increased representation flexibility which, as we'll see, endows BART with excellent predictive capabilities.
+This representational flexibility is obtained by rapidly increasing the number of parameters.
+Indeed, for fixed \( m \), each sum-of-trees model (4) is determined by \( (T_1, M_1), \ldots, (T_m, M_m) \) and \( \sigma \), which includes all the bottom node parameters as well as the tree structures and decision rules.
+Further, the representational flexibility of each individual tree leads to substantial redundancy across the tree components.
+Indeed, one can regard \( \{g(x; T_1, M_1), \ldots, g(x; T_m, M_m)\} \) as an "overcomplete basis" in the sense that many different choices of \( (T_1, M_1), \ldots, (T_m, M_m) \) can lead to an identical function \( \sum_{j=1}^m g(x; T_j, M_j) \).
+
+\\\\\textbf{Nonlinearity of BART.} From this simple example, we can see how BART handles nonlinearity.
+Each single regression tree is a simple stepwise function or ANOVA model.
+When we sum regression trees together, we are actually summing together these ANOVA models or stepwise functions, and as a result, we eventually obtain a more complicated stepwise function that can approximate the nonlinearities in the main effect and multiple-way interactions.
+
+\subsubsection{A regularization prior}
+In the examples above, we have taken the trees as a given, including which variables to split on, the splitting values, and the mean parameters at each terminal node.
+In practice, each \( g(x; T_j, M_j) \) is unknown.
+We therefore need a prior distributions.
+
+Hyperparameters of $p$ are set so that both the \(T_j\)'s and the 
+\(\mu_{ij}\)'s are likely to be small. Thus, $p$ keeps the contribution of each \(g(x; T_j, M_j)\) small, to 
+explain only a small portion of the fit. Bart builds up the fit, by adding up tiny bits of fit.  The prior of BART shrinks the model towards additive models with some interaction.
+
+\\\\The BART model specification is completed by imposing a prior over all the parameters of the sum-of-trees model, namely, \( (T_1, M_1), \ldots, (T_m, M_m) \) and \( \sigma \).
+There exist specifications of this prior that effectively regularize the fit by keeping the individual tree effects from being unduly influential.
+Without such a regularizing influence, large tree components would overwhelm the rich structure of (3), thereby limiting the advantages of the additive representation both in terms of function approximation and computation.
+
+Chipman et al.
+proposed a prior formulation in terms of just a few interpretable hyperparameters which govern priors on \( T_j \), \( M_j \), and \( \sigma \).
+When domain information is not available, the authors recommend using an empirical Bayes approach and calibrating the prior using the observed variation in \( y \).
+Or at least to obtain a range of plausible values and then perform cross-validation to select from these values.
+
+\textbf{Prior independence and Symmetry}
+The prior distribution for \eqref{likelihood} is \( p[(T_1, M_1), \ldots, (T_m, M_m), \sigma] \).
+n order to simplify the specification of the regularization prior, Chipman assumes that  that \( \{(T_1, M_1), \ldots, (T_m, M_m)\} \) are independent of each other and of  \( \sigma \).
+Then, the prior distribution can be written as:
+
+\[
+p[(T_1, M_1), \ldots, (T_m, M_m), \sigma] = p[(T_1, M_1), \ldots, (T_m, M_m)] \, p(\sigma).
+\]
+
+\[
+= \left[ \prod_{j=1}^m p(T_j, M_j) \right] \, p(\sigma).
+\]
+
+\[
+= \left[ \prod_{j=1}^m p(M_j \mid T_j) \, p(T_j) \right] \, p(\sigma).
+\]
+Consider the following.
+Given $T_j$, let  \( b_j \)  be the number of bottom nodes in the tree \( T_j \).
+Then \( M = \{\mu_1, \mu_2, \ldots, \mu_{b_j}\} \) is the vector of terminal node parameters associated with \( T_j \).
+We assume that each node parameter \( \mu_{ji} \) is iid of each other.
+This implies:
+\[
+p(M_j \mid T_j) = \prod_{i=1}^{b_j} p(\mu_{ij} \mid T_j) \quad \text{(5)}
+\]
+
+where \( \mu_{ij} \in M_j \) where \( \mu_i \sim p(\mu) \), iid.
+Finally we can write the prior distribution for \eqref{likelihood} as:
+\[
+p[(T_1, M_1), \ldots, (T_m, M_m), \sigma]= \left[ \prod_{j=1}^m \left( \prod_{i=1}^{b_j} p(\mu_{ji} \mid T_j) \right) \right] \, p(\sigma) \, p(T_j).
+\]
+Under the independence assumption, we only need to specify \( p(T_j) \), \( p(\mu_{ij} \mid T_j) \), and \( p(\sigma) \).
+Note that we want to have each $T$ small, each $\mu$ small,  and a "nice " $\sigma$  (e.g.
+smaller than the least squares estiamte).
+Indeed, we refer to $p$ as the regularization prior as it restraints the overall fit.
+In addition, it keeps the contribution of each $g(x;T_i,M_i)$ model component small.
+\\\\
+\textbf{The $T_j$ prior}
+The trees $T_j$ are assumed to be iid, so that it is sufficient to define a single $p(T_j)$. Chipman specifies a process 
+ that can be used to fraw a tree from the prior. The \( T_j \) prior, \( p(T_j) \), is specified by three aspects:
+\begin{enumerate}
+    \item The probability that a node at depth \( d = (0, 1, 2, \ldots) \) is nonterminal, given by:
+    \[
+    \alpha (1 + d)^\beta \quad \text{with} \quad \alpha \in (0, 1) \quad \text{and} \quad \beta \in [0, \infty).
+    \]
+    Node depth is defined as the distance from the root.
+Thus, the root itself has depth 0, its first child node has depth 1, etc.
+This prior controls the tree depth.
+For a sum-of-trees model with \( m \) large, we want the regularization prior to keep the individual tree components small.
+To do that, we usually use \( \alpha = 0.95 \) and \( \beta = 2 \).
+This makes nonnull but small tree more likely.
+Nonetheless,  trees with many terminal nodes can be grown if the data demands it.
+
+    \item The distribution on the splitting variable assignments at each interior node. Usually, this is the uniform prior on available variables.
+
+    \item The distribution on the splitting rule assignment in each interior node, conditional on the splitting variable.
+Usually, this is the uniform prior on the discrete set of available splitting values.
+\end{enumerate}
+
+\textbf{The \( \mu_{i,j} | T_j \) Prior}
+We assumed that $\mu_{ij} | T_j$ are iid.
+For convenience, we first shift and rescale \( y \) so that the observed transformed values range from \( y_{\text{min}} = -0.5 \) to \( y_{\text{max}} = 0.5 \).
+Then the  prior chosen is:
+\[
+\mu_{i,j}| T_j \sim N(0, \sigma_\mu^2)
+\]
+where the zero mean is justified by the centering.  Note that the normal is a conjugate prior. This will simplify posterior inference via MCMC algorithm.
+How do we choose the key parameter \( \sigma_\mu \)?
+The key observation is that:
+
+\[
+f(x | T_j, M_j) = \mu_1 + \mu_2 + \cdots + \mu_{b_j}, \quad \mu_i \sim N(0, \sigma_\mu^2)
+\]
+
+You can also just directly set the standard deviation of \( f(x) \):
+
+\[
+f(x) \sim N(0, \sigma_f^2)
+\]
+
+which is the same as just setting
+
+\[
+\sigma_\mu^2 = \frac{\sigma_f^2}{m}
+\]
+
+which gives us:
+
+\[
+f(x) | T \sim N(0, m\sigma_\mu^2)
+\]
+
+which does not depend on \( T \) or \( x \).
+
+\[
+\sigma_\mu = \frac{0.5 k}{\sqrt{m}}.
+\]
+This prior has the effect of shrinking the tree parameters \( \mu_{i,j} \) toward zero, limiting the effect of the individual tree components by keeping them small.
+Note that as \( k \) and/or \( m \) are increased, this prior becomes tighter and applies greater shrinkage to the \( \mu_{i,j} \).
+Chipman et al.
+(2010) found that a value of \( k \) between 1 and 3 yields good results.
+
+\\\\\\\\Or (\(^ {17}\)), we use the conjugate normal distribution \(N(\mu, \sigma^2)\), which offers tremendous computational benefits because \(\mu\) can be marginalized out.
+
+To guide the specification of the hyperparameters \(\mu\) and \(\sigma\), note that \(\mathbb{E}(Y \mid x)\) is the sum of \(m\) \(\mu_j\)'s under the sum-of-trees model, and because the \(\mu_j\)'s are a priori i.i.d., the induced prior on \(\mathbb{E}(Y \mid x)\) is \(N(\mu, \sigma^2)\).
+Note also that it is highly probable that \(\mathbb{E}(Y \mid x)\) is between \(y_{\text{min}}\) and \(y_{\text{max}}\), the observed minimum and maximum of \(Y\) in the data.
+
+The essence of our strategy is then to choose \(\mu\) and \(\sigma\) so that \(N(\mu, \sigma^2)\) assigns substantial probability to the interval \((y_{\text{min}}, y_{\text{max}})\).
+This can be conveniently done by choosing \(\mu\) and \(\sigma\) so that:
+
+\[
+m \mu - k \sigma / \sqrt{m} = y_{\text{min}} \quad \text{and} \quad m \mu + k \sigma / \sqrt{m} = y_{\text{max}}
+\]
+
+for some preselected value of \(k\). For example, \(k = 2\) would yield a 95\% prior probability that \(\mathbb{E}(Y \mid x)\) is in the interval \((y_{\text{min}}, y_{\text{max}})\).
+
+The strategy above uses an aspect of the observed data, namely, \(y_{\text{min}}\) and \(y_{\text{max}}\), to try to ensure that the implicit prior for \(\mathbb{E}(Y \mid x)\) is in the right "ballpark." That is to say, we want it to assign substantial probability to the entire region of plausible values of \(\mathbb{E}(Y \mid x)\) while avoiding overconcentration and overdispersion.
+We have found that, as long as this goal is met, BART is very robust to changes in the exact specification.
+Such a data-informed prior approach is especially useful in our problem, where reliable subjective information about \(\mathbb{E}(Y \mid x)\) is likely to be unavailable.
+
+For convenience, we implement our specification strategy by first shifting and rescaling \(Y\) so that the observed transformed \(y\) values range from \(y_{\text{min}} = -0.5\) to \(y_{\text{max}} = 0.5\), and then treating this transformed \(Y\) as our dependent variable.
+We then simply center the prior for \(\mu_j\) at zero (\(\mu = 0\)) and choose \(\sigma\) so that:
+
+\[
+k \sigma / \sqrt{m} = 0.5
+\]
+
+for a suitable value of \(k\), yielding:
+
+\[
+\mu_{ij} \sim N(0, \sigma^2) \quad \text{where} \quad \sigma = \frac{0.5}{k \sqrt{m}}
+\]
+
+\textbf{The \( \sigma \) Prior}
+We use the inverse chi-square distribution:
+
+\[
+\sigma^2 \sim \nu \lambda \chi^2_\nu
+\]
+
+Essentially, we calibrate the prior for the degree of freedom \( \nu \) and scale \( \lambda \) for this purpose using a rough data-based overestimate \( \hat{\sigma} \) of \( \sigma \).
+Two natural choices for \( \hat{\sigma} \) are:
+
+\begin{itemize}
+    \item The naive specification, where \( \hat{\sigma} \) is the sample standard deviation of \( y \).
+    \item The linear model specification, where \( \hat{\sigma} \) is the residual standard deviation from a least squares linear regression of \( y \) on the original \( X \).
+\end{itemize}
+
+We then pick a value of \( \nu \) between 3 and 10 to get an appropriate shape and a value of \( \lambda \) so that the \( q \)th quantile of the prior on \( \sigma \) is located at \( \hat{\sigma} \), i.e., \( P(\sigma < \hat{\sigma}) = q \).
+We consider values of \( q \) such as 0.75, 0.90, or 0.99 to center the distribution below \( \hat{\sigma} \).
+
+For automatic use, Chipman et al.
+(2010) recommend the default setting \( (\nu, q) = (3, 0.90) \).
+It is not recommended to choose \( \nu < 3 \) because it seems to concentrate too much mass on very small \( \sigma \) values, which leads to overfitting.
+
+\\\\\\\\The choice of \(m\).
+A major difference between BART and boosting methods is that, for a fixed number of trees \(m\), BART uses an iterative backfitting algorithm to cycle over and over through the \(m\) trees.
+BART involves many parameters to ensure flexibility in modeling.
+Unlike boosting and random forests, the BART algorithm iteratively updates a fixed set of \(m\) trees.
+A strong prior is used to shrink the model towards simple structures, effectively regularizing the model and encouraging it to align with additive models that incorporate some level of interaction.
+
+\textbf{the choice of m} Selecting a large \(m\) allows for better estimation of \(\mathbb{E}[Y|x]\) and improved predictive performance.
+With more trees, the model gains greater approximation flexibility, and the components \(g(x; T_1, M_1), \dots, g(x; T_m, M_m)\) become dimensionally adaptive.
+
+Conversely, choosing a smaller \(m\) highlights variable importance.
+In this setting, the fewer trees force the predictors \(x\) to compete for inclusion in the model, as seen in gradient boosting.
+The overall model fit is achieved through the cumulative effort of many “weak learners.”
+
+\textbf{Likelihood Specification for BART}
+
+BART is a sum-of-trees model, i.e., BART assumes that the label \( y \) for an input \( x \) is generated by an additive combination of \( m \) regression trees. More precisely,
+\[
+y = \sum_{j=1}^{m} g(x; T_j, M_j) + \epsilon,
+\]
+where \( \epsilon \sim \mathcal{N}(0, \sigma^2) \) is an independent Gaussian noise term with zero mean and variance \( \sigma^2 \). Hence, the likelihood for a single training instance is:
+\[
+p(y \mid T_j, M_j, \sigma^2, x) = \mathcal{N}\left(y \mid \sum_{j=1}^{m} g(x; T_j, M_j), \sigma^2\right),
+\]
+and the likelihood for the entire training dataset is:
+\[
+p(Y \mid T_j, M_j, \sigma^2, X) = \prod_{n=1}^{N} \mathcal{N}\left(y_n \mid \sum_{j=1}^{m} g(x_n; T_j, M_j), \sigma^2\right).
+\]
+
+\subsubsection{Posterior inference}
+
+Given the likelihood and the prior, our goal is to compute the posterior distribution:
+\[
+p\left( \{T_j, M_j \}_{j=1}^{m}, \sigma^2 \mid Y, X\right) \propto p\left(Y \mid \{ T_j, M_j \}_{j=1}^{m}, \sigma^2, X\right) p\left(\{T_j, M_j \}_{j=1}^{m}, \sigma^2 \mid X\right).
+\]
+Chipman et al.
+(2010) proposed a Bayesian backfitting MCMC to sample from the BART posterior.
+At a high level, the Bayesian backfitting MCMC is a Gibbs sampler that loops through the trees, sampling each tree \( T_j \) and associated parameters \( M_j \) conditioned on \( \sigma^2 \) and the remaining trees and their associated parameters \( T_{(j)}, M_{(j)}  \), and samples \( \sigma^2 \) conditioned on all the trees and parameters \( \{ T_j, M_j \}_{j=1}^{m} \).
+Let \( T_j^{(i)}, M_j^{(i)} \), and \( \sigma^{2(i)} \) respectively denote the values of \( T_j, M_j \), and \( \sigma^2 \) at the \( i \)-th MCMC iteration.
+Sampling \( \sigma^2 \) conditioned on \( \{T_j, M_j \}_{j=1}^{m} \) is straightforward due to conjugacy.
+To sample \( T_j, M_j \) conditioned on the other trees \( T_{(j)}, M_{(j)}  \), we first sample $T_j \mid T_{(j)}, M_{(j)}  , \sigma$, and then sample $M_j \mid T_j, T_{(j)}, M_{(j)},  \sigma$.
+\\More precisely, we compute the residual:
+\[
+R_j = Y - \sum_{j \neq i} g(X; T_j, M_j).
+\]
+Using the residual \( R_j^{(i)} \) as the target, Chipman et al. (2010) sample \( T_j^{(i)} \) by proposing local changes to \( T_j^{(i-1)} \).
+Finally, \( M_j \) is sampled from a Gaussian distribution conditioned on \( T_j, T_{(j)}, M_{(j)}  , \sigma^2 \).
+
+\subsection{ A Bayesian Backfitting MCMC Algorithm}
+NOTE THAT THE DRAWS ARE NOT IID
+Given the observed data \( y \), our Bayesian setup induces a posterior distribution:
+\[
+p\left(T_1, M_1, \ldots, T_m, M_m, \sigma^2 \mid y\right)
+\]
+on all the unknowns that determine a sum-of-trees model (4).
+Although the sheer size of the parameter space precludes exhaustive calculation, the following backfitting MCMC algorithm can be used to sample from this posterior.
+
+At a general level, our algorithm is a Gibbs sampler.
+For notational convenience, let \( T_{(-j)} \) be the set of all trees in the sum except \( T_j \), and similarly define \( M_{(-j)} \).
+Thus, \( T_{(-j)} \) will be a set of \( m - 1 \) trees, and the associated terminal node parameters.
+
+The Gibbs sampler here entails \( m \) successive draws of \( (T_j, M_j) \) conditionally on \( (T_{(-j)}, M_{(-j)}, \sigma^2, y) \):
+\[
+p(T_j, M_j \mid T_{(-j)}, M_{(-j)}, \sigma^2, y), \quad j = 1, \ldots, m,
+\]
+followed by a draw of \( \sigma^2 \) from the full conditional:
+\[
+p\left(\sigma^2 \mid T_1, \ldots, T_m, M_1, \ldots, M_m, y\right).
+\]
+
+Hastie and Tibshirani (2000) considered a similar application of the Gibbs sampler for posterior sampling for additive and generalized additive models with fixed bases and showed how it was a stochastic generalization of the backfitting algorithm for such models.
+For this reason, we refer to our algorithm as backfitting MCMC.
+
+The draw of \( \sigma^2 \) is simply a draw from an inverse gamma distribution and can be easily obtained by routine methods.
+More challenging is how to implement the \( m \) draws of \( (T_j, M_j) \).
+This can be done by taking advantage of the following reductions.
+First, observe that the conditional distribution:
+\[
+p\left(T_j, M_j \mid T_{(-j)}, M_{(-j)}, \sigma^2, y\right)
+\]
+depends on \( (T_{(-j)}, M_{(-j)}, y) \) only through:
+\[
+R_j = y - \sum_{k \neq j} g(x; T_k, M_k),
+\]
+the \( n \)-vector of partial residuals based on a fit that excludes the \( j \)-th tree.
+Thus, the \( m \) draws of \( (T_j, M_j) \) given \( (T_{(-j)}, M_{(-j)}, \sigma^2, y) \) are equivalent to \( m \) draws from:
+\[
+p(T_j, M_j \mid R_j, \sigma^2), \quad j = 1, \ldots, m.
+\]
+
+Now, this is formally equivalent to the posterior of the single tree model:
+\[
+R_j = g(x; T_j, M_j) + \epsilon,
+\]
+where \( R_j \) plays the role of the data \( y \). Because we have used a conjugate prior for \( M_j \):
+\[
+p(T_j \mid R_j, \sigma^2) \propto p(T_j) \int p(R_j \mid M_j, T_j, \sigma^2) p(M_j \mid T_j, \sigma^2) \, dM_j,
+\]
+this can be obtained in closed form up to a normalizing constant. This allows us to carry out each draw in two successive steps as:
+\[
+T_j \mid R_j, \sigma^2,
+\]
+and:
+\[
+M_j \mid T_j, R_j, \sigma^2.
+\]
+
+The draw of \( T_j \) can be obtained using the Metropolis-Hastings (MH) algorithm of CGM98.
+This algorithm proposes a new tree based on the current tree using one of four moves.
+The moves and their associated proposal probabilities are as follows: growing a terminal node (0.25), pruning a pair of terminal nodes (0.25), changing a non-terminal rule (0.40), and swapping a rule between parent and child (0.10).
+By integrating out \( M_j \), we avoid the complexities associated with reversible jumps between continuous spaces of varying dimensions.
+
+Finally, the draw of \( M_j \) is simply a set of independent draws of the terminal node \( \mu_j \)'s from a normal distribution.
+The draw of \( M_j \) enables the calculation of the subsequent residual \( R_{j+1} \), which is critical for the next draw of \( T_j \).
+
+We initialize the chain with \( m \) simple single-node trees and then iterate until satisfactory convergence is obtained.
+At each iteration, each tree may increase or decrease the number of terminal nodes by one or change one or two decision rules.
+Trees may change, cease to exist, or be created.
+It is not uncommon for a tree to grow large and then subsequently collapse back down to a single node as the algorithm iterates.
+
+The sum-of-trees model allows for "fit" to be freely reallocated from one tree to another.
+Because each move makes only small incremental changes to the fit, the algorithm can be imagined as analogous to sculpting a complex figure by adding and subtracting small dabs of clay.
+
+Compared to the single-tree model MCMC approach of CGM98, our backfitting MCMC algorithm mixes dramatically better.
+When only single-tree models are considered, the MCMC algorithm tends to quickly gravitate toward a single large tree and then gets stuck in a local neighborhood.
+In contrast, we have found that restarts of the backfitting MCMC algorithm give remarkably similar results even in difficult problems.
+Consequently, we run one long chain with BART rather than multiple starts.
+Although mixing does not appear to be an issue, the recently proposed modifications of Blanchard (2004) and Wu, Tjelmeland, and West (2007) might provide additional benefits.
+
+\section{Comparison}
+Various methods which combine a set of tree models, so-called ensemble methods, have attracted much attention.
+These include boosting \cite{freund1997} \cite{friedman2001}, bagging \cite{breiman1996} and random forests \cite{breiman2001}, each of which use different techniques to fit a linear combination of trees.
+
+Boosting fits a sequence of single trees, using each tree to fit data variation not explained by earlier trees in the sequence.
+Bagging and random forests use randomization to create a large number of independent trees, and then reduce prediction variance by averaging predictions across the trees.
+
+Yet another approach that results in a linear combination of trees is Bayesian model averaging applied to the posterior arising from a Bayesian single-tree model as in \cite{chipman1998} (hereafter CGM98), \cite{denison1998}, \cite{blanchard2004} and \cite{wu2007}.
+Such model averaging uses posterior probabilities as weights for averaging the predictions from individual trees.
 

### 2025-01-03 paper/paper.tex
cleaned repo

@@
 This comes from the fact that fitting a regression tree amounts to estimate a stepwise relation between $y$ and the covariates $x$ that can proxy well non-linear patterns.
-\\\\Greater flexibility of CART comes in handy also when dealing with missing values.
+Greater flexibility of CART comes in handy also when dealing with missing values.
 CART offers an effective way to overcome this issue, by constructing surrogate variables.
@@
 \\Correlation among the trees in turn decrease the effectiveness of averaging, as it does not lead to a large reduction in variance, making the entire bagging algorithm less useful. 
-\\\\ Random Forests starts from bagging and adds another kind of randomization to reduce correlation among trees.
+ Random Forests starts from bagging and adds another kind of randomization to reduce correlation among trees.
 Rather than searching over all the covariate space for each decision rule, as the greedy build of the big trees imposes, in random forests we randomly sample a subset of $m$ covariates to search over each time we make a split.
@@
 This makes the big trees explore more the space of models
-\\\\Usually, we choose $m \approx \sqrt{p}$.
+Usually, we choose $m \approx \sqrt{p}$.
 On average, $(p - m)/ = p$ of the splits will not even consider the strong predictor, and so other predictors will have more of a chance.
@@
 does not perform well. It is important to mention that in boosting, unlike in bagging, the construction of each tree strongly depends on the trees that have already grown.
-\\\\The model accounts for three different tuning parameters.
+The model accounts for three different tuning parameters.
 The number of of trees $B$, chosen trough cross-validation.
@@
 Like other ensemble methods (bagging, boosting, random forest), to solve this regression problem \eqref{regression}, BART approximates \( f(x) = \mathbb{E}(y \mid x) \) by a sum of $m$ regression trees \( h(x) \equiv \sum_{j=1}^m g(x, T_j, M_j)\), where each \( g \) is a regression tree function that given a tree $\theta_j$ to each observation $x$  assigns a leaf $\mu_{ij}$ of that regression tree.
-\\\\The BART Ensemble Model is therefore
+The BART Ensemble Model is therefore
 \[
@@
 Under \((3)\), \( \mathbb{E}(y \mid x) \) equals the sum of all the terminal node \( \mu_{ij} \)'s assigned to \( x \) by the \( g(x; T_j, M_j) \)'s.
-\\\\Like the other ensemble methods, to avoid overfitting each tree of the summation must be made a weak learner, explaining only part of the result.
+Like the other ensemble methods, to avoid overfitting each tree of the summation must be made a weak learner, explaining only part of the result.
 To do so, while boosting uses a shrinkage parameter $\lambda$, BART uses a regularization prior:
@@
 
-\\\\Overall, Bart's characteristics are in how it weakens the individual trees by instead using a prior, and how it performs the iterative fitting by instead using Bayesian backfitting on a fixed number of trees.
+Overall, Bart's characteristics are in how it weakens the individual trees by instead using a prior, and how it performs the iterative fitting by instead using Bayesian backfitting on a fixed number of trees.
 Conceptually, BART can be viewed as a Bayesian nonparametric approach that fits a parameter-rich model using a strongly influential prior distribution.
@@
 Given the way it is constructed, the tree is a full binary tree, that is, each node has exactly zero or two children.
-\\\\ Let \( g(x; T, M) \) denote the regression tree function which assigns a \( \mu_i \in M \) to \( x \).
+ Let \( g(x; T, M) \) denote the regression tree function which assigns a \( \mu_i \in M \) to \( x \).
 Each \( x \) value is associated with a single terminal node of \( T \) by the sequence of decision rules from top to bottom, and is then assigned the \( \mu_i \) value associated with this terminal node.
-\\\\We denote by \( T \) the binary tree structure itself—the various binary split decision rules of the form \( \{x_q < c\} \) versus \( \{x_q \geq c\} \), \( c \in \mathbb{R} \).
+We denote by \( T \) the binary tree structure itself—the various binary split decision rules of the form \( \{x_q < c\} \) versus \( \{x_q \geq c\} \), \( c \in \mathbb{R} \).
 The vector of parameters associated with \( T \), i.e., the collection of terminal node parameters, is denoted by \( M = \{\mu_1, \ldots, \mu_b\} \), where \( b \) is the number of terminal nodes.
 Defining a regression tree as \( g(x; T, M) \), we can view \( g(x; T, M) \) as a function that assigns the conditional mean \( \mathbb{E}[y \mid x] \) to the parameter \( \mu_i \in M \), i.e., \( \mu_i = g(x; T, M) \rightarrow \mathbb{E}[y \mid x] \) via binary decision rules denoted as \( T \).
-\\\\Thus,
+Thus,
 \begin{equation}
@@
 When the number of trees \( m > 1 \), each \( g(x; T_j, M_j) \) here is merely a part of \( \mathbb{E}(Y \mid x) \), unlike the single tree model.
-\\\\We can see that the quantity that is being ``summed'' and eventually allocated to \( \mathbb{E}(y \mid x) \) is not the regression tree or tree structure, but the value that each \( j \)th tree structure assigns to the subject.
+We can see that the quantity that is being ``summed'' and eventually allocated to \( \mathbb{E}(y \mid x) \) is not the regression tree or tree structure, but the value that each \( j \)th tree structure assigns to the subject.
 This is one way to think of a sum of regression trees.
@@
 
-\\\\A sum-of-trees model is an additive function, a sum of step functions of the individual components of \( x \).
+A sum-of-trees model is an additive function, a sum of step functions of the individual components of \( x \).
 With a large number of trees, a sum-of-trees model gains increased representation flexibility which, as we'll see, endows BART with excellent predictive capabilities.
@@
 
-\\\\\textbf{Nonlinearity of BART.} From this simple example, we can see how BART handles nonlinearity.
+\textbf{Nonlinearity of BART.} From this simple example, we can see how BART handles nonlinearity.
 Each single regression tree is a simple stepwise function or ANOVA model.
@@
 
-\\\\The BART model specification is completed by imposing a prior over all the parameters of the sum-of-trees model, namely, \( (T_1, M_1), \ldots, (T_m, M_m) \) and \( \sigma \).
+The BART model specification is completed by imposing a prior over all the parameters of the sum-of-trees model, namely, \( (T_1, M_1), \ldots, (T_m, M_m) \) and \( \sigma \).
 There exist specifications of this prior that effectively regularize the fit by keeping the individual tree effects from being unduly influential.
@@
 In addition, it keeps the contribution of each $g(x;T_i,M_i)$ model component small.
-\\\\
+
 \textbf{The $T_j$ prior}
@@
 
-\\\\\\\\Or (\(^ {17}\)), we use the conjugate normal distribution \(N(\mu, \sigma^2)\), which offers tremendous computational benefits because \(\mu\) can be marginalized out.
+Or (\(^ {17}\)), we use the conjugate normal distribution \(N(\mu, \sigma^2)\), which offers tremendous computational benefits because \(\mu\) can be marginalized out.
 
@@
 
-\\\\\\\\The choice of \(m\).
+The choice of \(m\).
 A major difference between BART and boosting methods is that, for a fixed number of trees \(m\), BART uses an iterative backfitting algorithm to cycle over and over through the \(m\) trees.

### 2025-01-03 code/tempCodeRunnerFile.r

@@ -1,1 +1,0 @@
-ytrain

### 2025-01-03 paper/paper.tex
Adding work on paper and bibliography

--- 2024-12-07 paper/paper.tex
@@
 \begin{abstract}
-Your abstract here.
+This paper provides a comprehensive comparison between Bayesian Additive Regression Trees (BART) and traditional non-Bayesian tree-based methods for regression problems.
+We begin by examining the theoretical foundations of decision trees and various ensemble methods, including bagging, random forests, and boosting.
+We then present BART as a Bayesian nonparametric approach that combines the flexibility of regression trees with the formal uncertainty quantification of Bayesian inference.
+The paper details BART's probability model, its regularization prior, and the Bayesian backfitting MCMC algorithm used for posterior inference.
+Through both simulated and real-world data applications, we demonstrate BART's effectiveness in capturing non-linear relationships and compare its predictive performance against other ensemble methods.
+Using the UCI Abalone dataset, we show that BART achieves superior prediction accuracy compared to random forests, boosting, and bagging when using sufficient numbers of trees, though at a higher computational cost.
+Our findings suggest that BART's automatic prior-based regularization and ability to quantify uncertainty make it a valuable addition to the tree-based regression toolkit, particularly for complex non-linear modeling tasks.
 \end{abstract}
 
-%\tableofcontents
+\newpage
 
@@
 
-Bayesian Regression Additive Trees (BART) were first introduced by \cite{chipmanBARTBayesianAdditive2010}
+The challenge of modeling complex, non-linear relationships in data has led to the development of various tree-based methods in statistical learning.
+While traditional parametric approaches like linear regression impose strict assumptions on the relationship between predictors and outcomes, tree-based methods offer greater flexibility in capturing non-linear patterns and interaction effects.
+This paper focuses on comparing Bayesian Additive Regression Trees (BART) with traditional non-Bayesian tree methods, examining their theoretical foundations, implementation approaches, and practical performance.
 
-\section{Non-Bayesian Tree Based Methods}
+Tree-based methods have evolved from simple decision trees to sophisticated ensemble approaches such as bagging, random forests, and boosting.
+These methods combine multiple trees to improve prediction accuracy and reduce overfitting.
+However, most traditional approaches lack formal uncertainty quantification and require careful tuning of regularization parameters.
+BART, introduced by \cite{chipmanBARTBayesianAdditive2010}, addresses these limitations by providing a Bayesian framework that automatically handles regularization through prior specifications while offering natural uncertainty estimates.
 
-Tree based method are more usefull compared to the linear regression model when the relation to be model is complex or non-linear. Indeed, the standard linear model is defined as:
+In our paper, we first provide an overview of the theoretical foundations of decision trees and their ensemble variants.
+Second, we present a comprehensive analysis of BART's methodology, including its probability model, regularization prior, and the Bayesian backfitting MCMC algorithm used for posterior inference.
+Third, we demonstrate the practical implementation and performance of these methods through both simulated data and a real-world application using the UCI Abalone dataset.
+
+The paper is organized as follows: Section \ref{decision} introduces decision trees and their fundamental concepts.
+Section \ref{frequentist} explores frequentist ensemble methods including bagging, random forests, and boosting.
+Section \ref{bart} provides a detailed examination of BART, including its model specification, prior structure, and posterior sampling approach.
+Section \ref{Comparison} offers a theoretical comparison of these methods, highlighting their relative strengths and weaknesses.
+Finally, Section \ref{data_app} presents empirical results from both simulated and real-world data applications, demonstrating the practical implications of choosing between these methods for regression tasks.\footnote{The code repository for this paper is available at this \href{https://github.com/giuliofrey/bart}{link}}
+
+\section{Decision trees}
+\label{decision}
+
+Consider a target variable \( Y \) and a \( p \)-dimensional vector of covariates, \( x = (x_1, \ldots, x_p) \).
+Decision trees are predictive models that make predictions about \( Y \) by partitioning the predictor space—the set of all possible values of \( x \)—into simple, non-overlapping regions.
+Notice that also linear
+regression in a sense partitions the predictor space, but restricts divisions to certain planes.
+To the contrary, all trees requires is that the partition can be achieved by successive binary splits based on the different predictors.
+Once we have a partition such as this we base our prediction on the average of the $Y$ in each partition.
+
+The structure of a decision tree reflects this partitioning process.
+The tree begins with a root node at the top, representing the entire dataset (or the full covariate space).
+The root node is split into two subsets according to decision rules based on the predictor variables \( (x_1, \ldots, x_p) \).
+Each subsequent subset may also be further split into two new subsets, continuing the process recursively.
+At each split, a new internal node is created, and the segments connecting nodes are called branches.
+This recursive splitting process, known as \textit{recursive partitioning}, proceeds iteratively until a termination condition is met, such as when further splitting no longer improves predictions or the subsets become too small.
+The final nodes, called \textit{terminal nodes} or \textit{leaves}, correspond to the smallest subsets created by the tree.
+
+The terminal nodes define a partition of the predictor space into \( J \) non-overlapping regions, \( R_1, R_2, \ldots, R_J \).
+Each region corresponds to a leaf of the tree.
+To make a prediction for a given covariate vector \( x \), the tree is traversed starting at the root node.
+At each internal node, the splitting rule determines whether to proceed to the left or right child node.
+This process continues until a terminal node is reached.
+The predicted value \( \hat{y} \) is then assigned based on the region \( x \) belongs to.
+
+When \( Y \) is continuous, the tree is called a \textit{regression tree}, and the predicted value \( \hat{y}_i \) for a new observation \( x_i \) falling into region \( R_j \) is the mean of \( Y \) values for the training observations in \( R_j \).
+For categorical \( Y \), the tree is referred to as a \textit{classification tree}, and the predicted value is the majority class of \( Y \) among the training observations in \( R_j \).
+Collectively, these approaches are known as \textit{CART} (Classification and Regression Trees), a term introduced by \cite{Gordon_Breiman_Friedman_Olshen_Stone_1984}.
+
+In this work, we focus specifically on regression trees, which aim to predict a continuous \( Y \) given a training dataset.
+Regression trees differ fundamentally from linear regression models because they do not impose any parametric form on the relationship between \( x \) and \( Y \).
+Instead, regression trees use a piecewise constant approximation of this relationship.
+
+To see this, suppose there are \( n \) observations of \( Y \) and \(\mathbf{x} = (x_1, \ldots, x_p) \). The relationship between the covariates and the response variable can be expressed as:
 \begin{equation}
-  f(X) = \beta_0 + \sum_{j=1}^p X_j \beta_j
+Y = f(X) + \epsilon \label{regression}
 \end{equation}
-Wheras the decision tree model as:
-\begin{equation}
-  f(X) = \sum_{m=1}^M c_m \cdot 1(X \in R_m)
-\end{equation}
-DESCRIBE BETTER THIS EQUATIONS
+where \( \epsilon \) is an error, whose distribution is not necessarily specified.
+Both regression trees and linear regression models seek to estimate the unknown function \( f(X) \) using the observed data.
 
-We take most of our arguments for this section from the \cite[Chapter~8]{jamesIntroductionStatisticalLearning2021}.
+The manner in which \( f(X) \) is approximated, however, depends on the model chosen. Linear regression assumes a parametric form:
+\[
+f(X) = \beta_0 + \sum_{j=1}^p X_j \beta_j,
+\]
+where \( \beta_0, \beta_1, \ldots, \beta_p \) are coefficients estimated from the data. In contrast, regression trees approximate \( f(X) \) as a stepwise constant function:
+\[
+f(X) = \sum_{j=1}^J \mu_j \cdot 1(X \in R_j),
+\]
+where \( R_1, \ldots, R_J \) are the regions defined by the leaves of the tree, and \( \mu_j \) is the mean of \( Y \) for the training observations in region \( R_j \).
+
+The choice between these models depends on the nature of the underlying relationship between the predictor variables and the response variable.
+If the relationship is approximately linear, a linear regression model is likely to outperform a regression tree.
+Conversely, if the relationship is highly nonlinear or complex, regression trees can capture such patterns more effectively.
+Additionally, regression trees have the advantage of being invariant to the scale of the predictor variables, eliminating the need for preprocessing steps like standardization.
+
+This section draws from \cite[Chapter~8]{jamesIntroductionStatisticalLearning2021} and \cite{Genuer_Poggi_2020} to illustrate these concepts.
 
 \subsection{Regression Trees}
+A regression tree partitions the \(X\)-space into disjoint regions \(R_j\) and provides a fitted value \(E(Y \mid X \in R_j)\) within each region.
+We consider the following notation.
+Denote by \( T \) the tree structure, consisting of a set of interior node decision rules and a set of $b$ terminal nodes.
+The decision rules are binary splits of the predictor space of the form \( \{x \in A\} \) vs \( \{x \notin A\} \), where \( A \) is a subset of the range of \( x \).
+We only consider rules based on the single components of \( x = (x_1, \dots, x_p) \) and of the form \( \{x_i \leq c\} \) vs \( \{x_i > c\} \) for continuous \( x_i \).
+Let \( M = \{\mu_1, \mu_2, \dots, \mu_b\} \) denote a set of parameter values associated with each of the \( b \) terminal nodes of \( T \).
+Given the way it is constructed, the tree is a full binary tree, that is, each node has exactly zero or two children.
 
-Tree based methods that we cover in this paper are all based on decision trees.
-Decision trees can be applied both to regression or classification problems, and are thus named accordingly to their application.
-Regression trees in particular are build in two steps.
+We now delve better in the process of building regression trees.
+As we saw before, they are built in two steps.
+First, the predictor space (the set of all possible values of the vector of observables $X_1, ...., X_p$) is split into $J$ regions $R_j$ that are distinct and non-overlapping.
+Secondly, given a new observation $x$ that falls in a region $R_j$, the model predicts the corresponding outcome variable $\hat{y}$ to be the mean of the outcome variables corresponding to the observations in the training set that fall in that same region.
 
-First, the prediction space (the set of all possible values of the vector of observables) is split into $J$ regions $R_j$ that are distinct and non-overlapping.
-Secondly, a prediction for the outcome variable is made for each observation in each $R_j$ regions, as the mean of the training observations in that same regions.
-
-For a predictor space of $P$ elements the regions are $p$ dimensional boxes, that are build by minimizing the Residual Sum Squared, as the equation \ref{rss_tree} shows.
+To begin, we have to construct the regions $R_1,..., R_j$.
+Typically, regions are chosen as $p$-dimensional boxes, where $p$ is the number of regressors.
+The partition is chosen is such a way to minimize the RSS:
 \begin{equation}
-  R_1,...,R_j=\operatorname*{argmin}_{R_j}\sum^J_{j=1}\sum_{i\in R_j}\left(y_i-\hat{y}_{r_j}\right)^2
+  \{R_1,...,R_J \}=\operatorname*{argmin}_{ \{R_j\}_{j=1}^J}\sum^J_{j=1}\sum_{i\in R_j}\left(y_i-\mu_{R_j}\right)^2
   \label{rss_tree}
 \end{equation}
+where $\mu_{R_j}$ is the mean of the training observation in the $j$-th box.
 
-Instead of considering all the possible partitions of the predictor space, a top-down approach is used.
-First, all observations are considered such that they belong to the same region.
-Then, this process chooses the best split minimizing the RSS irrespective of future outcomes.
-This is then repeated for a number of times, until a stopping criterion is reached (usually a minimum number of observations in all the regions).
+It is computationally infeasible to do the optimization problem in \eqref{rss_tree}.
+Instead of considering all the possible partitions of the predictor space, a top-down, greedy approach, known as recursive binary splitting is used.
+The approach is \texttt{top-down} since it begins at the top of the tree, and then sequentially splits the predictor space; each split is then indicated by two new branches (hence it is binary).
+Moreover, the process is \textit{greedy}, since for each step we make the split that generates the greatest reduction of the RSS at that particular step, and we do not choose the best sequence of steps looking ahead.
+\\To perform recursive binary splitting, the algorithm considers the set of all predictors $X_1, \ldots, X_p$, and the set of all possible values of the cutpoint $s$, and then choose the predictor $X_j$ and the cutpoint $s$ such that the branching of the tree corresponding to the decision rule:  $$\{X \mid X_j < s\} \quad \text{and} \quad \{X \mid X_j \geq s\}$$ minimizes the RSS.
+Then, we repeat the process within the resulting subregion, looking again for the predictor and the cutpoint such that the subsequent decision rule minimize the RSS.
+It is important to note that this time we do not split the entire predictor space but one of the two previously identified regions.
+This process continues until a stopping condition is reached; the more common requires a minimum number of observation within each splitting nodes.
 
-As the regions visualizations are complex when more than 2 regressors are present, a tree based graph is used.
+\subsection{Fitting trees: growing and pruning}
 
-The process described above will likely tend to overfit the data.
-There are numerous solutions to this, the most simple one being increasing the number of AND GENERALLY THE ENSAMBLING MODELSobservations that rise the stopping rule.
-A valid alternative is to grow a very large tree, and then reduce it to obtain a smaller subtree.
-This is known as tree pruning.
+Recursive binary splitting to build a regression tree  is likely to produce good prediction on the training-set but overfit the training data, leading to poor out-of-sample performance.The reason is that the above mentioned procedure is likely to give us a complex tree, which has probably a low bias but high variance in prediction.
+Indeed, small fluctuations in the training set may lead to different predictions.
 
-ADD COST COMPLEXITY PRUNING
+Constraining ex ante the size of trees would be a short-sighted solution, because one could possibly stop too soon, leaving potentially good splits further down in tree unachieved.
+\\A better strategy is pruning an already complex tree.
+This amounts to search for a good intermediate between two extreme cases: a fully developed tree which has high variance but low bias, and a tree consisting of only the root, with high bias but low variance.
+The strategy consists in growing a very large tree $T_0$, and then prune it back in order to obtain a subtree. To do so, we use cost complexity pruning, also know as weakest link pruning.
 
-One of the main disadvantages of this setting, is that trees are non-robust.
-In fact a small change in the data can lead to a large change in the estimation.
-This follows from the fact that Regression Trees suffer from high variance.
-This is indeed quite problematic and the main reason that simple regression trees are not used by practitioners.
-Ensemble methods aim at solving this issue.
-Those methods in the context of trees combine simpler models (called weak learners), whichy may lead to mediocre predictions, in a single more powerful and stable model.
-QUelli di cui sotto sopra abuiao detto sono tutti ensable models.
+Consider a sequence of subtrees of $T_0$, $\{T\}_\alpha$ indexed by a non-negative tuning parameter $\alpha$.
+For each value of $\alpha$, there is a subtree $T \subset T_0$ such that
+$$\sum_{m=1}^{|T|} \sum_{i : x_i \in R_m} \left( y_i - \mu_{R_m} \right)^2 + \alpha |T|$$
+is as small as possible, where $|T|$ represents the number of terminal nodes of the tree T, $R_m$ is the subset of the predictor space corresponding to the $m$-th terminal node, $\mu_{R_m}$ is the predicted response associated with $R_m$.
+The tuning parameter $\alpha$ controls the trade-off between a sub-tree complexity and and its fit to the training data, the higher the $\alpha$, the more terminal nodes a tree has, the smaller a subtree is.
+The value for $\alpha$ can be chosen using a validation set or trough cross-validation.
+
+\subsection{Pros and Cons of decision trees}
+Overall, regression trees offer much greater flexibility in handling the relation between covariates and output than linear models do.
+This comes from the fact that fitting a regression tree amounts to estimate a stepwise relation between $y$ and the covariates $x$ that can proxy well non-linear patterns.
+They can more naturally incorpo
+rate interaction effects.
+
+Greater flexibility comes in handy also when dealing with missing values.
+Tree offer an effective way to overcome this issue, by constructing surrogate variables.
+Exploiting correlation between predictors, they may decrease the effect of missing data.
+
+So far we considered only binary splits, thus an interesting question would be if a model with more than two splits at each note is a valid idea.
+However, it turns out that in general is not a good strategy.
+The issue is that multiple splits fragment the dataset too quickly.
+Moreover, since multiple splits can be achieved by a series of binary splits, the latter approach is preferred.
+
+The interpretability of trees has been one of the driving forces of its success.
+Indeed, thanks to their structure, is straightforward to understand why, for a given $x$, we expect a particular value of $y$, following a series of decision rules.
+Small trees are visually easy to interpret.
+Moreover, trees build in such a way are a computationally fast method, that easily scales with large datasets.
+
+However, one of the main disadvantages of this setting, is that trees are non-robust.
+They suffer from high variance: a small change in the data can lead to a large change in the estimation.
+The main reason of this instability is the hierarchical structure of the process as the effect of an error in one of the first nodes is propagated to the other nodes below it.
+This is heightened by the fact that single trees tend to overfit, and spurious correlations in the training data can be mistaken as structural patterns and included in the decision rules.
+This may lead to bad out-of-sample predictive performance.
+High variance is the main reasons that simple regression trees are generally not used.
+Instead, ensemble methods are preferred.
+
+\section{Frequentist ensemble methods}
+\label{frequentist}
+
+We highlighted that individual decision trees tend to overfit and to be sensitive to the training data they were built on. To counteract this, ensemble methods were proposed where multiple models
+are aggregated or averaged over to give a more stable and
+generalisable solution.
+
+In general, ensemble methods combine the fit from many (hundreds, thousands) simpler models to get an overall predictor. 
+All the models we are going to discuss, namely Bagging, Boosting, Random Forest and Bayesian Additive Regression Trees are ensemble methods, for which the simpler models consist of regression trees.
 
 \subsection{Bagging}
-Bagging is a generic method that aims at reducing the variance of a statistical learning method.
-Bagging is based on the fact that if we have independent observations $(Z_1,...,Z_n)$ with same variance $\sigma$, then the vairance of the sample mean of the observation will be lower, as it is $V(\bar{Z})=\frac{\sigma}{n}$.
+Bagging, which stands for Bootstrap Aggregation, is a general technique that aims at reducing the variance of a statistical learning method.
+It is baseed on the fact that, given independent observation, averaging over them reduces variance.
 
-Bagging uses a bootstrap sampling procedure to obtain different samples from the data, and then fits a tree for each of the bootstrapped samples.
-The single estimates for each sample are then averaged, as equation \ref{eq_bagging} shows.
+For the case of regression trees, the main problem of using a single tree as a prediction method is that perturbing the learning set can cause significant changes in the predictor constructed. 
+We would thus like to be able to take many different training sets from the population, to build a different trees for each training set and then to average the resulting predictions.
+However this is unfeasible since we do not have access to many training sets.
+
+Hence, we resort to bootstrap.
+Bootstrapping consists of taking repeated samples, with replacement, from the single available training set.
+In such a way, we are able to construct $B$ different bootstrapped training sets.
+Bagging involves creating multiple $B$ copies of the original training set using hte bootstrap, fitting a separate decision tree to each copy.
+Then it combines the results from each of the $B$ trees to get an overall prediction.
+The single predictions for the target function $f(X)$ for each sample are averaged, as
 \begin{equation}
@@
 \end{equation}
-Where $B$ is the number of samples of the bootstrapping procedure.
+where $B$ is the number of samples of the bootstrapping procedure.
+
+In bagging the trees are grown big and are not pruned.
+Therefore, each single tree has high variance and low bias.
+Averaging these different trees reduce the variance.
+Bagging can dramatically reduce the variance of unstable procedures such
+as trees, leading to improved prediction.
+
+The drawback of bagging is that we lose interpretability to improve accuracy. When we bag a large number of trees, we can no longer represent the model with a single tree.
+
+\subsection{Random Forest}
+A possible problem with bagging is due to the lack of randomization between trees.
+Indeed, since trees are separately fitted on the bootstrapped training set, it may happen that they share to great amount the same decision rules.
+This is the case, for example, if there is very strong predictor in the dataset.
+If this happens, in the collection of bagged trees a vast majority of the trees will use this strong predictor in the top split and they will all be hihgly correlated among themselves.
+
+Correlation among the trees decrease the effectiveness of averaging, as it does not lead to a large reduction in variance, making the entire bagging algorithm less useful.
+
+ Random Forests starts from bagging and adds another kind of randomization to reduce correlation among trees.
+Rather than searching over all the covariate space for each decision rule, as the greedy building of the big trees imposes, in random forests each time we make a split in a tree we randomly sample a subset of $m$ covariates to search over.
+Therefore, the split is allowed to consider only one of those $m$ predictors.
+Moreover, at each split is considered a new sample of $m$ predictors.
+This makes the big trees explore more the space of models.
+
+ Usually, we choose $m \approx \sqrt{p}$.
+On average, $(p - m)/ = p$ of the splits will not even consider the strong predictor, and so other predictors will have more of a chance.
+Therefore, we can think of this process as decorrelating the trees, making the average of the resulting trees less variable and hence more reliable.
 
@@
 
-Similar to bagging, boosting is not specific to regression trees, it is rather a generic approach.
-Said so, it is quite effective when used on trees.
-In fact ,it approaches the overfitting problem but without using bootstrapping, as bagging does.
-Boosting relies on fitting trees sequentially, as algorithm \ref{alg_boosting} shows.
-
-The number of sequential trees to fit, as chosen by the parameter $B$ in algorithm  \ref{alg_boosting}, if chosen too large, can cause overfitting.
-The learning rate $\lambda$ is another tunable parameter.
-Often small values are chosen ($\lambda=10^{-2}$ or even smaller).
-
-\begin{algorithm}
+Differently from bagging and random forests, boosting does not involve fitting independent trees from many bootstrap samples and averaging the predictions to reduce variance.
+Instead, it relies on fitting trees sequentially, that is, using information from previously grown trees.
+Indeed, trees are not fit to bootstrapped versions of the original datasets, but rather to the residual not explained by the remainder of
+the trees.
+Then, its prediction is added, with a weight, to the prediction of the previous trees.
+This procedure is iterarated until we express the overall prediction of the boosted trees as trhe weighted average of the prediction of the single trees.
+This is shown in algorithm \ref{alg_boosting}. 
+\begin{algorithm} [h]
   \caption{Boosting for Regression Trees}
@@
 
+By fitting sequentially new trees on the residual of the previous tree, boosting improves at each iteration the fit marginally.
+The parameter $d$, which determines the number of terminal nodes, andh the parameter $\lambda$, which weights the contirbution of each tree to the overall prediction, constraint the single trees to be weak learners.
+By  fitting small trees to the residuals, we slowly improve $\hat{f}$ in areas where it
+does not perform well. It is important to mention that in boosting, unlike in bagging, the construction of each tree strongly depends on the trees that have already grown.
+
+The model accounts for three different tuning parameters.
+The number of of trees $B$, chosen trough cross-validation.
+Interestingly, unlike bagging and random forests, boosting can suffer of overfitting if $B$ is too large, since tree are strongly dependent from each other.
+The shrinkage parameter $\lambda$, a small positive number.
+Common values for $\lambda$ are 0.01 or 0.001, depending on the problem at hand.
+However, very small values of $\lambda$ often require a very large B to achieve good performances.
+Lastly, $d$ represents the number of splits in each tree.
+Often $d=1$ works well.
+However in this case the tree is called stump since it contains only a single split.
+
+\section{Bayesian additive regression trees}
+\label{bart}
+
+The idea for Bayesian Additive Regression Trees (BART) was first developed by Chipman, George, McCulloch (2010 AOAS).
+Like other ensemble methods (bagging, boosting, random forest), to solve this regression problem \eqref{regression}, BART approximates \( f(x) = \mathbb{E}(y \mid x) \) by a sum of regression trees.
+However, being a Bayesian method, BART requires specifying a full probability model for the data, composed by a likelihood and a prior. 
+The likelihood is obtained by assuming normality of the error in the regression \eqref{regression}. In other words, BART assumes 
+\begin{equation}
+Y = f(X) + \epsilon, \quad \epsilon \sim \mathcal{N}(0, \sigma^2), \label{regression}
+\end{equation}
+where \( \sigma \) is a parameter.
+
+The function $f$ is modeled with a sum of $m$ regression trees, which are the remaining parameters of the model.
+As before, we separate the parameter space into the tree structure $T$, which consists of all the decision rules, and $M$, which is the set of leaves or terminal nodes.
+Leaves are going to be denoted by $\mu$, as they are simply means in a  regression setting.
+
+Thus, in BART the relationship between $(y, X)$ is modeled as: 
+\begin{equation}
+y = \sum_{j=1}^m g(x; T_j, M_j) + \epsilon, \quad \epsilon \sim N(0, \sigma^2) \label{firstlikelihood}
+\end{equation}
+where for each binary regression tree \( T_j \) and its associated terminal nodes  \( M_j \), \( g(x; T_j, M_j) \) is the function assigning \( \mu_{ij} \in M_j \)  to \( x \), and \( \mu_{ij} \in M_j \) will be the predicted value of $y$ for $x$.
+
+ The BART model is completed by a prior on the parameters $ (\{T_j, M_j \}_{j=1}^m, \sigma)$, 
+\[
+p((T_1, M_1), (T_2, M_2), \ldots, (T_m, M_m), \sigma).
+\]
+The essential idea of BART is to exploit this prior to keep the individual tree effects in the fit small and avoid overfitting.
+Indeed, while the number $m$ of trees is fixed, the overall number of parameters (that is the depth of each trees and the number of terminal nodes) is left unspecified.
+This should make BART prone to overfit the data by growing large and complex trees.
+To overcome this problem, BART uses a regularization prior, that is a prior set in a way to shrink the model towards simplicity.
+As an effect of shrinkage, each tree is forced to explain only a limited subset of the relationships between the covariates and the predictor variable.
+This is similar to what boosting does, with the difference that in boosting, being a frequentist method, shrinkage is regulated by a parameter $\lambda$, which needs tuning.
+
+Since the space of parameters $ (\{T_j, M_j \}_{j=1}^m, \sigma)$ is high and random dimensional (remember that only the number of trees is specified, but not the numbers of decision rules or of leaves), BART uses Markov Chain Monte Carlo (MCMC) to get draws from posterior distribution.
+At every iteration of the MCMC, the software generates a representation of $f$ and a value for $\sigma$ given the data.
+A single posterior mean estimate of \( f(x) = \mathbb{E}(Y \mid x) \) at any input value $x$ is obtained by a simple average of these successive draws evaluated at \( x \).
+Further, pointwise uncertainty intervals for \( f(x) \) are easily obtained from the corresponding quantiles of the sample of draws.
+
+Overall, BART can be viewed as a Bayesian nonparametric approach (each tree has unknown size) that fits a parameter-rich model like \eqref{firstlikelihood} using a regularization prior to avoid overfitting.
+We are going to examine the a sum-of-trees model and implied likelihood first, and then look at the regularization prior.
+
+\subsection{A Sum-of-Trees Model}
+As before, and for technical reasons that will be clear in the posterior sampling, BART models each tree separating it into $T$ and $M$.
+We denote by \( T \) the binary tree structure —the various binary split decision rules of the form \( \{x_q < c\} \) versus \( \{x_q \geq c\} \), with \( c \in \mathbb{R} \) and $x_q$ an element of the vector of covariates $x$.
+The vector of parameters associated with \( T \), i.e., the collection of terminal node parameters, is denoted by \( M = \{\mu_1, \ldots, \mu_b\} \), where \( b \) is the number of terminal nodes.
+The $(T, M)$ are to be thought about as a parameter $\theta$ in prior specification and posterior sampling.
+
+We define a regression tree as a function \( g(x; T, M) \) that assigns the conditional mean \( \mathbb{E}[y \mid x] \) to the parameter \( \mu_i \in M \), i.e., \( \mu_i = g(x; T, M) \rightarrow \mathbb{E}[y \mid x] \) via binary decision rules denoted as \( T \).
+With this notation, the sum-of-trees model can be more explicitly expressed as
+\begin{equation}
+Y = \sum_{j=1}^m g(x; T_j, M_j) + \epsilon, \quad \epsilon \sim N(0, \sigma^2) \label{likelihood}
+\end{equation}
+where for each binary regression tree \( T_j \) and its associated terminal node parameters \( M_j \), \( g(x; T_j, M_j) \) is the function which assigns \( \mu_{j} \in M_j \) to \( x \).
+Under \eqref{likelihood}, \( \mathbb{E}(Y \mid x) \) equals the sum of all the terminal node \( \mu_{j} \)'s assigned by the $m$ \( g(x \mid T_j, M_j) \)'s.
+
+ As all others sum-of-tree models, the quantity that is being summed and eventually allocated to \( \mathbb{E}(y \mid x) \) is not the regression tree or tree structure, but the value that each \( j \)th tree structure assigns to the subject, that is \( \mu_{j} \).
+This makes single regression tree is a simple stepwise function, and summing trees amounts to summing together these stepwise functions.
+As a result, the resulting function $f(x)$  can approximate the nonlinearities and  interactions effects between the covariates $x$.
+
+Given \eqref{likelihood},  the likelihood for a single training instance is:
+\[
+p(y \mid T_j, M_j, \sigma^2, x) = \mathcal{N}\left(y \mid \sum_{j=1}^{m} g(x; T_j, M_j), \sigma^2\right),
+\]
+and the likelihood for the entire training dataset (note that $Y = (y_1, \ldots, y_n)$) is:
+\[
+p(Y \mid T_j, M_j, \sigma^2, X) = \prod_{i=1}^{N} \mathcal{N}\left(y_i \mid \sum_{j=1}^{m} g(x_i; T_j, M_j), \sigma^2\right).
+\]
+
+Consider the parameter space $(\{T_j, M_j\}_{j=1}^m, \sigma)$ of model \eqref{likelihood}.
+The only identified parameter is $\sigma$, because it's the only one that can be estimated as the variance of the residulas.
+For the trees, only their number $m$ is fixed, while the individual trees remain unidentified and interchangeable.
+Indeed, the dimension of each tree (namely the number of internal and terminal nodes) is left unbounded.
+If data demands it, the trees can grow complex, with many leaves and nodes, but they can also remain simple.
+Moreover, trees can always be swapped one with the other, so they are fundamentally interchangeable.
+As a result, the only interpretable parameter is $\sigma$-
+
+For this reason, BART can be defined as a nonparametric Bayes approach: since the dimension of $T$ and $M$ is left unbounded, the model does not have a fixed number of parameters.
+The dimension of each tree is going to be inferred from the data.
+This high-dimensional and random structure underpins BART's ability to model complex data flexibly and effectively.This interchangeability means the trees themselves lack direct interpretability.
+Despite the large number of parameters, BART avoids overfitting through the use of an appropriate regularization prior.
+
+\subsection{A regularization prior}
+
+Since each tree \( g(x; T_j, M_j) \) in \eqref{likelihood}, including the variables to split on, the splitting values, and the mean parameters at each terminal node, is unknown, a prior distribution 
+\[
+p((T_1, M_1), (T_2, M_2), \ldots, (T_m, M_m), \sigma)
+\]
+over all the parameters of the sum-of-trees model is needed.
+
+Since the parameter space is high and random dimensional, to avoid overfitting the prior must be chosen in order to keep the contribution of each \(g(x; T_j, M_j)\) small.
+The prior is thus a regularization prior, that shrinks the model towards simplicity.
+But there is also a connection to boosting: indeed, the prior ensures that each tree in the ensemble acts as a weak learner.
+Rather than attempting to fit the overall pattern in the data, each tree focuses on capturing the residual pattern that other trees did not capture.
+
+\subsubsection{Prior independence and Symmetry}
+In order to simplify the specification of the regularization prior, Chipman assumes that  the trees \( \{(T_1, M_1), \ldots, (T_m, M_m)\} \) are independent of each other and of  \( \sigma \).
+Then, the prior distribution can be written, using the same prior for each $(T, M)$ pair, as:
+
+\[
+p[(T_1, M_1), \ldots, (T_m, M_m), \sigma] = \left[ \prod_{j=1}^m p(T_j, M_j) \right] \, p(\sigma).
+\]
+For each  $(T, M)$ pair, the joint prior is represented as a marginal for $T$ and a conditional for $M$ given $T$.
+The choice of conditioning on $T_j$ is due the fact that the dimension of $M$ is determined by the number of bottom nodes in the tree, so once $T$ is known, $M$ is just a vector.
+\[
+= \left[ \prod_{j=1}^m p(M_j \mid T_j) \, p(T_j) \right] \, p(\sigma).
+\]
+Given $T_j$, let  \( b_j \)  be the number of terminal nodes in the tree \( T_j \).
+Then \( M_j = \{\mu_1, \mu_2, \ldots, \mu_{b_j}\} \) is the vector of terminal node parameters associated with \( T_j \).
+Chipman at al.
+assume that node parameters \( \mu_{ij} \in M_j\) are iid, that is $\mu_{ij} \overset{\mathrm{iid}}{\sim} p(\mu_j)$ This implies:
+\[
+p(M_j \mid T_j) = \prod_{i=1}^{b_j} p(\mu_{ij} \mid T_j)
+\]
+Thus we can write the prior distribution for \eqref{likelihood} as:
+\[
+p[(T_1, M_1), \ldots, (T_m, M_m), \sigma]= \left[ \prod_{j=1}^m \left( \prod_{i=1}^{b_j} p(\mu_{ji} \mid T_j) \right) p(T_j) \right] \, p(\sigma).
+\]
+Under these assumptions, for the BART prior is sufficient to specify \( p(T_j) \), \( p(\mu_{ij} \mid T_j) \), and \( p(\sigma) \) (note that each $p$ here denote a different probability distribution depending on its input).
+The idea of regularization prior is to choose them so to have each $T$ small, each $\mu$ small,  and an appropriate $\sigma$  (e.g.
+smaller than the least squares estimate).
+
+\subsubsection{The $T_j$ prior}
+ Instead of a probability distribution, Chipman specifies a process that can be used to draw a tree from the prior. The \( T_j \) prior, \( p(T_j) \), is specified by three aspects:
+\begin{enumerate}
+    \item The probability that a node at depth \( d = (0, 1, 2, \ldots) \) is nonterminal is given by:
+    \[
+    \frac{\alpha}{(1 + d)^\beta}  \quad \text{with} \quad \alpha \in (0, 1) \quad \text{and} \quad \beta \in [0, \infty),
+    \]
+    where node depth $d$ is defined as the distance from the root.
+In default \texttt{wbart}, $\alpha$ is set 0.95 and $\beta$ is set equal to 2.
+As the tree grows and the depth increases, the probability of splitting decreases.
+This shrinkage mechanism prevents the tree from growing indefinitely.
+Nonetheless, trees with many terminal nodes can be grown if the data demands it.
+    \item The distribution used to select the covariate to split upon in an internal node.This is the uniform prior on available variables in vector $x$.
+    \item The distribution used to select the cutoff point in an internal node conditional on the covariate selected.
+In \texttt{wbart}, this is the uniform prior on the discrete set of available splitting values $c$.
+\end{enumerate}
+
+\subsubsection{The \( \mu_{i,j} | T_j \) Prior}
+$\mu_{ij} | T_j$ are assumed to be iid. The prior specification for the leaves is $$ \mu_{ji} |T_j \overset{\text{iid}}{\sim}N(\mu_\mu, \sigma_\mu^2)$$
+To specify the hyperparameters $\mu_\mu, \sigma_\mu$, \cite{chipmanBARTBayesianAdditive2010} turn to empirical Bayes, that is they  calibrate the prior using the observed variation in \( y \).
+In particular, note that under equation \eqref{likelihood}, each $y_i$ is the sum of an error $\varepsilon_i$ and $m$ $\mu_{ij}$.
+Thus, given $\mu_{ij} \mid T_j \sim N(\mu_\mu, \sigma^2_\mu)$, the induced prior on $E(Y \mid x)$ is \(N(m\mu_\mu, \sigma^2_\mu)\).
+Since we expect that \(\mathbb{E}(Y \mid x)\) lies between \(y_{\text{min}}\) and \(y_{\text{max}}\), we choose \(\mu_\mu\) and \(\sigma_\mu\) such that the induced prior distribution of $E(Y \mid x)$ assigns substantial probability to the interval \((y_{\text{min}}, y_{\text{max}})\).
+
+To choose $\mu_\mu$, \( y \) is centered around zero by subtracting its sample meand and it is rescaled so that the observed transformed values range from \( y_{\text{min}} = -0.5 \) to \( y_{\text{max}} = 0.5 \).
+Then the prior we can choose for $\mu_{ij}$ is zero.
+On the other hand, $\sigma_\mu$ is picked such that
+\[
+\sigma_\mu k\sqrt{m} = 0.5.
+\]
+for some suitable value of $k$ (generally $k=2$).
+
+As \(k\) and/or \(m\) (the number of trees) increase, such a prior shrinks the tree leaves \(\mu_{ij}\) toward zero, limiting the influence of individual tree components and thus making it a weak learner.
+
+\subsubsection{The \( \sigma \) Prior}
+For the prior of the variance of the idiosyncratic error $\varepsilon$ in \eqref{likelihood}, the inverse chi-square distribution is used:
+\[
+\sigma^2 \sim \frac{\nu \lambda}{\chi^2_\nu} 
+\]
+We use the the inverse chi-square because it is a conjugate prior for the variance of the normal distribution, and the distribution of the errors $\varepsilon$ in \eqref{likelihood} is assumed to be normal in our probability model: this will make the draw of $\sigma$ from the posterior easy.
+
+The hyperparameters we need to choose are the degrees of freedom $\nu$ and the scale parameter $\lambda$. Again, an empirical Bayesian approach is adopted to assign 
+substantial probability to the entire region of plausible values of $\sigma$.
+The estimate $\hat{\sigma}$ produced by the software \texttt{wbart} we use in our simulation can be either the sample standard deviation of \( y \) (if the number of covariates is more of the numebr of observations), or the residual standard deviation from a least squares linear regression of \( y \) on the original \( X \).
+The value of \( \nu \) is then picked between 3 and 10 to get an appropriate shape of the distribution of $\sigma$, while the value of \( \lambda \) is chosen so that \( p(\sigma < \hat{\sigma}) = q \), where $q = 0.75, 0.95, 0.99$.
+
+\subsection{ Posterior sampling}
+Given the observed data \( y \), our Bayesian setup induces a posterior distribution:
+\[
+\begin{aligned}
+p\left( \{T_j, M_j\}_{j=1}^m, \sigma^2 \mid Y, X \right) &\propto p\left( Y \mid \{T_j, M_j\}_{j=1}^m, \sigma^2, X \right) \times p\left( \{T_j, M_j\}_{j=1}^m, \sigma \right)
+\end{aligned}
+\]
+Since the parameter space is highly dimensional, exact calculation of the posterior is infeasible.
+However, a MCMC algorithm can be used to sample from this posterior.
+The algorithm allows to obtain a dependent sample from the posterior
+ distribution over the space of sum-of-trees models.
+
+This algorithm is a Gibbs sampler, in which one parameter is drawn from the full conditional, and then you cycle through the others.
+In particular, the draw of all parameters from the posterior can be simplified into two major draws from full conditionals: first $m$ successive draws of the trees, and then a final draw for $\sigma$.
+
+For notational convenience, let \( T_{(-j)} \) be the set of all trees except \( T_j \), and similarly define \( M_{(-j)} \).
+Thus, \( T_{(-j)} \) will be a set of \( m - 1 \) trees, and the associated terminal node parameters.
+The Gibbs sampler loops through the trees, sampling:
+\begin{enumerate}
+    \item  first,  \( m \) successive draws of \( (T_j, M_j) \) conditionally on \( (T_{(-j)}, M_{(-j)}, \sigma^2, y) \):
+\begin{equation}
+p(T_j, M_j \mid T_{(-j)}, M_{(-j)}, \sigma^2, y), \quad j = 1, \ldots, m, \label{priortree}
+\end{equation}
+\item then,  draw of \( \sigma^2 \) from the full conditional:
+\begin{equation}
+p\left(\sigma^2 \mid\{T_j, M_j\}_{j=1}^m, y\right). \label{priorsigma}
+\end{equation}
+\end{enumerate}
+The draw of $\sigma$ from \eqref{priorsigma} is made easy by the choice of a conjugate prior.
+Indeed, if all the tree models $(T_i, M_i)$ are known, then the function $f$ is known, and we can compute $y$ minus$ f(x)$ to get the residuals of \eqref{firstlikelihood}.
+Since hte residulas are assumed to be distributed normally, and the inverse chi-squared distribution is (conditionally) the conjugate prior for the variance of the normal distribution, the draw of \( \sigma^2 \) from the posterior is simply a draw from an inverse chi-squared distribution with updated parameters.
+To get the \( m \) draws of the tree models \( (T_j, M_j) \) from \eqref{priortree}, we take advantage of the following reductions. First, observe that the full conditional:
+\[
+p\left(T_j, M_j \mid T_{(-j)}, M_{(-j)}, \sigma^2, y\right)
+\]
+depends on \( (T_{(-j)}, M_{(-j)}, y) \) only through:
+\begin{equation}
+R_j = y - \sum_{k \neq j} g(x; T_k, M_k), \label{residual}
+\end{equation}
+which for each $j$ is the $n$-vector of residuals of the $m-1$ regression sum-of-trees excluding the \( j \)-th tree.
+Thus, the \( m \) draws of \( (T_j, M_j) \) given \( (T_{(-j)}, M_{(-j)}, \sigma^2, y) \) are equivalent to \( m \) draws from:
+\begin{equation}
+p(T_j, M_j \mid R_j, \sigma^2), \quad j = 1, \ldots, m. \label{13}
+\end{equation}
+
+Now we use the fact that, by the definition of conditional probability, we can write:
+\[
+p(T_j, M_j \mid R_j, \sigma^2) = p(M_j \mid T_j, R_j, \sigma^2) \; p(T_j \mid R_j, \sigma^2).
+\]
+We start from \( p(T_j \mid R_j, \sigma^2) \). Using Bayes' Theorem and standard probability laws, 
+\[
+p(T_j \mid R_j, \sigma^2) \propto p(T_j) \; p(R_j \mid T_j, \sigma^2) 
+\propto p(T_j) \; \int p(R_j, M_j \mid T_j, \sigma^2) \, dM_j.
+\]
+
+Expanding further:
+\[
+p(T_j \mid R_j, \sigma^2) \propto p(T_j) \; \int p(R_j \mid T_j, M_j, \sigma^2) \; p(M_j \mid T_j, \sigma^2) \, dM_j
+\]
+\[
+= p(T_j) \; \int p(R_j \mid T_j, M_j, \sigma^2) \; p(M_j \mid T_j, \sigma^2) \; p(\sigma^2) \, dM_j.
+\]
+
+And finally:
+\begin{equation}
+p(T_j \mid R_j, \sigma^2) \propto p(T_j) \int p(R_j \mid M_j, T_j, \sigma^2) \; p(M_j \mid T_j, \sigma^2) \, dM_j. \label{posterioeT}
+\end{equation}
+
+Notice that $R_j$, the $n$-vector of partial residuals, is conditionally normally distributed,and that we have used a conjugate normal prior for $M_j$.
+Thus, it can easily be integrated it out from \eqref{posterioeT} to obtain $p(T_j \mid R_j, \sigma^2)$ in closed form up to a normalizing constant.
+Thus we can just draw $T$ alone from \eqref{posterioeT} and then draw $M_j$ conditionally on it.
+In other words, the draw from \eqref{13} can be split in two successive steps.
+\begin{enumerate}
+    \item  first,  we draw  $T_j$ conditional on all the other trees/ on the residula $R_j$ and on $\sigma$ from \eqref{posterioeT}:
+\begin{equation}
+ T_j \mid R_j, \sigma^2, \label{MH}
+\end{equation}
+\item then, you draw $M_j$conditional on all the trees and on $\sigma$:
+\begin{equation}
+M_j \mid T_j, R_j, \sigma^2 \label{drawM}
+\end{equation}
+\end{enumerate}
+The draw of $T_j$ from \eqref{MH} is obtained using a Metropolis-Hastings (MH) algorithm.
+Let \( T_j^{(i)} \), \( M_j^{(i)} \), and \( \sigma^{(i)} \) denote the values of \( T_j \), \( M_j \), and \( \sigma \) at the \( i \)th MCMC iteration, respectively.
+Each draw  \( T_j^{(i)} \mid R_j^{(i)}, \sigma^{(i)} \) from \eqref{MH} is obtained using MH.
+At the MCMC iteration $i$, MH samples $T_j^{(i)}$ by proposing local changes to $T_j^{(i-1)}$.
+CHanges are either accepted or rejected in such a way that it is still true that the posterior is the stationary distribution.
+The proposal changes to $T_J^{(i-1)}$ that MH can carry are the following: 
+\begin{itemize}
+    \item \textbf{Grow:} randomly chooses a leaf node and splits it further into left and right children (with probability 0.25). 
+    \item \textbf{Prune:} randomly chooses an internal node where both the children are leaf nodes and prunes the two leaf nodes, thereby making the internal node a leaf node (probability 0.25). 
+    \item \textbf{Change:} changes the decision rule at a randomly chosen internal node (probbility 0.40)
+    \item \textbf{Swap:} swaps the decision rules at a parent-child pair where both the parent and child are internal nodes (probability 0.10)
+    \end{itemize}
+    Clearly, these moves change the dimension of the tree and thus of the parameter space.
+The fact that the dimension of the parameter space changes as inference is carried out is indeed a feature of Bayesian nonparametrics.
+
+Finally, once we have $p(T_j \mid R_j, \sigma^2)$, we are left with the draw $M_j$ from the distribution \eqref{drawM}. Given that the $\mu_{ji}$ are iid, we have: 
+$$ p(M_j \mid T_j, R_j, \sigma^2) = \prod_{i=1}^{b_j} p(\mu_{ij} \mid T_j, R_j, \sigma^2),
+$$ where the prior on $\mu_{ji}$ is  $N(0, \sigma_\mu^2)$.
+Notice that, if you call  \( R_{ji} \) the subsets of residuals in the $n$-vector  $R_j$ that are allocated to the terminal node parameter  \( \mu_{ji} \), and call \( n_i \) their number, then \( R_{ji} \mid T_j, \mu_{ji}, \sigma \sim \mathcal{N}(\mu_{ji}, \sigma^2) \).
+Thus the prior of $\mu_{ji}$ is actually conjugate, and the posterior we are drawing from, $p(\mu_{ji} \mid T_j, \sigma, R_j) \propto p(R_{ji} \mid T_j, \mu_{ji}, \sigma) p(\mu_{ji} \mid T_j)$, is going to be a normal distribution.
+The draw of \( M_j \) from this normal enables the calculation of the subsequent residual \( R_{j+1} \), which allows the successive draw of \( T_j \).
+
+The MCMC algorithm is initialized with $m$ simple single node trees, and then iterations are
+ repeated until convergence.
+At each iteration, the dimension of the trees changes.
+Contrarily to boosting, where trees are fitted sequentially on the residual of the previous, in this algorithm you have a set of trees, and then you cycle through all the trees and keep adjusting one tree, given all the others.
+For this reason, it is also called Bayesian backfitting.
+The algorithm is summarized in the following figure.
+\begin{algorithm} [h]
+  \caption{Bayesian backfitting MCMC for posterior inference in BART}
+  \label{MCMC}
+  \SetAlgoLined
+  \DontPrintSemicolon
+  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
+
+  \textbf{Inputs:} Training data $(X, Y)$, BART hyperparameters $(\alpha, \beta, m, k, \nu, q)$.\;
+
+  \For{$j = 1, \ldots, m$}{ 
+      (a) Initialize $T_j^{(0)}$ with a single leaf node.\;
+      (b) Sample $M_j^{(0)}$. \tcp*{\textit{$\mu_j = \frac{\overline{Y}}{m}$}}
+  }
+
+  \For{$i = 1, \ldots, \texttt{maxiter}$}{     
+      (c) Sample $\sigma^{(i)} \mid \{T_j^{(i-1)}, M_j^{(i-1)}\}_{j=1}^m$ \tcp*{\textit{Sample from invgamma }}
+
+      \For{$j = 1, \ldots, m$}{
+          (d) Compute residual $R_j^{(i)}$ \tcp*{\textit{Using \eqref{residual}}}
+
+          (e) Sample $T_j^{(i)} \mid R_j^{(i)}, \sigma^{(i)}, T_j^{(i-1)}$ \tcp*{\textit{Using MH}}
+
+          (f) Sample $M_j^{(i)} \mid R_j^{(i)}, \sigma^{(i)}, T_j^{(i)}$ \tcp*{\textit{Sample from Gaussian distribution}}
+      }
+  }
+\end{algorithm}
+\textit{Adapted from the blog}.
+
+\section{Comparison}
+\label{Comparison}
+
+We examined different ensemble methods that combine regression trees to estimate an unknown function $f(X)$ in a regression like 
+\[
+Y = f(X) + \varepsilon
+\]
+without assuming linearity.
+These ensembles of regression trees allows to model nonlinearity and interaction effects between covariates by fitting some sort of stepwise function on the data.
+According to Lakshminarayanan (2015), ensembles of regression trees can be
+broadly classified into two families: randomized independent regression trees, wherein the trees are grown independently and predictions are averaged to reduce variance, and additive regression trees, wherein each
+tree fits the residual not explained by the remainder of
+the trees.
+In the first category there are bagged trees, and random forests, which use use randomization to create a large number of independent trees, and then reduce prediction variance by averaging predictions across the tree \parencite{chipmanBARTBayesianAdditive2010}.
+
+Addititive regression trees methods can be further categorized into those that are fit in a serial fashion, like in boosting, and those fit in an iterative fashion, like
+Bayesian additive regression trees (BART).
+In moth methods, trees are grown dependently one from the other and thus are prone to overfitting and must be regularized.
+For example, boosting fits a sequence of single trees, using the resedual of the previous tree to fit the subsequent tree.
+Regularization in boosting is governed by parameters $\lambda$ and $d$, which shrink the contribution of each tree toward zero.
+In this way, each tree is forced to explain only a part of the overall fit (it is a weak learner).
+
+On the other hand, BART is a Bayesian method defined by a probability model with a likelihood and a prior.
+Regularization is performed by the prior distribution, which is set in such a way to prefer simple tree structures $T_j$ and non-extreme predictions at leaves $\mu_{ji}$.
+The prior thus constrain each tree to be a weak learner, just like in boosting.
+Since the dimension of the trees (the number of nodes) is left unbounded,  In effect, BART is a nonparametric Bayesian method: the trees are dimensionally adaptive \parencite{chipmanBARTBayesianAdditive2010}.
+
+To fit the sum-of-trees model, BART uses a Bayesian backfitting MCMC algorithm that iteratively constructs and fits successive residuals.
+Here there is another difference with boosting: while in boosting trees are built sequentially, with each tree explicitly addressing the residual errors of the previous ensemble of trees, in BART a fixed number $m$ of trees is set and trees are adjusted collectively during Markov Chain Monte Carlo (MCMC) iterations to fit the data.
+In other words, at each MCMC iterations the dimension of the trees is updated.
+
+Overall, BART's strength comes from the fact that being a Bayesian method its shrinkage parameters does not need tuning, ad the prior will perform shrinkage automatically.
+In terms of performance, the literature (Chipman et a.) shows that BART has predictive performance comparable to other ensemble methods and far superior to linear regression when nonlinearities are present in the data.
+
 \section{Data Application}
+\label{data_app}
 
 We now move to applications of the theoretical insights derived in the previous sections.
-Firstly, we will focus on simulated data, and then we will cover an application using real word data on abalones.
+Firstly, we will focus on simulated data, and then we will cover an application using real word data on Abalones.
 
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{../outputs/sin_plot.pdf}
+  \includegraphics[width=0.7\textwidth]{outputs/sin_plot.pdf}
   \caption{Plot of the estimates (blue) against observations (black) and generating function(green)}
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{../outputs/sigma_plot.pdf}
+  \includegraphics[width=0.7\textwidth]{outputs/sigma_plot.pdf}
   \caption{Value of $\sigma$ over MCMC iterations}
@@
   \centering
-  \includegraphics[width=0.7\textwidth]{../outputs/sin_plot_diff_sigma.pdf}
+  \includegraphics[width=0.7\textwidth]{outputs/sin_plot_diff_sigma.pdf}
   \caption{DIfferent prior mean of $\sigma$}
@@
 We run BART from \cite{mccullochBARTBayesianAdditive2024}; Boosting from \cite{ridgewayGbmGeneralizedBoosted2024}; Random Forest and Bagging from \cite{breimanRandomForestBreimanCutlers2024}.Our aim is to compare the predictive performance of those methods.
-In fact,  we use half of the sample to fit $\hat{y}$, that is the age variable, and half the sample to give us out of bag Mean Squared Errors (MSE) for predicting the same variable.
+In order to do so, we use half of the sample to fit $\hat{y}$, that is the age variable, and half the sample to give us out of bag Mean Squared Errors (MSE) for predicting the same variable.
 Models are run for various values of T, and for each value, both the mean squared error (MSE) of out-of-sample predictions (shown in plot \ref{plot_mse}) and the computation time (depicted in plot \ref{plot_time}) are recorded.
@@
 Although it is the best performer, it is also the model that takes the most to be run, and time of computations scales linearly with the number of trees that are fitted.
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.7\textwidth]{outputs/mse_plot.pdf}
+  \caption{Mean Squared Error for different values of T}
+  \label{plot_mse}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics[width=0.7\textwidth]{outputs/time_plot.pdf}
+  \caption{Time of computations for different values of T}
+  \label{plot_time}
+\end{figure}
 
@@
 
-\begin{figure}
-  \centering
-  \includegraphics[width=0.7\textwidth]{../outputs/mse_plot.pdf}
-  \caption{Mean Squared Error for different values of T}
-  \label{plot_mse}
-\end{figure}
-
-\begin{figure}
-  \centering
-  \includegraphics[width=0.7\textwidth]{../outputs/time_plot.pdf}
-  \caption{Time of computations for different values of T}
-  \label{plot_time}
-\end{figure}
+\newpage
+$ $
+\newpage
 

### 2025-01-03 code/plots.r
Adding plot to readme

@@ -30,4 +30,5 @@ time_plot <-ggplot(data, aes(x = ntree, y = time, color = method)) +
 
 #save the plots
 ggsave("outputs/mse_plot.pdf", mse_plot, width = 6, height = 4)
+ggsave("outputs/mse_plot.png", mse_plot, width = 6, height = 4)
 ggsave("outputs/time_plot.pdf", time_plot, width = 6, height = 4)

### 2025-01-03 code/plots.r
Adding plot to readme

@@ -30,5 +30,5 @@ time_plot <-ggplot(data, aes(x = ntree, y = time, color = method)) +
 
 #save the plots
 ggsave("outputs/mse_plot.pdf", mse_plot, width = 6, height = 4)
-ggsave("outputs/mse_plot.png", mse_plot, width = 6, height = 4)
+ggsave("outputs/mse_plot.png", mse_plot, width = 6, height = 4, bg = "white")
 ggsave("outputs/time_plot.pdf", time_plot, width = 6, height = 4)

### 2025-01-10 code/ code.r
final work on code

@@ -18,7 +18,9 @@ if (any(installed_packages == FALSE)) {
 # Packages loading
 invisible(lapply(packages, library, character.only = TRUE, warn.conflicts = FALSE))
 
-# Settting the seed
+# Settting the Warning message:
+Continuous x aesthetic
+ℹ did you forget `aes(group = ...)`? seed
 set.seed(123)
 
 
@@ -38,16 +40,23 @@ bart_sin <- gbart(xtrain, ytrain, x.test = xtest)
 tree_sin <- tree(ytrain ~ xtrain)
 print(summary(tree_sin))
 
+plot_data <- data.frame(xtest, ytest, yhat_mean = bart_sin$yhat.test.mean)
+plot_data$true_function <- 2 * sin(xtest) + xtest
+plot_data$lower_quantile <- apply(bart_sin$yhat.test, 2, quantile, probs = 0.025)
+plot_data$upper_quantile <- apply(bart_sin$yhat.test, 2, quantile, probs = 0.975)
 
-# plot the results using ggplot2
-plot_sin <- ggplot(data = data.frame(xtest, ytest), aes(x = xtest, y = ytest)) +
-  geom_point() +  
-  geom_line(aes(x = xtest, y = bart_sin$yhat.test.mean), color = "blue") +
-  geom_line(aes(x = xtest, y = 2*sin(xtest) + xtest), color = "green") +
+plot_sin <- ggplot(plot_data, aes(x = xtest, y = ytest)) +
+  geom_point(aes(color = "Observed Data"), color = "black") +
+  geom_line(aes(y = yhat_mean, color = "BART Prediction")) +
+  geom_line(aes(y = true_function, color = "True Function"), linetype = "dashed") +
+  geom_ribbon(aes(ymin = lower_quantile, ymax = upper_quantile, fill = "95% CI"), alpha = 0.2) +
   labs(title = "BART vs. True Function",
        x = "x",
-       y = "2sin(x)+x") +
-  theme_minimal()
+       y = "2sin(x) + x",
+       color = "Legend",
+       fill = "Legend") +
+  theme_minimal() +
+  theme(legend.position = "bottom")
 
 ggsave("outputs/sin_plot.pdf", plot_sin, width = 6, height = 4)
 
@@ -55,6 +64,7 @@ ggsave("outputs/sin_plot.pdf", plot_sin, width = 6, height = 4)
 
[…]

### 2025-01-10 code/boxplot.r

@@ -1,0 +1,58 @@
+# Package names
+packages <- c(
+    "ggplot2", 
+    "dplyr",
+    "tree",
+    "xtable",
+    "tidyr",
+    "BART",
+    "gbm", # For Generalized Boosted Regression Models
+    "randomForest", # For Random Forest,
+    "ucimlrepo" # For the Abalone age prediction dataset,
+)
+# Install packages not yet installed
+installed_packages <- packages %in% rownames(installed.packages())
+if (any(installed_packages == FALSE)) {
+  install.packages(packages[!installed_packages])
+}
+
+# Packages loading
+invisible(lapply(packages, library, character.only = TRUE, warn.conflicts = FALSE))
+
+# Settting the seed
+set.seed(123)
+
+#load the crab dataset
+abalone <- fetch_ucirepo(name = "Abalone")
+
+abalone_df <- abalone$data$original
+
+
+#convert the categorical variables to factors
+abalone_df$Sex <- as.factor(abalone_df$Sex)
+
+#convert rings to age
+abalone_df$Age <- abalone_df$Rings + 1.5
+#remove the rings column
+abalone_df <- abalone_df[, setdiff(names(abalone_df), "Rings")]
+
+
+train <- sample(1:nrow(abalone_df), nrow(abalone_df) / 2)
+x <- abalone_df[,  setdiff(names(abalone_df), "Age")]
+y <- abalone_df[, "Age"]
+xtrain <- x[train, ]
+ytrain <- y[train]
+xtest <- x[-train, ]
+ytest <- y[-train]
+
+
+bart <- gbart(xtrain, ytrain, x.test = xtest, ntree = 500, mc.cores = 4)
+bart.mse <- mean((bart$yhat.test.mean - ytest)^2)
+
+ii <- order(bart$yhat.train.mean)
+
+# Select a subset of the training data to make the plot less crowded
+subset_indices <- seq(1, length(ii), length.out = 100)
+pdf("outputs/bart_boxplot.pdf")
+boxplot(bart$yhat.train[, ii[subset_indices]], ylim = range(ytrain), xlab = "Ordered Indices", ylab = "Predicted Values", main = "BART Model Training Predictions")
+dev.off()

### 2025-01-10 paper/paper.tex
updated draft

@@
 
+\usepackage{listings}
 \usepackage[utf8]{inputenc}
@@
 
-\title{BART: A Comparison with Non-bayesian Tree Methods}
+\title{BART: A Comparison with Frequentist Tree Methods}
 \author{
@@
 \begin{abstract}
-This paper provides a comprehensive comparison between Bayesian Additive Regression Trees (BART) and traditional non-Bayesian tree-based methods for regression problems.
-We begin by examining the theoretical foundations of decision trees and various ensemble methods, including bagging, random forests, and boosting.
-We then present BART as a Bayesian nonparametric approach that combines the flexibility of regression trees with the formal uncertainty quantification of Bayesian inference.
-The paper details BART's probability model, its regularization prior, and the Bayesian backfitting MCMC algorithm used for posterior inference.
-Through both simulated and real-world data applications, we demonstrate BART's effectiveness in capturing non-linear relationships and compare its predictive performance against other ensemble methods.
-Using the UCI Abalone dataset, we show that BART achieves superior prediction accuracy compared to random forests, boosting, and bagging when using sufficient numbers of trees, though at a higher computational cost.
-Our findings suggest that BART's automatic prior-based regularization and ability to quantify uncertainty make it a valuable addition to the tree-based regression toolkit, particularly for complex non-linear modeling tasks.
+This paper provides a comparison between Bayesian Additive Regression Trees (BART) and traditional non-Bayesian tree-based methods for regression problems.
+We begin by examining the theoretical foundations of decision trees.
+We then turn to frequentist ensemble methods, including bagging, random forests, and boosting.
+Finally, we introduce BART, a Bayesian nonparametric approach that combines the flexibility of regression trees with the formal uncertainty quantification of Bayesian inference.
+The paper details BART's likelihood, regularization prior, and the Bayesian backfitting MCMC algorithm used for posterior inference.
+Our final section is a data application, in which we compare the predictive performance of BART against other ensemble methods through both simulated and real-world data.
+In particular, using the UCI Abalone dataset, we show that BART achieves superior prediction accuracy compared to random forests, boosting, and bagging when using sufficient numbers of trees, though at a higher computational cost.
+Overall, our findings suggest that BART's automatic prior-based regularization and ability to quantify uncertainty is particularly well suited for complex non-linear modeling tasks.
 \end{abstract}
@@
 While traditional parametric approaches like linear regression impose strict assumptions on the relationship between predictors and outcomes, tree-based methods offer greater flexibility in capturing non-linear patterns and interaction effects.
-This paper focuses on comparing Bayesian Additive Regression Trees (BART) with traditional non-Bayesian tree methods, examining their theoretical foundations, implementation approaches, and practical performance.
+This paper focuses on comparing Bayesian Additive Regression Trees (BART) with traditional non-Bayesian tree methods, examining both their theoretical foundations and practical performance.
 
-Tree-based methods have evolved from simple decision trees to sophisticated ensemble approaches such as bagging, random forests, and boosting.
+Tree-based methods have evolved from simple decision trees to ensemble approaches such as bagging, random forests, and boosting.
 These methods combine multiple trees to improve prediction accuracy and reduce overfitting.
-However, most traditional approaches lack formal uncertainty quantification and require careful tuning of regularization parameters.
-BART, introduced by \cite{chipmanBARTBayesianAdditive2010}, addresses these limitations by providing a Bayesian framework that automatically handles regularization through prior specifications while offering natural uncertainty estimates.
+However, most traditional approaches lack formal uncertainty quantification and require careful tuning of regularization parameters, for example boosting.
+BART, introduced by \cite{chipmanBARTBayesianAdditive2010}, addresses these limitations by providing a Bayesian framework that automatically handles regularization through prior specifications while offering uncertainty estimates.
 
@@
 Section \ref{bart} provides a detailed examination of BART, including its model specification, prior structure, and posterior sampling approach.
-Section \ref{Comparison} offers a theoretical comparison of these methods, highlighting their relative strengths and weaknesses.
-Finally, Section \ref{data_app} presents empirical results from both simulated and real-world data applications, demonstrating the practical implications of choosing between these methods for regression tasks.\footnote{The code repository for this paper is available at this \href{https://github.com/giuliofrey/bart}{link}}
+Section \ref{data_app} presents empirical results from both simulated and real-world data applications.
+Finally, Section \ref{conclusion} concludes comparing all the tree ensemble methods both on the theoretical and empirical ground.
+\footnote{The code repository for this paper is available at this \href{https://github.com/giuliofrey/bart}{link}}
 
@@
 
-Consider a target variable \( Y \) and a \( p \)-dimensional vector of covariates, \( x = (x_1, \ldots, x_p) \).
-Decision trees are predictive models that make predictions about \( Y \) by partitioning the predictor space—the set of all possible values of \( x \)—into simple, non-overlapping regions.
+Consider a target variable \( y \) and a \( p \)-dimensional vector of covariates, \( \mathbf{x} = (x_1, \ldots, x_p) \).
+Decision trees are predictive models that make predictions about \( y \) by partitioning the predictor space—the set of all possible values of the random vector \( \mathbf{x} \)—into simple, non-overlapping regions.
 Notice that also linear
 regression in a sense partitions the predictor space, but restricts divisions to certain planes.
-To the contrary, all trees requires is that the partition can be achieved by successive binary splits based on the different predictors.
-Once we have a partition such as this we base our prediction on the average of the $Y$ in each partition.
+To the contrary, trees only require that the partition can be achieved by successive splits of the predictor space based on the different predictors.
 
-The structure of a decision tree reflects this partitioning process.
-The tree begins with a root node at the top, representing the entire dataset (or the full covariate space).
-The root node is split into two subsets according to decision rules based on the predictor variables \( (x_1, \ldots, x_p) \).
-Each subsequent subset may also be further split into two new subsets, continuing the process recursively.
-At each split, a new internal node is created, and the segments connecting nodes are called branches.
-This recursive splitting process, known as \textit{recursive partitioning}, proceeds iteratively until a termination condition is met, such as when further splitting no longer improves predictions or the subsets become too small.
+The structure of a decision tree reflects this partitioning process.We consider only binary trees, where each split of the predictor space can create only two regions.
+This is in general a good strategy, as multiple splits can be achieved by a series of binary splits.
+Each tree begins with a root node at the top, representing the entire dataset (the full covariate space).
+The initial split divides the covariate space into two non-overlapping subsets, according to a decision rule based on the predictor variables \( (x_1, \ldots, x_p) \).
+Each of these two subsets may be in turn further split into two new (sub-)subsets, and so on.
+Each split is represented by a new node in the tree.
+Nodes that represents non-finals splits are called \textit{internal nodes}, while the segments of the tree connecting these nodes are called \textit{branches}.
+This splitting process, known as \textit{recursive partitioning}, proceeds iteratively until a termination condition is met, such as when further splitting no longer improves predictions or the subsets become too small.
 The final nodes, called \textit{terminal nodes} or \textit{leaves}, correspond to the smallest subsets created by the tree.
@@
 Each region corresponds to a leaf of the tree.
-To make a prediction for a given covariate vector \( x \), the tree is traversed starting at the root node.
+To make a prediction for a given covariate vector \( \mathbf{x}_i \), the tree is traversed starting at the root node.
 At each internal node, the splitting rule determines whether to proceed to the left or right child node.
 This process continues until a terminal node is reached.
-The predicted value \( \hat{y} \) is then assigned based on the region \( x \) belongs to.
+The predicted value \( \hat{y}_i \) is then assigned based on the final region \( \mathbf{x}_i \) falls into.
 
-When \( Y \) is continuous, the tree is called a \textit{regression tree}, and the predicted value \( \hat{y}_i \) for a new observation \( x_i \) falling into region \( R_j \) is the mean of \( Y \) values for the training observations in \( R_j \).
-For categorical \( Y \), the tree is referred to as a \textit{classification tree}, and the predicted value is the majority class of \( Y \) among the training observations in \( R_j \).
+When \( y \) is continuous, the tree is called a \textit{regression tree}, and the predicted value \( \hat{y}_i \) for a new observation \( \mathbf{x}_i \) falling into region \( R_j \) is the mean of \( y \) values for the training observations in \( R_j \).
+For categorical \( y \), the tree is referred to as a \textit{classification tree}, and the predicted value is the majority class of \( y \) among the training observations in \( R_j \).
 Collectively, these approaches are known as \textit{CART} (Classification and Regression Trees), a term introduced by \cite{Gordon_Breiman_Friedman_Olshen_Stone_1984}.
 
-In this work, we focus specifically on regression trees, which aim to predict a continuous \( Y \) given a training dataset.
-Regression trees differ fundamentally from linear regression models because they do not impose any parametric form on the relationship between \( x \) and \( Y \).
-Instead, regression trees use a piecewise constant approximation of this relationship.
+In this work, we focus specifically on regression trees, which aim at predicting a continuous variable \( Y \) given a training dataset.
+Regression trees differ fundamentally from linear regression models because they do not impose any parametric form on the relationship between \( \mathbf{x} \) and \( Y \).
+Instead, regression trees use a piecewise-constant approximation of this relationship.
 
-To see this, suppose there are \( n \) observations of \( Y \) and \(\mathbf{x} = (x_1, \ldots, x_p) \). The relationship between the covariates and the response variable can be expressed as:
+To see this, suppose that the training data is composed by $n$ iid samples
+\(\mathbf{X} = \{\mathbf{x}_i\}_{i=1}^n\), where \(\mathbf{x} \in \mathbb{R}^p\), 
+along with corresponding response variable
+\(Y = \{y_i\}_{i=1}^n\), where \(y_i \in \mathbb{R}\). The relationship between the covariates and the response variable can be expressed as:
 \begin{equation}
-Y = f(X) + \epsilon \label{regression}
+y = f(\mathbf{x}) + \epsilon \label{regression}
 \end{equation}
 where \( \epsilon \) is an error, whose distribution is not necessarily specified.
-Both regression trees and linear regression models seek to estimate the unknown function \( f(X) \) using the observed data.
+Both regression trees and linear regression models seek to estimate the unknown function \( f(\mathbf{x}) \) using the observed data.
 
-The manner in which \( f(X) \) is approximated, however, depends on the model chosen. Linear regression assumes a parametric form:
+The manner in which \( f(\mathbf{x}) \) is approximated depends on the model chosen. Linear regression assumes a parametric form:
 \[
-f(X) = \beta_0 + \sum_{j=1}^p X_j \beta_j,
+f(\mathbf{x}) = \beta_0 + \sum_{j=1}^p \beta_jx_j ,
 \]
-where \( \beta_0, \beta_1, \ldots, \beta_p \) are coefficients estimated from the data. In contrast, regression trees approximate \( f(X) \) as a stepwise constant function:
+where \( \beta_0, \beta_1, \ldots, \beta_p \) are coefficients estimated from the data. In contrast, regression trees approximate \( f(\mathbf{x}) \) as a stepwise constant function:
 \[
-f(X) = \sum_{j=1}^J \mu_j \cdot 1(X \in R_j),
+f(\mathbf{x}) = \sum_{j=1}^J \mu_j \cdot 1(\mathbf{x}\in R_j),
 \]
-where \( R_1, \ldots, R_J \) are the regions defined by the leaves of the tree, and \( \mu_j \) is the mean of \( Y \) for the training observations in region \( R_j \).
+where \( R_1, \ldots, R_J \) are the regions defined by the leaves of the tree, and \( \mu_j \) is the mean of \( y \) for the training observations in region \( R_j \).
 
@@
 If the relationship is approximately linear, a linear regression model is likely to outperform a regression tree.
-Conversely, if the relationship is highly nonlinear or complex, regression trees can capture such patterns more effectively.
-Additionally, regression trees have the advantage of being invariant to the scale of the predictor variables, eliminating the need for preprocessing steps like standardization.
-
-This section draws from \cite[Chapter~8]{jamesIntroductionStatisticalLearning2021} and \cite{Genuer_Poggi_2020} to illustrate these concepts.
+Conversely, if the relationship is highly nonlinear, regression trees can capture such patterns more effectively.
 
 \subsection{Regression Trees}
-A regression tree partitions the \(X\)-space into disjoint regions \(R_j\) and provides a fitted value \(E(Y \mid X \in R_j)\) within each region.
-We consider the following notation.
-Denote by \( T \) the tree structure, consisting of a set of interior node decision rules and a set of $b$ terminal nodes.
-The decision rules are binary splits of the predictor space of the form \( \{x \in A\} \) vs \( \{x \notin A\} \), where \( A \) is a subset of the range of \( x \).
-We only consider rules based on the single components of \( x = (x_1, \dots, x_p) \) and of the form \( \{x_i \leq c\} \) vs \( \{x_i > c\} \) for continuous \( x_i \).
-Let \( M = \{\mu_1, \mu_2, \dots, \mu_b\} \) denote a set of parameter values associated with each of the \( b \) terminal nodes of \( T \).
-Given the way it is constructed, the tree is a full binary tree, that is, each node has exactly zero or two children.
 
-We now delve better in the process of building regression trees.
-As we saw before, they are built in two steps.
-First, the predictor space (the set of all possible values of the vector of observables $X_1, ...., X_p$) is split into $J$ regions $R_j$ that are distinct and non-overlapping.
-Secondly, given a new observation $x$ that falls in a region $R_j$, the model predicts the corresponding outcome variable $\hat{y}$ to be the mean of the outcome variables corresponding to the observations in the training set that fall in that same region.
+This section draws from \cite[Chapter~8]{jamesIntroductionStatisticalLearning2021} and \cite{Genuer_Poggi_2020}.
 
-To begin, we have to construct the regions $R_1,..., R_j$.
+A regression tree splits the predictor space into $J$ non-overlapping regions $R_j$.
+Given a new observation $\mathbf{x}$ that falls in a region $R_j$, the model predicts the corresponding outcome variable $\hat{y}$ to be the mean of the outcome variables corresponding to the observations in the training set that fall in that same region.
+ We denote by \( T \) the (binary) tree structure, consisting of a set of interior node decision rules and a set of $b$ terminal nodes.
+The decision rules are binary splits of the predictor space of the form \( \{\mathbf{x} \in A\} \) vs \( \{\mathbf{x} \notin A\} \), where \( A \) is a subset of the predictor space.
+Following  \cite{chipmanBARTBayesianAdditive2010}, we only consider rules based on the single components of \( \mathbf{x} = (x_1, \dots, x_p) \) and of the form \( \{x_i \leq c\} \) vs \( \{x_i > c\} \) for continuous \( x_i \).
+Let \( M = \{\mu_1, \mu_2, \dots, \mu_b\} \) denote the set of parameter values associated with each of the \( b \) terminal nodes of \( T \).
+
+The first step to build a regression tree is to construct the partition $R_1,..., R_j$.
 Typically, regions are chosen as $p$-dimensional boxes, where $p$ is the number of regressors.
-The partition is chosen is such a way to minimize the RSS:
+The partition is chosen is such a way to minimize the sum of squared errors:
 \begin{equation}
@@
 It is computationally infeasible to do the optimization problem in \eqref{rss_tree}.
-Instead of considering all the possible partitions of the predictor space, a top-down, greedy approach, known as recursive binary splitting is used.
-The approach is \texttt{top-down} since it begins at the top of the tree, and then sequentially splits the predictor space; each split is then indicated by two new branches (hence it is binary).
-Moreover, the process is \textit{greedy}, since for each step we make the split that generates the greatest reduction of the RSS at that particular step, and we do not choose the best sequence of steps looking ahead.
-\\To perform recursive binary splitting, the algorithm considers the set of all predictors $X_1, \ldots, X_p$, and the set of all possible values of the cutpoint $s$, and then choose the predictor $X_j$ and the cutpoint $s$ such that the branching of the tree corresponding to the decision rule:  $$\{X \mid X_j < s\} \quad \text{and} \quad \{X \mid X_j \geq s\}$$ minimizes the RSS.
-Then, we repeat the process within the resulting subregion, looking again for the predictor and the cutpoint such that the subsequent decision rule minimize the RSS.
-It is important to note that this time we do not split the entire predictor space but one of the two previously identified regions.
-This process continues until a stopping condition is reached; the more common requires a minimum number of observation within each splitting nodes.
+Instead of considering all the possible partitions of the predictor space, a top-down, greedy algorithm, known as recursive binary splitting, is generally used.
+This approach is \textit{top-down} because it begins at the top of the tree, and then sequentially splits the predictor space.
+It is also \textit{greedy}, since for each step we choose the split that generates the greatest immediate reduction of the RSS: we do not choose the best sequence of steps looking ahead.
+To perform the splitting, the algorithm considers the set of all predictors $ \mathbf{x}= (x_1, \ldots, x_p)$, and the set of all possible values of the cutpoint $s$, and then choose the predictor $x_j$ and the cutpoint $s$ such that the splitting generated by the decision rule  $$\{\mathbf{x} \mid x_j < s\} \quad \text{and} \quad \{\mathbf{x} \mid x_j \geq s\}$$ minimizes the RSS.
+The process of looking for the predictor and the cutpoint such that the implied decision rule minimize the RSS continues within each subregion of the space until a stopping condition is reached; the more common requires a minimum number of observation within each splitting nodes.
 
-\subsection{Fitting trees: growing and pruning}
-
-Recursive binary splitting to build a regression tree  is likely to produce good prediction on the training-set but overfit the training data, leading to poor out-of-sample performance.The reason is that the above mentioned procedure is likely to give us a complex tree, which has probably a low bias but high variance in prediction.
-Indeed, small fluctuations in the training set may lead to different predictions.
-
-Constraining ex ante the size of trees would be a short-sighted solution, because one could possibly stop too soon, leaving potentially good splits further down in tree unachieved.
-\\A better strategy is pruning an already complex tree.
-This amounts to search for a good intermediate between two extreme cases: a fully developed tree which has high variance but low bias, and a tree consisting of only the root, with high bias but low variance.
-The strategy consists in growing a very large tree $T_0$, and then prune it back in order to obtain a subtree. To do so, we use cost complexity pruning, also know as weakest link pruning.
-
-Consider a sequence of subtrees of $T_0$, $\{T\}_\alpha$ indexed by a non-negative tuning parameter $\alpha$.
-For each value of $\alpha$, there is a subtree $T \subset T_0$ such that
-$$\sum_{m=1}^{|T|} \sum_{i : x_i \in R_m} \left( y_i - \mu_{R_m} \right)^2 + \alpha |T|$$
-is as small as possible, where $|T|$ represents the number of terminal nodes of the tree T, $R_m$ is the subset of the predictor space corresponding to the $m$-th terminal node, $\mu_{R_m}$ is the predicted response associated with $R_m$.
-The tuning parameter $\alpha$ controls the trade-off between a sub-tree complexity and and its fit to the training data, the higher the $\alpha$, the more terminal nodes a tree has, the smaller a subtree is.
-The value for $\alpha$ can be chosen using a validation set or trough cross-validation.
-
-\subsection{Pros and Cons of decision trees}
+\subsection{Pros and cons of decision trees}
 Overall, regression trees offer much greater flexibility in handling the relation between covariates and output than linear models do.
-This comes from the fact that fitting a regression tree amounts to estimate a stepwise relation between $y$ and the covariates $x$ that can proxy well non-linear patterns.
-They can more naturally incorpo
-rate interaction effects.
-
+This comes from the fact that fitting a regression tree amounts to estimate a stepwise relation between $y$ and the covariates $x$ that can proxy well non-linear patterns and interaction effects between covariates.
 Greater flexibility comes in handy also when dealing with missing values.
 Tree offer an effective way to overcome this issue, by constructing surrogate variables.
-Exploiting correlation between predictors, they may decrease the effect of missing data.
-
-So far we considered only binary splits, thus an interesting question would be if a model with more than two splits at each note is a valid idea.
-However, it turns out that in general is not a good strategy.
-The issue is that multiple splits fragment the dataset too quickly.
-Moreover, since multiple splits can be achieved by a series of binary splits, the latter approach is preferred.
-
-The interpretability of trees has been one of the driving forces of its success.
-Indeed, thanks to their structure, is straightforward to understand why, for a given $x$, we expect a particular value of $y$, following a series of decision rules.
-Small trees are visually easy to interpret.
+Moreover, trees are visually easy to interpret.
+Thanks to their structure, is straightforward to understand why, for a given $x$, we expect a particular value of $y$, following a series of decision rules.
 Moreover, trees build in such a way are a computationally fast method, that easily scales with large datasets.
 
-However, one of the main disadvantages of this setting, is that trees are non-robust.
-They suffer from high variance: a small change in the data can lead to a large change in the estimation.
+However, one of the main disadvantages of single trees is that trees are non-robust.
+In general, they suffer from high variance: a small change in the data can lead to a large change in the estimation.
 The main reason of this instability is the hierarchical structure of the process as the effect of an error in one of the first nodes is propagated to the other nodes below it.
-This is heightened by the fact that single trees tend to overfit, and spurious correlations in the training data can be mistaken as structural patterns and included in the decision rules.
-This may lead to bad out-of-sample predictive performance.
-High variance is the main reasons that simple regression trees are generally not used.
-Instead, ensemble methods are preferred.
+Indeed, recursive binary splitting is likely to produce good prediction on the training-set, but overfit the training data, leading to poor out-of-sample predictive performance.
+The reason is that this procedure is likely to generate a complex tree, with a low bias but high variance in prediction.
+A possible solution could be constraining ex ante the size of trees, generating a small tree with low variance but possibly high bias, since potentially good splits further down in tree would possibly remain unexploited.
+A better strategy, called \textit{weakest link pruning}, is to grow a very large tree $T_0$, and then prune it back in order to obtain a subtree.
 
-\section{Frequentist ensemble methods}
+However, the predictive performance of pruned trees still falls short of what models combining multiple trees can achieve.
+These models are called tree ensemble methods and are discussed in the following section.
+
+\section{Frequentist tree ensemble methods}
 \label{frequentist}
 
-We highlighted that individual decision trees tend to overfit and to be sensitive to the training data they were built on. To counteract this, ensemble methods were proposed where multiple models
-are aggregated or averaged over to give a more stable and
-generalisable solution.
-
-In general, ensemble methods combine the fit from many (hundreds, thousands) simpler models to get an overall predictor. 
+We highlighted that individual decision trees tend to overfit and their prediction to be highly sensitive to the training data they were built on.
+A general solution, not only in the case of trees, to improve predictive performance of a model are ensemble methods.
+Ensemble methods combine the fit from many (hundreds or thousands) simpler models to get an overall predictor.
 All the models we are going to discuss, namely Bagging, Boosting, Random Forest and Bayesian Additive Regression Trees are ensemble methods, for which the simpler models consist of regression trees.
@@
 \subsection{Bagging}
-Bagging, which stands for Bootstrap Aggregation, is a general technique that aims at reducing the variance of a statistical learning method.
-It is baseed on the fact that, given independent observation, averaging over them reduces variance.
-
+Bagging, which stands for Bootstrap Aggregation, is an ensemble method that aims at reducing the variance of a generic statistical model.
+It is based on the fact that, given independent observation, averaging over them reduces variance.
 For the case of regression trees, the main problem of using a single tree as a prediction method is that perturbing the learning set can cause significant changes in the predictor constructed.
@@
 
-Hence, we resort to bootstrap.
-Bootstrapping consists of taking repeated samples, with replacement, from the single available training set.
+Bootstrapping consists of taking repeated samples, with replacement, from the single available training set to simulate sampling many different training sets.
 In such a way, we are able to construct $B$ different bootstrapped training sets.
-Bagging involves creating multiple $B$ copies of the original training set using hte bootstrap, fitting a separate decision tree to each copy.
-Then it combines the results from each of the $B$ trees to get an overall prediction.
-The single predictions for the target function $f(X)$ for each sample are averaged, as
+In bagging we fit a separate decision tree to each copy of the original training set, and then average the predictions $\hat{y}$ from each of the $B$ trees to get an overall prediction.
+The single predictions for the target function $f(\mathbf{x})$ for each sample are averaged, as
 \begin{equation}
   \label{eq_bagging}
-  \hat{f}_{\text{bag}}(x) = \frac{1}{B} \sum_{b=1}^{B} \hat{f}^{*b}(x).
+  \hat{f}_{\text{bag}}(\mathbf{x}) = \frac{1}{B} \sum_{b=1}^{B} \hat{f}^{*b}(\mathbf{x}).
 \end{equation}
@@
 Averaging these different trees reduce the variance.
-Bagging can dramatically reduce the variance of unstable procedures such
-as trees, leading to improved prediction.
-
-The drawback of bagging is that we lose interpretability to improve accuracy. When we bag a large number of trees, we can no longer represent the model with a single tree.
+However, the drawback of bagging (and of ensemble methods in general) is the loss of interpretability required in order to improve accuracy.
+Indeed, when we bag a large number of trees, we can no longer represent the model with a single tree and directly interpret the relations between covariates throught the structure of this tree.
 
@@
 This is the case, for example, if there is very strong predictor in the dataset.
-If this happens, in the collection of bagged trees a vast majority of the trees will use this strong predictor in the top split and they will all be hihgly correlated among themselves.
-
-Correlation among the trees decrease the effectiveness of averaging, as it does not lead to a large reduction in variance, making the entire bagging algorithm less useful.
+If this happens, in the collection of bagged trees a vast majority of the trees will use this strong predictor in the top split and they will all be highly correlated among themselves.
+Correlation among the trees decrease the effectiveness of averaging, as it does not lead to a large reduction in variance as it would be the case with independent trees.
+This makes the entire bagging algorithm less useful.
 
 Random Forests starts from bagging and adds another kind of randomization to reduce correlation among trees.
-Rather than searching over all the covariate space for each decision rule, as the greedy building of the big trees imposes, in random forests each time we make a split in a tree we randomly sample a subset of $m$ covariates to search over.
+Rather than searching over all the covariate space for each decision rule, as the greedy algorithm used for building of the single trees imposes, in random forests each time we make a split in a tree we randomly sample a subset of $m$ covariates to search over.
 Therefore, the split is allowed to consider only one of those $m$ predictors.
 Moreover, at each split is considered a new sample of $m$ predictors.
-This makes the big trees explore more the space of models.
-
- Usually, we choose $m \approx \sqrt{p}$.
-On average, $(p - m)/ = p$ of the splits will not even consider the strong predictor, and so other predictors will have more of a chance.
-Therefore, we can think of this process as decorrelating the trees, making the average of the resulting trees less variable and hence more reliable.
+On average, $(p - m)/ = p$ of the splits will not even consider the strong predictor, and so other predictors will have a higher probability to be selected.
+This increases the exploration of the  space of models and reduces correlation among trees, making the average of the resulting trees less variable and hence more reliable.
 
@@
 Differently from bagging and random forests, boosting does not involve fitting independent trees from many bootstrap samples and averaging the predictions to reduce variance.
-Instead, it relies on fitting trees sequentially, that is, using information from previously grown trees.
-Indeed, trees are not fit to bootstrapped versions of the original datasets, but rather to the residual not explained by the remainder of
-the trees.
-Then, its prediction is added, with a weight, to the prediction of the previous trees.
-This procedure is iterarated until we express the overall prediction of the boosted trees as trhe weighted average of the prediction of the single trees.
-This is shown in algorithm \ref{alg_boosting}. 
+Instead, it relies on fitting trees sequentially, that is using information from previously grown trees.
+Indeed, trees are not fit to bootstrapped versions of the original datasets, but rather to the residual not explained by the remainder of the trees.
+Then, each new tree's prediction is added, with a weight, to the prediction of the previous trees.
+This procedure is iterated until the prediction of the boosted trees is the weighted average of the prediction of all the single trees.
+This is shown in algorithm \ref{alg_boosting}, adapted from \cite{jamesIntroductionStatisticalLearning2021}.
 \begin{algorithm} [h]
@@
 
-  Set $\hat{f}(x) = 0$ and $r_i = y_i$ for all $i$ in the training set.\;
+  Set $\hat{f}(\mathbf{x}) = 0$ and $r_i = y_i$ for all $i$ in the training set ($i= 1, \ldots, n$).\;
   \For{$b = 1, 2, \ldots, B$}{
-      (a) Fit a tree $f^b$ with $d$ splits ($d+1$ terminal nodes) to the training data $(X, r)$.\;
+      (a) Fit a tree $f^b$ with $d$ splits ($d+1$ terminal nodes) to the training data $[(\mathbf{x}_1, \ldots \mathbf{x_n)}, (r_1, \ldots, r_n)]$.\;
       (b) Update $\hat{f}$ by adding in a shrunken version of the new tree:
       \begin{equation}
-      \hat{f}(x) \gets \hat{f}(x) + \lambda f^b(x).
+      \hat{f}(\mathbf{x}) \gets \hat{f}(\mathbf{x}) + \lambda f^b(\mathbf{x}).
       \end{equation}\;
@@
       \begin{equation}
-      r_i \gets r_i - \lambda f^b(x_i).
+      r_i \gets r_i - \lambda f^b(\mathbf{x}_i).
       \end{equation}\;
@@
   \begin{equation}
-  \hat{f}(x) = \sum_{b=1}^B \lambda f^b(x).
+  \hat{f}(\mathbf{x}) = \sum_{b=1}^B \lambda f^b(\mathbf{x}).
   \end{equation}\;
@@
 By fitting sequentially new trees on the residual of the previous tree, boosting improves at each iteration the fit marginally.
-The parameter $d$, which determines the number of terminal nodes, andh the parameter $\lambda$, which weights the contirbution of each tree to the overall prediction, constraint the single trees to be weak learners.
-By  fitting small trees to the residuals, we slowly improve $\hat{f}$ in areas where it
-does not perform well. It is important to mention that in boosting, unlike in bagging, the construction of each tree strongly depends on the trees that have already grown.
+The parameter $d$, which determines the number of terminal nodes, and the parameter $\lambda$, which weights the contribution of each tree to the overall prediction, constrain each single trees to be a \textit{weak learner}: rather than attempting to fit the overall pattern in the data, each tree focuses on capturing the residual pattern that other trees did not capture.
+By sequentially fitting small trees to the residuals, we slowly improve $\hat{f}$ in areas where it
+does not perform well.
+It is important to mention that in boosting, unlike in bagging, the construction of each tree strongly depends on the trees that have already grown: trees are highly dependent from each other.
 
-The model accounts for three different tuning parameters.
-The number of of trees $B$, chosen trough cross-validation.
-Interestingly, unlike bagging and random forests, boosting can suffer of overfitting if $B$ is too large, since tree are strongly dependent from each other.
-The shrinkage parameter $\lambda$, a small positive number.
-Common values for $\lambda$ are 0.01 or 0.001, depending on the problem at hand.
-However, very small values of $\lambda$ often require a very large B to achieve good performances.
+The model accounts for three different parameters that must be tuned.
+Firstly, the number of of trees $B$, is chosen trough cross-validation.
+Unlike bagging and random forests, boosting can suffer of overfitting if $B$ is too large, since tree are strongly dependent from each other.
+Then there is the shrinkage parameter $\lambda$, a small positive number.
+Common values for $\lambda$ are $0.01$ or $0.001$, depending on the problem at hand.
+However, very small values of $\lambda$ often require a very large B to achieve good performance.
 Lastly, $d$ represents the number of splits in each tree.
-Often $d=1$ works well.
-However in this case the tree is called stump since it contains only a single split.
+Often, $d=1$ works well: int this case the tree is called a stump since it contains only a single split.
 
@@
 
-The idea for Bayesian Additive Regression Trees (BART) was first developed by Chipman, George, McCulloch (2010 AOAS).
-Like other ensemble methods (bagging, boosting, random forest), to solve this regression problem \eqref{regression}, BART approximates \( f(x) = \mathbb{E}(y \mid x) \) by a sum of regression trees.
-However, being a Bayesian method, BART requires specifying a full probability model for the data, composed by a likelihood and a prior. 
-The likelihood is obtained by assuming normality of the error in the regression \eqref{regression}. In other words, BART assumes 
+The idea for Bayesian Additive Regression Trees (BART) was first developed by \cite{chipmanBARTBayesianAdditive2010}.
+Like other ensemble methods (bagging, boosting, random forest), to solve this regression problem \eqref{regression}, BART approximates \( f(\mathbf{x}) = \mathbb{E}(y \mid \mathbf{x}) \) by a sum of regression trees.
+Its specificify comes from the fact that, being a Bayesian method, BART requires a full probability model for the data, composed by a likelihood and a prior.
+The likelihood is obtained by assuming normality of the error in the regression \eqref{regression}. BART thus assumes 
 \begin{equation}
-Y = f(X) + \epsilon, \quad \epsilon \sim \mathcal{N}(0, \sigma^2), \label{regression}
+y = f(\mathbf{x}) + \epsilon, \quad \epsilon \sim \mathcal{N}(0, \sigma^2), \label{regression2}
 \end{equation}
@@
 The function $f$ is modeled with a sum of $m$ regression trees, which are the remaining parameters of the model.
-As before, we separate the parameter space into the tree structure $T$, which consists of all the decision rules, and $M$, which is the set of leaves or terminal nodes.
-Leaves are going to be denoted by $\mu$, as they are simply means in a  regression setting.
+As before, we separate the parameter space into the tree structure $T$, which consists of all the decision rules, and $M$, which is the set of leaves or terminal nodes, which are denoted by $\mu$.
 
-Thus, in BART the relationship between $(y, X)$ is modeled as: 
+BART is completed by a prior distribution on the parameters $ (\{T_j, M_j \}_{j=1}^m, \sigma)$, 
 \begin{equation}
-y = \sum_{j=1}^m g(x; T_j, M_j) + \epsilon, \quad \epsilon \sim N(0, \sigma^2) \label{firstlikelihood}
+p((T_1, M_1), (T_2, M_2), \ldots, (T_m, M_m), \sigma). \label{firstprior}
 \end{equation}
-where for each binary regression tree \( T_j \) and its associated terminal nodes  \( M_j \), \( g(x; T_j, M_j) \) is the function assigning \( \mu_{ij} \in M_j \)  to \( x \), and \( \mu_{ij} \in M_j \) will be the predicted value of $y$ for $x$.
-
- The BART model is completed by a prior on the parameters $ (\{T_j, M_j \}_{j=1}^m, \sigma)$, 
-\[
-p((T_1, M_1), (T_2, M_2), \ldots, (T_m, M_m), \sigma).
-\]
 The essential idea of BART is to exploit this prior to keep the individual tree effects in the fit small and avoid overfitting.
 Indeed, while the number $m$ of trees is fixed, the overall number of parameters (that is the depth of each trees and the number of terminal nodes) is left unspecified.
-This should make BART prone to overfit the data by growing large and complex trees.
-To overcome this problem, BART uses a regularization prior, that is a prior set in a way to shrink the model towards simplicity.
-As an effect of shrinkage, each tree is forced to explain only a limited subset of the relationships between the covariates and the predictor variable.
-This is similar to what boosting does, with the difference that in boosting, being a frequentist method, shrinkage is regulated by a parameter $\lambda$, which needs tuning.
+In absence of an appropriate prior, this would make BART prone to grow large trees and overfit the data.
+To avoid this, the prior \eqref{firstprior} is set in a way to shrink the model towards simplicity: it is a so called regularization prior.
+As an effect of shrinkage, each tree is forced to be a weak learner and explain only a limited subset of the relationships between the covariates and the predictor variable.
+This is similar to what boosting does, with the difference that in boosting, being a frequentist method, shrinkage is regulated by a parameter $\lambda$, which needs tuning, while in BART it is done automatically by the prior.
 
-Since the space of parameters $ (\{T_j, M_j \}_{j=1}^m, \sigma)$ is high and random dimensional (remember that only the number of trees is specified, but not the numbers of decision rules or of leaves), BART uses Markov Chain Monte Carlo (MCMC) to get draws from posterior distribution.
-At every iteration of the MCMC, the software generates a representation of $f$ and a value for $\sigma$ given the data.
-A single posterior mean estimate of \( f(x) = \mathbb{E}(Y \mid x) \) at any input value $x$ is obtained by a simple average of these successive draws evaluated at \( x \).
-Further, pointwise uncertainty intervals for \( f(x) \) are easily obtained from the corresponding quantiles of the sample of draws.
+Since the space of parameters $ (\{T_j, M_j \}_{j=1}^m, \sigma)$ is high and random dimensional (only the number of trees is specified, but not the numbers of decision rules or of leaves), BART uses Markov Chain Monte Carlo (MCMC) to get draws from posterior distribution.
+Every iteration of the MCMC generates a representation of $f$ and a value for $\sigma$ given the data.
+A single posterior mean estimate of \( f(\mathbf{x}) = \mathbb{E}(y \mid \mathbf{x}) \) at any input value $\mathbf{x}$ is obtained by a simple average of these successive draws evaluated at \( \mathbf{x} \).
+Further, pointwise uncertainty intervals for \( f(\mathbf{x}) \) are easily obtained from the corresponding quantiles of the sample of draws.
 
-Overall, BART can be viewed as a Bayesian nonparametric approach (each tree has unknown size) that fits a parameter-rich model like \eqref{firstlikelihood} using a regularization prior to avoid overfitting.
-We are going to examine the a sum-of-trees model and implied likelihood first, and then look at the regularization prior.
+BART can be implement in R using the \texttt{gbart} package, created by \cite{chipmanBARTBayesianAdditive2010}.
+We will refer to this package when commenting the default choice of the hyperparamters of the prior.
+In this section, we examine the tree constituents of BART, the likelihood, the prior and the MCMC sampler form a theoretical standpoint.
 
 \subsection{A Sum-of-Trees Model}
-As before, and for technical reasons that will be clear in the posterior sampling, BART models each tree separating it into $T$ and $M$.
-We denote by \( T \) the binary tree structure —the various binary split decision rules of the form \( \{x_q < c\} \) versus \( \{x_q \geq c\} \), with \( c \in \mathbb{R} \) and $x_q$ an element of the vector of covariates $x$.
-The vector of parameters associated with \( T \), i.e., the collection of terminal node parameters, is denoted by \( M = \{\mu_1, \ldots, \mu_b\} \), where \( b \) is the number of terminal nodes.
-The $(T, M)$ are to be thought about as a parameter $\theta$ in prior specification and posterior sampling.
+As before, and for technical reasons that will be clear in posterior sampling, BART models separately $T$, the binary tree structure, composed by the decision rules, and $M$, the collection of terminal node parameters.
+The $(T, M)$ couple are the parameters of the likelihood.
 
-We define a regression tree as a function \( g(x; T, M) \) that assigns the conditional mean \( \mathbb{E}[y \mid x] \) to the parameter \( \mu_i \in M \), i.e., \( \mu_i = g(x; T, M) \rightarrow \mathbb{E}[y \mid x] \) via binary decision rules denoted as \( T \).
-With this notation, the sum-of-trees model can be more explicitly expressed as
+We define a regression tree as the function \( g(\mathbf{x}; T, M) \) that assigns the conditional mean \( \mathbb{E}[y \mid \mathbf{x}] \) to the parameter \( \mu_i \in M \), i.e., \( \mu_i = g(\mathbf{x}; T, M) \rightarrow \mathbb{E}[y \mid \mathbf{x}] \), via the binary decision rules denoted as \( T \).
+With this notation, the sum-of-trees model in \eqref{regression2} can be more explicitly expressed as
 \begin{equation}
-Y = \sum_{j=1}^m g(x; T_j, M_j) + \epsilon, \quad \epsilon \sim N(0, \sigma^2) \label{likelihood}
+y= \sum_{j=1}^m g(\mathbf{x}; T_j, M_j) + \epsilon, \quad \epsilon \sim N(0, \sigma^2) \label{likelihood}
 \end{equation}
-where for each binary regression tree \( T_j \) and its associated terminal node parameters \( M_j \), \( g(x; T_j, M_j) \) is the function which assigns \( \mu_{j} \in M_j \) to \( x \).
-Under \eqref{likelihood}, \( \mathbb{E}(Y \mid x) \) equals the sum of all the terminal node \( \mu_{j} \)'s assigned by the $m$ \( g(x \mid T_j, M_j) \)'s.
+where for each binary regression tree \( T_j \) and its associated terminal node parameters \( M_j \), \( g(\mathbf{x}; T_j, M_j) \) is the function which assigns \( \mu_{j} \in M_j \) to \( \mathbf{x} \).
+Under \eqref{likelihood}, \( \mathbb{E}(y \mid \mathbf{x}) \) equals the sum of all the terminal node \( \mu_{j} \)'s assigned by the $m$ \( g(\mathbf{x} \mid T_j, M_j) \)'s.
 
- As all others sum-of-tree models, the quantity that is being summed and eventually allocated to \( \mathbb{E}(y \mid x) \) is not the regression tree or tree structure, but the value that each \( j \)th tree structure assigns to the subject, that is \( \mu_{j} \).
+ As all others sum-of-tree models, the quantity that is being summed and eventually allocated to \( \mathbb{E}(y \mid \mathbf{x}) \) is not the regression tree or tree structure, but the value that each \( j \)th tree structure assigns to the subject, that is \( \mu_{j} \).
 This makes single regression tree is a simple stepwise function, and summing trees amounts to summing together these stepwise functions.
-As a result, the resulting function $f(x)$  can approximate the nonlinearities and  interactions effects between the covariates $x$.
+As a result, the resulting function $f(\mathbf{x})$  can approximate the nonlinearities and interactions effects between the covariates $x$.
 
@@
 \[
-p(y \mid T_j, M_j, \sigma^2, x) = \mathcal{N}\left(y \mid \sum_{j=1}^{m} g(x; T_j, M_j), \sigma^2\right),
+p(y \mid T_j, M_j, \sigma^2, \mathbf{x}) = \mathcal{N}\left(y \mid \sum_{j=1}^{m} g(\mathbf{x}; T_j, M_j), \sigma^2\right),
 \]
-and the likelihood for the entire training dataset (note that $Y = (y_1, \ldots, y_n)$) is:
+and the likelihood for the entire training dataset, calling $Y= (y_1, \ldots, y_n)$ and $\mathbf{X} =(\mathbf{x}_1,\ldots, \mathbf{x}_n)$, is:
 \[
-p(Y \mid T_j, M_j, \sigma^2, X) = \prod_{i=1}^{N} \mathcal{N}\left(y_i \mid \sum_{j=1}^{m} g(x_i; T_j, M_j), \sigma^2\right).
+p(Y \mid T_j, M_j, \sigma^2, \mathbf{X}) = \prod_{i=1}^{N} \mathcal{N}\left(y_i \mid \sum_{j=1}^{m} g(\mathbf{x}_i; T_j, M_j), \sigma^2\right).
 \]
@@
 Consider the parameter space $(\{T_j, M_j\}_{j=1}^m, \sigma)$ of model \eqref{likelihood}.
-The only identified parameter is $\sigma$, because it's the only one that can be estimated as the variance of the residulas.
-For the trees, only their number $m$ is fixed, while the individual trees remain unidentified and interchangeable.
-Indeed, the dimension of each tree (namely the number of internal and terminal nodes) is left unbounded.
+The only identified parameter is $\sigma$, because it is the only one that can be estimated as the variance of the residuals.
+For the trees, only their number $m$ is fixed, while the individual trees remain unidentified: thus, the dimension of each tree (namely the number of internal and terminal nodes) is left unbounded.
 If data demands it, the trees can grow complex, with many leaves and nodes, but they can also remain simple.
@@
 The dimension of each tree is going to be inferred from the data.
-This high-dimensional and random structure underpins BART's ability to model complex data flexibly and effectively.This interchangeability means the trees themselves lack direct interpretability.
+This high-dimensional and random structure underpins BART's ability to model complex data flexibly and effectively.
 Despite the large number of parameters, BART avoids overfitting through the use of an appropriate regularization prior.
@@
 
-Since each tree \( g(x; T_j, M_j) \) in \eqref{likelihood}, including the variables to split on, the splitting values, and the mean parameters at each terminal node, is unknown, a prior distribution 
+Since each tree \( g(\mathbf{x}; T_j, M_j) \) in \eqref{likelihood}, including the variables to split on, the splitting values, and the mean parameters at each terminal node, is unknown, a prior distribution
 \[
@@
 
-Since the parameter space is high and random dimensional, to avoid overfitting the prior must be chosen in order to keep the contribution of each \(g(x; T_j, M_j)\) small.
+Since the parameter space is high and random dimensional, to avoid overfitting the prior must be chosen in order to keep the contribution of each \(g(\mathbf{x}; T_j, M_j)\) small.
 The prior is thus a regularization prior, that shrinks the model towards simplicity.
-But there is also a connection to boosting: indeed, the prior ensures that each tree in the ensemble acts as a weak learner.
-Rather than attempting to fit the overall pattern in the data, each tree focuses on capturing the residual pattern that other trees did not capture.
+The shrinkage is not carried out by reducing the number of trees, but instead by ensuring that each tree in the ensemble acts as a weak learner.
+This is a clear connection to boosting, where this function is performed by parameters $\lambda$, and $d$, and that needs tuning.
 
 \subsubsection{Prior independence and Symmetry}
-In order to simplify the specification of the regularization prior, Chipman assumes that  the trees \( \{(T_1, M_1), \ldots, (T_m, M_m)\} \) are independent of each other and of  \( \sigma \).
+In order to simplify the specification of the regularization prior, \cite{chipmanBARTBayesianAdditive2010} assumes that the trees \( \{(T_1, M_1), \ldots, (T_m, M_m)\} \) are independent of each other and from \( \sigma \).
 Then, the prior distribution can be written, using the same prior for each $(T, M)$ pair, as:
-
 \[
@@
 Then \( M_j = \{\mu_1, \mu_2, \ldots, \mu_{b_j}\} \) is the vector of terminal node parameters associated with \( T_j \).
-Chipman at al.
-assume that node parameters \( \mu_{ij} \in M_j\) are iid, that is $\mu_{ij} \overset{\mathrm{iid}}{\sim} p(\mu_j)$ This implies:
+\cite{chipmanBARTBayesianAdditive2010} assume that node parameters \( \mu_{ij} \in M_j\) are iid, that is $\mu_{ij} \overset{\mathrm{iid}}{\sim} p(\mu_j)$ This implies:
 \[
@@
 The idea of regularization prior is to choose them so to have each $T$ small, each $\mu$ small, and an appropriate $\sigma$  (e.g.
-smaller than the least squares estimate).
+smaller than the least squares estimate, if we reasonably assume that BART will fit a better model than a simple linear regression).
 
 \subsubsection{The $T_j$ prior}
- Instead of a probability distribution, Chipman specifies a process that can be used to draw a tree from the prior. The \( T_j \) prior, \( p(T_j) \), is specified by three aspects:
+ Instead of a probability distribution, \cite{chipmanBARTBayesianAdditive2010} specifies a process that can be used to draw a tree from the prior. The prior \( p(T_j) \) is specified by three aspects:
 \begin{enumerate}
-    \item The probability that a node at depth \( d = (0, 1, 2, \ldots) \) is nonterminal is given by:
+    \item The probability that a node at depth \( d = (0, 1, 2, \ldots) \) is nonterminal, given by:
     \[
@@
     where node depth $d$ is defined as the distance from the root.
-In default \texttt{wbart}, $\alpha$ is set 0.95 and $\beta$ is set equal to 2.
+In default \texttt{gbart}, $\alpha$ is set 0.95 and $\beta$ is set equal to 2.
 As the tree grows and the depth increases, the probability of splitting decreases.
@@
 Nonetheless, trees with many terminal nodes can be grown if the data demands it.
-    \item The distribution used to select the covariate to split upon in an internal node.This is the uniform prior on available variables in vector $x$.
-    \item The distribution used to select the cutoff point in an internal node conditional on the covariate selected.
-In \texttt{wbart}, this is the uniform prior on the discrete set of available splitting values $c$.
+    \item The distribution used to select the covariate to split upon in an internal node. This is the uniform prior on available variables in vector $\mathbf{x}$.
+    \item The distribution used to select the cutoff point $c$ in an internal node conditional on the covariate selected.
+In \texttt{gbart}, this is the uniform prior on the discrete set of available splitting values $c$.
 \end{enumerate}
@@
 \subsubsection{The \( \mu_{i,j} | T_j \) Prior}
-$\mu_{ij} | T_j$ are assumed to be iid. The prior specification for the leaves is $$ \mu_{ji} |T_j \overset{\text{iid}}{\sim}N(\mu_\mu, \sigma_\mu^2)$$
+The leaves $\mu_{ij} | T_j$ are assumed to be iid. Their prior specification is $$ \mu_{ji} |T_j \overset{\text{iid}}{\sim}N(\mu_\mu, \sigma_\mu^2)$$
 To specify the hyperparameters $\mu_\mu, \sigma_\mu$, \cite{chipmanBARTBayesianAdditive2010} turn to empirical Bayes, that is they calibrate the prior using the observed variation in \( y \).
-In particular, note that under equation \eqref{likelihood}, each $y_i$ is the sum of an error $\varepsilon_i$ and $m$ $\mu_{ij}$.
-Thus, given $\mu_{ij} \mid T_j \sim N(\mu_\mu, \sigma^2_\mu)$, the induced prior on $E(Y \mid x)$ is \(N(m\mu_\mu, \sigma^2_\mu)\).
-Since we expect that \(\mathbb{E}(Y \mid x)\) lies between \(y_{\text{min}}\) and \(y_{\text{max}}\), we choose \(\mu_\mu\) and \(\sigma_\mu\) such that the induced prior distribution of $E(Y \mid x)$ assigns substantial probability to the interval \((y_{\text{min}}, y_{\text{max}})\).
+Under equation \eqref{likelihood}, each $y_i$ is the sum of an error $\varepsilon_i$ and $m$ $\mu_{ij}$.
+Thus, given the prior distribution $\mu_{ij} \mid T_j \sim N(\mu_\mu, \sigma^2_\mu)$, the induced prior on $E(y \mid \mathbf{x})$ is \(N(m\mu_\mu, \sigma^2_\mu)\).
+Since we expect that \(\mathbb{E}(y \mid \mathbf{x})\) lies between \(y_{\text{min}}\) and \(y_{\text{max}}\), the hyperparameters \(\mu_\mu\) and \(\sigma_\mu\) are chosen so that the induced prior distribution of $E(y \mid \mathbf{x})$ assigns substantial probability to the interval \((y_{\text{min}}, y_{\text{max}})\).
 
-To choose $\mu_\mu$, \( y \) is centered around zero by subtracting its sample meand and it is rescaled so that the observed transformed values range from \( y_{\text{min}} = -0.5 \) to \( y_{\text{max}} = 0.5 \).
+To choose $\mu_\mu$, \( y \) is centered around zero by subtracting its sample mean $\overline{y}$ and it is rescaled so that the observed transformed values range from \( y_{\text{min}} = -0.5 \) to \( y_{\text{max}} = 0.5 \).
 Then the prior we can choose for $\mu_{ij}$ is zero.
@@
 for some suitable value of $k$ (generally $k=2$).
-
-As \(k\) and/or \(m\) (the number of trees) increase, such a prior shrinks the tree leaves \(\mu_{ij}\) toward zero, limiting the influence of individual tree components and thus making it a weak learner.
+As \(k\) and/or \(m\) (the number of trees) increase, such a prior shrinks the tree leaves \(\mu_{ij}\) toward zero, limiting the influence of individual tree components and thus contributing to make it a weak learner.
 
@@
 \]
-We use the the inverse chi-square because it is a conjugate prior for the variance of the normal distribution, and the distribution of the errors $\varepsilon$ in \eqref{likelihood} is assumed to be normal in our probability model: this will make the draw of $\sigma$ from the posterior easy.
+This choice is motivated by the fact that the inverse chi-square because is the conjugate prior for the variance of the normal distribution, and the distribution of the errors $\varepsilon$ in \eqref{likelihood} is assumed to be normal in our probability model.
+Thus the posterior value of $\sigma$ will have a closed form and posterior inference for it will be easy.
 
@@
 substantial probability to the entire region of plausible values of $\sigma$.
-The estimate $\hat{\sigma}$ produced by the software \texttt{wbart} we use in our simulation can be either the sample standard deviation of \( y \) (if the number of covariates is more of the numebr of observations), or the residual standard deviation from a least squares linear regression of \( y \) on the original \( X \).
+The estimate $\hat{\sigma}$ produced by the software \texttt{wbart} we use in our simulation can be either the sample standard deviation of \( y \) (if the number of covariates $p$ is more of the number of observations $n$), or the residual standard deviation from a least squares linear regression of \( y \) on the original \( X \).
 The value of \( \nu \) is then picked between 3 and 10 to get an appropriate shape of the distribution of $\sigma$, while the value of \( \lambda \) is chosen so that \( p(\sigma < \hat{\sigma}) = q \), where $q = 0.75, 0.95, 0.99$.
@@
 \subsection{ Posterior sampling}
-Given the observed data \( y \), our Bayesian setup induces a posterior distribution:
+Given the observed data \( Y = (y_1, \ldots, y_n) \), our Bayesian setup (the likelihood and prior) induces a posterior distribution:
 \[
 \begin{aligned}
-p\left( \{T_j, M_j\}_{j=1}^m, \sigma^2 \mid Y, X \right) &\propto p\left( Y \mid \{T_j, M_j\}_{j=1}^m, \sigma^2, X \right) \times p\left( \{T_j, M_j\}_{j=1}^m, \sigma \right)
+p\left( \{T_j, M_j\}_{j=1}^m, \sigma^2 \mid Y, \mathbf{X}\right) &\propto p\left( Y \mid \{T_j, M_j\}_{j=1}^m, \sigma^2, \mathbf{X}\right) \times p\left( \{T_j, M_j\}_{j=1}^m, \sigma \right)
 \end{aligned}
@@
 However, a MCMC algorithm can be used to sample from this posterior.
-The algorithm allows to obtain a dependent sample from the posterior
- distribution over the space of sum-of-trees models.
+The algorithm generates a Markov chain of samples from the posterior distribution over the space of parameters, each of which is correlated with nearby samples.
 
-This algorithm is a Gibbs sampler, in which one parameter is drawn from the full conditional, and then you cycle through the others.
-In particular, the draw of all parameters from the posterior can be simplified into two major draws from full conditionals: first $m$ successive draws of the trees, and then a final draw for $\sigma$.
+In the BART model, the general MCMC algorithm is a Gibbs sampler, in which, starting from an initial value of the random vector of parameters $ (\{T_j, M_j\}_{j=1}^m, \sigma^2) $, each parameters is drawn from its full conditional, and then you cycle through the others.
+In particular, the draw of all parameters from the posterior can be simplified into two major draws from full conditionals: first $m$ successive draws of the trees $(T_j,M_j)$, and then a final draw for $\sigma$.
 
@@
 The draw of $\sigma$ from \eqref{priorsigma} is made easy by the choice of a conjugate prior.
-Indeed, if all the tree models $(T_i, M_i)$ are known, then the function $f$ is known, and we can compute $y$ minus$ f(x)$ to get the residuals of \eqref{firstlikelihood}.
-Since hte residulas are assumed to be distributed normally, and the inverse chi-squared distribution is (conditionally) the conjugate prior for the variance of the normal distribution, the draw of \( \sigma^2 \) from the posterior is simply a draw from an inverse chi-squared distribution with updated parameters.
+Indeed, conditional on all the tree models $(T_i, M_i)$, the function $f(\mathbf{x})=\sum_{j=1}^m g(\mathbf{x}; T_j, M_j) $ is known, and we can compute $y - f(\mathbf{x}) = \varepsilon $, the residuals of \eqref{likelihood}.
+Since the residuals are assumed to be distributed normally, and since the inverse chi-squared distribution of $\sigma$ is the conjugate prior for the variance of the normal distribution, the draw of \( \sigma^2 \) from the full posterior is simply a draw from an inverse chi-squared distribution with updated parameters.
+
 To get the \( m \) draws of the tree models \( (T_j, M_j) \) from \eqref{priortree}, we take advantage of the following reductions. First, observe that the full conditional:
@@
 \]
-depends on \( (T_{(-j)}, M_{(-j)}, y) \) only through:
+depends on \( (T_{(-j)}, M_{(-j)}, y) \) only through a vector of $n$ residuals:
 \begin{equation}
-R_j = y - \sum_{k \neq j} g(x; T_k, M_k), \label{residual}
+E_j = Y - \sum_{k \neq j} g(\mathbf{X}; T_k, M_k), \label{residual}.
 \end{equation}
-which for each $j$ is the $n$-vector of residuals of the $m-1$ regression sum-of-trees excluding the \( j \)-th tree.
+For each $j$, $E_j$ is the $n$-vector of residuals of the $m-1$ regression sum-of-trees excluding the \( j \)-th tree, ($E_j = [\hat{\varepsilon}_1, \ldots, \hat{\varepsilon}_n]$ with $\hat{\varepsilon}$ coming from the estimated \eqref{likelihood}).
 Thus, the \( m \) draws of \( (T_j, M_j) \) given \( (T_{(-j)}, M_{(-j)}, \sigma^2, y) \) are equivalent to \( m \) draws from:
 \begin{equation}
-p(T_j, M_j \mid R_j, \sigma^2), \quad j = 1, \ldots, m. \label{13}
+p(T_j, M_j \mid E_j, \sigma^2), \quad j = 1, \ldots, m. \label{13}
 \end{equation}
+According to the definition of conditional probability, the draw from \eqref{13} can be split as such:
+\[
+p(T_j, M_j \mid E_j, \sigma^2) = p(M_j \mid T_j, E_j, \sigma^2) \; p(T_j \mid E_j, \sigma^2).
+\]
+We will sample $T_j, M_j$ conditioned on all the other trees in two successive steps, by first sampling $T_j$ conditional on the other trees and $\sigma$, and then sampling $M_j$ conditioned on all the trees including $T_j$ and $\sigma$.
+In other words, the draw from \eqref{13} is carried out as follows:
+\begin{enumerate}
+    \item  first,  we draw  $T_j$ from :
+\begin{equation}
+ T_j \mid E_j, \sigma^2, \label{MH}
+\end{equation}
+\item then, you draw $M_j$ from :
+\begin{equation}
+M_j \mid T_j, E_j, \sigma^2 \label{drawM}
+\end{equation}
+\end{enumerate}
+The reason behind starting from \( p(T_j \mid E_j, \sigma^2) \) is that the draw of the vector of leaves $M_j$ will be easy when conditioned on $T_j$ due to conjugacy.
 
-Now we use the fact that, by the definition of conditional probability, we can write:
+Using Bayes' Theorem and standard probability laws, 
 \[
-p(T_j, M_j \mid R_j, \sigma^2) = p(M_j \mid T_j, R_j, \sigma^2) \; p(T_j \mid R_j, \sigma^2).
+p(T_j \mid E_j, \sigma^2) \propto p(T_j) \; p(E_j \mid T_j, \sigma^2) 
+\propto p(T_j) \; \int p(E_j, M_j \mid T_j, \sigma^2) \, dM_j.
 \]
-We start from \( p(T_j \mid R_j, \sigma^2) \). Using Bayes' Theorem and standard probability laws, 
-\[
-p(T_j \mid R_j, \sigma^2) \propto p(T_j) \; p(R_j \mid T_j, \sigma^2) 
-\propto p(T_j) \; \int p(R_j, M_j \mid T_j, \sigma^2) \, dM_j.
-\]
-
 Expanding further:
 \[
-p(T_j \mid R_j, \sigma^2) \propto p(T_j) \; \int p(R_j \mid T_j, M_j, \sigma^2) \; p(M_j \mid T_j, \sigma^2) \, dM_j
+p(T_j \mid E_j, \sigma^2) \propto p(T_j) \; \int p(E_j \mid T_j, M_j, \sigma^2) \; p(M_j \mid T_j, \sigma^2) \, dM_j
 \]
-\[
-= p(T_j) \; \int p(R_j \mid T_j, M_j, \sigma^2) \; p(M_j \mid T_j, \sigma^2) \; p(\sigma^2) \, dM_j.
-\]
-
 And finally:
 \begin{equation}
-p(T_j \mid R_j, \sigma^2) \propto p(T_j) \int p(R_j \mid M_j, T_j, \sigma^2) \; p(M_j \mid T_j, \sigma^2) \, dM_j. \label{posterioeT}
+p(T_j \mid E_j, \sigma^2) \propto p(T_j) \int p(E_j \mid M_j, T_j, \sigma^2) \; p(M_j \mid T_j, \sigma^2) \, dM_j. \label{posterioeT}
 \end{equation}
+We notice that $E_j$, the $n$-vector of partial residuals, is normally distributed conditionally on $(T_j, M_j), \sigma^2$.
+Indeed, from the definition of the vector of partial residuals in \eqref{residual}, we have that:
+\[
+E_j = g(\mathbf{X}, T_j, M_j) + \varepsilon
+\]
+and since $\varepsilon \sim N(0, \sigma^2 \mathbf{I}_n)$ by assumption, it follows that $p(E_j \mid M_j, T_j, \sigma^2) $ in \eqref{posterioeT} is a normal.
+Moreover, we have used a conjugate normal prior for $M_j$:
+\[ \mu_{ji} \mid T_j  \overset{\text{iid}}{\sim} N(\mu_\mu, \sigma_\mu^2)
+\]
+Thus, $M_j$ can be integrated it out from \eqref{posterioeT} to obtain $p(T_j \mid E_j, \sigma^2)$ in closed form up to a normalizing constant.
+This is the only thing we need in order to get a sample of $T_j \mid E_j, \sigma^2$ using a Metropolis-Hastings algorithm (MH).
 
-Notice that $R_j$, the $n$-vector of partial residuals, is conditionally normally distributed,and that we have used a conjugate normal prior for $M_j$.
-Thus, it can easily be integrated it out from \eqref{posterioeT} to obtain $p(T_j \mid R_j, \sigma^2)$ in closed form up to a normalizing constant.
-Thus we can just draw $T$ alone from \eqref{posterioeT} and then draw $M_j$ conditionally on it.
-In other words, the draw from \eqref{13} can be split in two successive steps.
-\begin{enumerate}
-    \item  first,  we draw  $T_j$ conditional on all the other trees/ on the residula $R_j$ and on $\sigma$ from \eqref{posterioeT}:
-\begin{equation}
- T_j \mid R_j, \sigma^2, \label{MH}
-\end{equation}
-\item then, you draw $M_j$conditional on all the trees and on $\sigma$:
-\begin{equation}
-M_j \mid T_j, R_j, \sigma^2 \label{drawM}
-\end{equation}
-\end{enumerate}
-The draw of $T_j$ from \eqref{MH} is obtained using a Metropolis-Hastings (MH) algorithm.
-Let \( T_j^{(i)} \), \( M_j^{(i)} \), and \( \sigma^{(i)} \) denote the values of \( T_j \), \( M_j \), and \( \sigma \) at the \( i \)th MCMC iteration, respectively.
-Each draw  \( T_j^{(i)} \mid R_j^{(i)}, \sigma^{(i)} \) from \eqref{MH} is obtained using MH.
-At the MCMC iteration $i$, MH samples $T_j^{(i)}$ by proposing local changes to $T_j^{(i-1)}$.
-CHanges are either accepted or rejected in such a way that it is still true that the posterior is the stationary distribution.
-The proposal changes to $T_J^{(i-1)}$ that MH can carry are the following: 
+Similarly to Gibbs sampler, which is indeed a particular case of Metropolis-Hastings, the Metropolis–Hastings algorithm is a Markov chain Monte Carlo (MCMC) method for obtaining a sequence of random samples from a probability distribution from which direct sampling is difficult, with the only requirement that we know a function proportional to the target distribution.
+In our case, thanks to our prior choice we know the right hand side of \eqref{posterioeT}, so we are able to obtain draws from $p(T_j \mid E_j, \sigma^2)$.
+Let \( T_j^{(i)} \), \( M_j^{(i)} \), and \( \sigma^{(i)} \) denote the values of \( T_j \), \( M_j \), and \( \sigma \) at the \( i^\text{th} \) MCMC iteration.
+At the MCMC iteration $i$, MH samples $T_j^{(i)} \mid E_j^{(i)}, \sigma^{(i)}$ by proposing local changes to $T_j^{(i-1)}$.
+Changes are either accepted or rejected in such a way that it is still true that the posterior is the stationary distribution.
+The proposal changes to $T_j^{(i-1)}$ that MH can carry are the following: 
 \begin{itemize}
@@
 
-Finally, once we have $p(T_j \mid R_j, \sigma^2)$, we are left with the draw $M_j$ from the distribution \eqref{drawM}. Given that the $\mu_{ji}$ are iid, we have: 
-$$ p(M_j \mid T_j, R_j, \sigma^2) = \prod_{i=1}^{b_j} p(\mu_{ij} \mid T_j, R_j, \sigma^2),
-$$ where the prior on $\mu_{ji}$ is  $N(0, \sigma_\mu^2)$.
-Notice that, if you call  \( R_{ji} \) the subsets of residuals in the $n$-vector  $R_j$ that are allocated to the terminal node parameter  \( \mu_{ji} \), and call \( n_i \) their number, then \( R_{ji} \mid T_j, \mu_{ji}, \sigma \sim \mathcal{N}(\mu_{ji}, \sigma^2) \).
-Thus the prior of $\mu_{ji}$ is actually conjugate, and the posterior we are drawing from, $p(\mu_{ji} \mid T_j, \sigma, R_j) \propto p(R_{ji} \mid T_j, \mu_{ji}, \sigma) p(\mu_{ji} \mid T_j)$, is going to be a normal distribution.
-The draw of \( M_j \) from this normal enables the calculation of the subsequent residual \( R_{j+1} \), which allows the successive draw of \( T_j \).
+Finally, once we have $p(T_j \mid E_j, \sigma^2)$, we are left with the draw of $M_j \mid T_j, E_j, \sigma^2$. Given that the $\mu_{ji}$ are iid, we have: 
+$$ p(M_j \mid T_j, E_j, \sigma^2) = \prod_{i=1}^{b_j} p(\mu_{ji} \mid T_j, E_j, \sigma^2),
+$$ where the prior on $\mu_{ji} \mid T_j$ is  $N(0, \sigma_\mu^2)$.
+Notice that, if you call  \( E_{ji} \) the subsets of residuals in the $n$-vector  $E_j$ that are allocated to the terminal node parameter  \( \mu_{ji} \), and call \( n_i \) their number, then \( E_{ji} \mid T_j, \mu_{ji}, \sigma^2 \sim N(\mu_{ji}, \sigma^2) \).
+Thus the prior of $\mu_{ji}$ is actually conjugate, since it is a normal prior for the mean of a normal with known variance.
+Hence the posterior we are drawing from, $p(\mu_{ji} \mid T_j, \sigma, E_j) \propto p(E_{ji} \mid T_j, \mu_{ji}, \sigma) p(\mu_{ji} \mid T_j)$, is going to be a normal distribution.
+The draw of \( M_j \) is just a set of independent draws of $\mu_{ji}$ from this normal distribution.
+Finally, the draw of $M_{ji}$  from this normal enables the calculation of the subsequent residual \( E_{j+1} \), which allows the successive draw of \( T_j \).
 
@@
 
-  \textbf{Inputs:} Training data $(X, Y)$, BART hyperparameters $(\alpha, \beta, m, k, \nu, q)$.\;
+  \textbf{Inputs:} Training data $(\mathbf{X}, Y)$, BART hyperparameters $(\alpha, \beta, m, k, \nu, q)$.\;
 
@@
       \For{$j = 1, \ldots, m$}{
-          (d) Compute residual $R_j^{(i)}$ \tcp*{\textit{Using \eqref{residual}}}
+          (d) Compute residual $E_j^{(i)}$ \tcp*{\textit{Using \eqref{residual}}}
 
-          (e) Sample $T_j^{(i)} \mid R_j^{(i)}, \sigma^{(i)}, T_j^{(i-1)}$ \tcp*{\textit{Using MH}}
+          (e) Sample $T_j^{(i)} \mid E_j^{(i)}, \sigma^{(i)}, T_j^{(i-1)}$ \tcp*{\textit{Using Metropolis-Hastings}}
 
-          (f) Sample $M_j^{(i)} \mid R_j^{(i)}, \sigma^{(i)}, T_j^{(i)}$ \tcp*{\textit{Sample from Gaussian distribution}}
+          (f) Sample $M_j^{(i)} \mid E_j^{(i)}, \sigma^{(i)}, T_j^{(i)}$ \tcp*{\textit{Sample from Gaussian distribution}}
       }
@@
 \end{algorithm}
-\textit{Adapted from the blog}.
+\textit{Adapted from the \cite{ParticleGibbs}}.
 
-\section{Comparison}
+\subsection{Comparison}
 \label{Comparison}
 
-We examined different ensemble methods that combine regression trees to estimate an unknown function $f(X)$ in a regression like 
+We examined different ensemble methods that combine regression trees to estimate an unknown function $f(\mathbf{x})$ in a regression like 
 \[
-Y = f(X) + \varepsilon
+y = f(\mathbf{x}) + \varepsilon
 \]
@@
 These ensembles of regression trees allows to model nonlinearity and interaction effects between covariates by fitting some sort of stepwise function on the data.
-According to Lakshminarayanan (2015), ensembles of regression trees can be
+According to \cite{ParticleGibbs}, ensembles of regression trees can be
 broadly classified into two families: randomized independent regression trees, wherein the trees are grown independently and predictions are averaged to reduce variance, and additive regression trees, wherein each
@@
 In moth methods, trees are grown dependently one from the other and thus are prone to overfitting and must be regularized.
-For example, boosting fits a sequence of single trees, using the resedual of the previous tree to fit the subsequent tree.
+For example, boosting fits a sequence of single trees, using the residual of the previous tree to fit the subsequent tree.
 Regularization in boosting is governed by parameters $\lambda$ and $d$, which shrink the contribution of each tree toward zero.
@@
 
-Overall, BART's strength comes from the fact that being a Bayesian method its shrinkage parameters does not need tuning, ad the prior will perform shrinkage automatically.
-In terms of performance, the literature (Chipman et a.) shows that BART has predictive performance comparable to other ensemble methods and far superior to linear regression when nonlinearities are present in the data.
+Overall, BART's strength comes from the fact that being a Bayesian method its shrinkage parameters do not need tuning, and the prior will perform shrinkage automatically.
+In terms of performance, the literature \cite{chipmanBARTBayesianAdditive2010} shows that BART has predictive performance comparable to other ensemble methods and far superior to linear regression when nonlinearities are present in the data.
 
@@
 We now move to applications of the theoretical insights derived in the previous sections.
-Firstly, we will focus on simulated data, and then we will cover an application using real word data on Abalones.
+Firstly, we will focus on simulated data, and then we will cover an application using real word data on Abalones (\textit{Haliotidae}).
 
@@
 \begin{equation}
-  y = 2\sin(x)+x+Z \quad Z \sim \mathcal{N}(0,1)
+  y = 2\sin(x)+x+\varepsilon \quad \varepsilon \sim \mathcal{N}(0,1)
 \end{equation}
-The dataframe is split in two smaller dataframes to obtain training and testing data.
+The dataframe is split into a training (in-sample) dataset and a testing (out-of-sample) dataset.
 
-We first run a single decision tree on the simulated data and we obtain as a result 5 terminal nodes and a 
-residual mean deviance of 1.164.
+We first run a single decision tree on the simulated data using the \texttt{tree}, and we obtain as a result $5$ terminal nodes and a 
+residual mean deviance of $1.164$.
 
-We now fit a BART model with default parameters:  MCMC have 1100 iterations in total of which 100 are burn-in.
+We now run a BART from \cite{chipmanBARTBayesianAdditive2010}.
+We keep the default parameters and we choose the number of iterations of the MCMC to be 1100, of which 100 are burn-in.
 We see from figure \ref{plot_sin1} that the BART model is able to fit correctly this non-linear trend.
-
-\begin{figure}
+Moreover, our Bayesian setting allows us to also estimate credible intervals to the data.
+\begin{figure}[h]
   \centering
@@
 
-To check convergence of the posterior, we plot in figure \ref{plot_sin2} the posterior draws of $\sigma$, the uniquely identified parameter.
-We observe that the MCMC draw converges quickly to the true parameter, although they oscillate with some autocorrelation due to the posterior sampling procedure.
+To check convergence of the posterior, we plot in figure \ref{plot_sin2} the posterior draws of $\sigma$, the uniquely identified parameter, including the burn-in period (at the left of the red line).
+We observe that the MCMC draw converges quickly to the true parameter.
+after a burn-in period of 100 iterations, although the draws oscillate with some autocorrelation due to the posterior sampling procedure, which generates dependent samples.
 
-\begin{figure}
+\begin{figure}[h]
   \centering
@@
 
-To check robustness of BART to different prior specifications, we consider different values for the prior mean of $\sigma$,  that are 100, 0.001 and 1 (the true one).
-With a large $\sigma$, we observe in plot \ref{plot_sin3} underfitting of the data.
-With a small $\sigma$, holding all the other parameter constant, the regularization prior is still strong enough to avoid the overfitting.
+To check robustness of BART to different prior specifications, we consider different values for the prior mean of $\sigma$: 100, 0.001 and 1 (the true one).
+With $\sigma = 100$ (the blue line), we observe in plot \ref{plot_sin3} underfitting of the data, because the prior is telling to BART that a large part of the variation is due to the idiosyncratic error $\varepsilon$ in \eqref{likelihood}.
+A small $\sigma$, like $\sigma = 0.001$ (the red line), should in theory push BART to overfit the training set.
+However, from the graph \ref{plot_sin3}, it is clear that the regularization prior is still strong enough to avoid the overfitting.
+The green line corresponds instead to BART estimated with the true population $\sigma$.
 
-\begin{figure}
+\begin{figure}[h]
   \centering
   \includegraphics[width=0.7\textwidth]{outputs/sin_plot_diff_sigma.pdf}
-  \caption{DIfferent prior mean of $\sigma$}
+  \caption{Different prior mean of $\sigma$}
   \label{plot_sin3}
@@
 
-Abalone are marine snails found in various regions around the world.
+Abalone are marine snails from the family \textit{Haliotidae}, which are found in various regions around the world.
 Determining their age is a labor-intensive process, as it requires cutting through the shell, staining it, and then counting the growth rings under a microscope.
@@
 
-We use data abalone characteristics data coming from the UC Irvine Machine Learning Repository \parencite{warwicknashAbalone1994a}, that is described in table \ref{table1}.
+We use Abalone characteristics data coming from the UC Irvine Machine Learning Repository \parencite{warwicknashAbalone1994a}, described in table \ref{table1}.
+Our dataset contains 4177 observations with no missing values.
 No data manipulation is necessary, aside from creating a new variable, \( \text{Age} = \text{Rings} + 1.5 \), and removing the column identifying the number of rings.
 
-\begin{table}[]
+\begin{table}[h]
   \centering
@@
   \end{table}
+Our target variable $y$ in this dataset is represented by Age. Our $p = 8$ covariates are the ones labelled as "Feature" in table \ref{table1}.
 
-We run BART from \cite{mccullochBARTBayesianAdditive2024}; Boosting from \cite{ridgewayGbmGeneralizedBoosted2024}; Random Forest and Bagging from \cite{breimanRandomForestBreimanCutlers2024}.Our aim is to compare the predictive performance of those methods.
-In order to do so, we use half of the sample to fit $\hat{y}$, that is the age variable, and half the sample to give us out of bag Mean Squared Errors (MSE) for predicting the same variable.
-Models are run for various values of T, and for each value, both the mean squared error (MSE) of out-of-sample predictions (shown in plot \ref{plot_mse}) and the computation time (depicted in plot \ref{plot_time}) are recorded.
+Our aim is to compare the predictive performance of the different tree ensemble models discussed in theory in the sections above.
+We thus run BART from \cite{mccullochBARTBayesianAdditive2024}; Boosting from \cite{ridgewayGbmGeneralizedBoosted2024}; Random Forest and Bagging from \cite{breimanRandomForestBreimanCutlers2024}.
+Notice that since BART adopts a Bayesian framework, it is able to provide a measure of the uncertantiy regarding its prediction, while other frequentist methods cannot.
+Indeed, BART allows to show boxplots of the posterior samples of predictions.
+Figure \ref{fig:bart_box} displays the boxplots of the draws of $\hat{y}$ (on the $y$-axis) ordered by the average predicted age (on the $x$-axis).
 
-We observe with a number of trees estimated larger than 500, the BART model is the best performing one, and it reaches also the best performance overall when run with 500 iterations.
-Although it is the best performer, it is also the model that takes the most to be run, and time of computations scales linearly with the number of trees that are fitted.
-
-\begin{figure}
+\begin{figure}[h]
     \centering
-  \includegraphics[width=0.7\textwidth]{outputs/mse_plot.pdf}
-  \caption{Mean Squared Error for different values of T}
-  \label{plot_mse}
+    \includegraphics[width=0.7\linewidth]{outputs/bart_boxplot.pdf}
+    \caption{Ordered boxplot of BART estimates}
+    \label{fig:bart_box}
 \end{figure}
 
-\begin{figure}
-  \centering
-  \includegraphics[width=0.7\textwidth]{outputs/time_plot.pdf}
-  \caption{Time of computations for different values of T}
-  \label{plot_time}
-\end{figure}
+In order to assess the predictive performance of the models, we divide the sample set in half: a training set used to fit  $\hat{y}$ (predicted age), and a testing set to compute the out of bag Mean Squared Errors (MSE) for predicting the same variable.
+We use the classic formula for the MSE,
+\[\text{MSE} = \frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2
+\]
+Models are run for various values of $m$, the number of trees, and for each value, both the mean squared error (MSE) of out-of-sample predictions and the computation time are recorded.
 
-\begin{table}[ht]
+Table \ref{table2} displays the MSE and the computation time for all the tree ensemble models with the number of trees equal to 500.
+Consistently to what was showed by \cite{chipmanBARTBayesianAdditive2010} for 42 different datasets, the performance of BART in terms of MSE is comparable or even slightly superior to that of other tree ensemble methods.
+However, since BART is a Bayesian method, it requires time-consuming MCMC iterations for posterior inference, even in the default prior setting.
+This increases the time needed for BART to be run relative to the other models.
+\begin{table}[h]
   \centering
@@
   \end{tabular}
-  \caption{Model performance with $T=500$}
+  \caption{Model performance with $m=500$} \label{table2}
   \end{table}
 
-\newpage
-$ $
+The same findings hold when we let the number of trees vary.
+In Figure \ref{plot_mse}, we see that the MSE for BART is lower than that of the other models, but it quickly stabilizes after 500 trees, similarly to the other models.
+It is worth noting that boosting, which in theory should be the frequentist model more similar to BART, has a significantly higher MSE for all numbers of trees, probably due to ineffective parameter tuning.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.7\textwidth]{outputs/mse_plot.pdf}
+  \caption{Mean Squared Error for different values of $m$}
+  \label{plot_mse}
+\end{figure}
+
+Figure \ref{plot_time} displays the number of seconds needed for the model to converge.
+Although it is the best performer, BART is also the model that takes the longest to be run, and computation time scales linearly with the number of trees that are fitted.
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.7\textwidth]{outputs/time_plot.pdf}
+  \caption{Time of computations for different values of $m$}
+  \label{plot_time}
+\end{figure}
+
+\section{Conclusion}
+\label{conclusion}
+BART is a Bayesian sum-of-trees model that exploits a regularization prior to constrains each tree to be a weak learner.
+Fitting and posterior inference is perfomed via a MCMC algorithm that combines Gibbs and Metropolis-Hastings.
+Since the parameter space of BART has unbounded dimension, BART is a nonparametric model, that is a Bayesian model on an infinite-dimensional
+parameter space. BART chooses the dimension of the parameter space $ (\{T_j, M_j\}_{j=1}^m, \sigma^2) $ depending on the sample, such that the effective complexity of the model (as measured by the
+number of dimensions used) adapts to the data. In particular, each tree is allowed to change in its number of decision rules and leaves. 
+The Bayesian setting of BART allows posterior inference including point and interval estimates of the unknown regression function $f(\mathbf{x})$.
+
+The data application shows that the theoretical features of BART (the prior and the posterior inference algorithm) influence its predictive performance compared to other ensembe methods.
+In particular, the necessity of doing MCMC iterations in order to fit the model implies that BART requires more computation time respect to Random Forests, Bagging, and Boosting.
+However, the flexibility due to the fact that it is a nonparamtric Bayesian model allows it to perform well even in the default setting, i.e.
+using the default hyperparameters of \texttt{gbart}.
+
+ In the simulation experiment, BART obtained posterior
+ mean and interval estimates of the true responde variable $\hat{y}$.
+BART's performance was robust to different hyperparameter specifications, for example different choices of prior $\sigma$.
+In the Abalone dataset, BART was able to perform reliable age prediction for Abalone shells given other features.
+
+\section{Acknowledgments}
+LLMs were used in this project for checking clarity and grammar of the initial drafted version, specifically in the introduction and abstract part.
+Moreover, were used to improve and check the correctness of the R-code.
+The LLMs used were ChatGPT, Claude and Github Copilot.
+
 \newpage
